---
title: "Genetic Offset Analysis for Climate Change Vulnerability"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = 'center'
)
```

# Introduction

## What is Genetic Offset?

**Genetic offset** quantifies the mismatch between a population's current genetic composition and the genetic composition predicted to be optimal under future environmental conditions. It serves as a measure of climate change vulnerability.

**Key concepts:**

- **Current adaptation**: Populations are adapted to their current local environments
- **Climate change**: Future climates will differ from current conditions
- **Genetic lag**: Populations may not evolve fast enough to track environmental change
- **Vulnerability**: High genetic offset = high vulnerability to climate change

## How Does This Analysis Work?

This workflow integrates three key components:

1. **Genomic data (VCF)**: SNP genotypes from populations across the landscape
2. **Environmental data**: Current climate variables (from WorldClim)
3. **Future climate projections**: CMIP6 models predicting future conditions

**Analysis steps:**

1. Identify genotype-environment associations using LFMM (Latent Factor Mixed Models)
2. Calculate genetic offset based on environmental shifts
3. Project vulnerability across geographic space
4. Visualize results globally and regionally

## Applications

- **Conservation planning**: Identify vulnerable populations
- **Assisted gene flow**: Guide translocation decisions
- **Climate adaptation**: Predict which populations need intervention
- **Breeding programs**: Select climate-resilient genotypes

# Setup

## Load Required Packages

```{r load-packages}
# Required packages
required_pkgs <- c("vcfR", "LEA", "lfmm", "terra", "geodata", 
                   "dplyr", "fields", "maps", "knitr", "kableExtra")

# Check for missing packages
missing <- required_pkgs[!required_pkgs %in% installed.packages()[, "Package"]]

if (length(missing) > 0) {
  cat("Installing missing packages:", paste(missing, collapse = ", "), "\n")
  install.packages(missing, repos = "https://cloud.r-project.org")
}

# Load libraries
library(vcfR)
library(LEA)
library(lfmm)
library(terra)
library(geodata)
library(dplyr)
library(fields)
library(maps)
library(knitr)
library(kableExtra)

cat("All required packages loaded successfully\n")
```

## Set Analysis Parameters

```{r set-parameters}
# ==================== INPUT FILES ====================
VCF_FILE <- "your_genotypes.vcf"                    # Path to VCF file
ENV_CSV <- "your_sample_coordinates.csv"            # CSV with sample coordinates

# ==================== WORLDCLIM PARAMETERS ====================
RES_MINUTES <- 10                                   # WorldClim resolution (2.5, 5, or 10 minutes)

# ==================== CMIP6 CLIMATE MODELS ====================
CMIP6_MODELS <- c(
  "ACCESS-ESM1-5",      # Australia
  "BCC-CSM2-MR",        # China
  "CanESM5",            # Canada
  "CNRM-CM6-1",         # France
  "IPSL-CM6A-LR",       # France
  "MIROC6"              # Japan
)

# ==================== CLIMATE SCENARIOS ====================
SSP_SCENARIOS <- c("245", "585")                    # SSP2-4.5 (moderate), SSP5-8.5 (high)
FUTURE_PERIODS <- c("2041-2060", "2081-2100")      # Mid-century and end-century

# ==================== ANALYSIS PARAMETERS ====================
GRID_RESOLUTION <- 200                              # Number of grid cells for projection
K_LFMM <- 7                                         # Number of latent factors in LFMM
PVALUE_CUTOFF <- 1e-5                               # P-value threshold for candidate SNPs
MAX_PROJ_CLIP <- NA                                 # Set numeric value to clip extreme offsets

# ==================== REGIONAL PLOTTING ====================
REGIONAL_BUFFER_DEGREES <- 2                        # Buffer around data extent for regional plots

# ==================== OUTPUT DIRECTORIES ====================
WORK_DIR <- file.path(getwd(), "cmip6_cache")      # Cache for downloaded climate data
OUT_DIR <- file.path(getwd(), "genetic_offset_results")
REGIONAL_DIR <- file.path(OUT_DIR, "regional_plots")

# Create directories
dir.create(WORK_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(REGIONAL_DIR, showWarnings = FALSE, recursive = TRUE)

# Display parameter summary
param_summary <- data.frame(
  Parameter = c("VCF File", "Coordinate CSV", "WorldClim Resolution", 
                "Climate Models", "SSP Scenarios", "Future Periods",
                "Grid Resolution", "LFMM K", "P-value Cutoff"),
  Value = c(VCF_FILE, ENV_CSV, paste(RES_MINUTES, "minutes"),
            paste(length(CMIP6_MODELS), "models"), 
            paste(SSP_SCENARIOS, collapse = ", "),
            paste(FUTURE_PERIODS, collapse = ", "),
            GRID_RESOLUTION, K_LFMM, PVALUE_CUTOFF)
)

kable(param_summary, caption = "Analysis Parameters") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Helper Functions

```{r helper-functions}
# Normalize bioclim layer names to bio1, bio2, etc.
normalize_bio_names <- function(n) {
  n2 <- tolower(n)
  digits <- gsub(".*?([0-9]+).*", "\\1", n2)
  digits[grepl("^[^0-9]+$", digits)] <- NA
  idx <- suppressWarnings(as.integer(digits))
  idx[is.na(idx)] <- NA
  paste0("bio", idx)
}

# Safe raster coercion
to_spatraster <- function(x) {
  if (inherits(x, "SpatRaster")) return(x)
  rast(x)
}

# Check if matrix is valid for plotting
is_valid_map <- function(m) {
  is.matrix(m) && any(is.finite(m))
}

# Check if site is polymorphic
is_polymorphic <- function(g) {
  g2 <- g[g != 9]  # Remove missing (coded as 9)
  if (length(g2) == 0) return(FALSE)
  u <- unique(g2)
  length(u) > 1  # Polymorphic if at least 2 genotypes
}
```

# Data Import and Preparation

## Read VCF and Prepare Genotype Matrix

```{r read-vcf}
cat("Reading VCF and preparing genotype matrix...\n")

# Check if file exists
if (!file.exists(VCF_FILE)) {
  stop("VCF file not found: ", VCF_FILE)
}

# Convert VCF to LEA .geno format
geno_file <- vcf2geno(VCF_FILE)

# Read genotype matrix (0/1/2/9, where 9 = missing)
genotype <- LEA::read.geno(geno_file)

cat("Raw genotype dimensions (individuals x SNPs):", dim(genotype), "\n")

# Remove monomorphic sites
keep <- apply(genotype, 2, is_polymorphic)

cat("Monomorphic SNPs removed:", sum(!keep), "\n")

genotype <- genotype[, keep, drop = FALSE]

cat("Filtered genotype dimensions:", dim(genotype), "\n")

# Create genotype summary
geno_summary <- data.frame(
  Metric = c("Total individuals", "Raw SNPs", "Monomorphic SNPs removed", 
             "Polymorphic SNPs retained", "Missing genotype rate"),
  Value = c(
    nrow(genotype),
    length(keep),
    sum(!keep),
    sum(keep),
    paste0(round(100 * mean(genotype == 9), 2), "%")
  )
)

kable(geno_summary, caption = "Genotype Data Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Read Sample Coordinates

```{r read-coordinates}
cat("Reading sample coordinates from:", ENV_CSV, "\n")

# Read CSV
env_df <- read.csv(ENV_CSV, stringsAsFactors = FALSE)
colnames(env_df) <- tolower(colnames(env_df))

# Detect latitude and longitude columns
if (!any(grepl("^lat", colnames(env_df)))) {
  stop("No latitude column detected in CSV")
}
if (!any(grepl("^lon|^long", colnames(env_df)))) {
  stop("No longitude column detected in CSV")
}

lat_col <- grep("^lat", colnames(env_df), value = TRUE)[1]
lon_col <- grep("^lon|^long", colnames(env_df), value = TRUE)[1]
sample_col <- colnames(env_df)[1]

# Create coordinate dataframe
coords <- data.frame(
  Longitude = env_df[[lon_col]], 
  Latitude = env_df[[lat_col]]
)
rownames(coords) <- env_df[[sample_col]]

cat("Loaded", nrow(coords), "sample coordinates\n")

# Calculate data extent
min_lon <- min(coords$Longitude, na.rm = TRUE)
max_lon <- max(coords$Longitude, na.rm = TRUE)
min_lat <- min(coords$Latitude, na.rm = TRUE)
max_lat <- max(coords$Latitude, na.rm = TRUE)

cat("Data extent:\n")
cat("  Longitude: [", min_lon, ",", max_lon, "]\n")
cat("  Latitude:  [", min_lat, ",", max_lat, "]\n")

# Display coordinate summary
coord_summary <- data.frame(
  Statistic = c("Min Longitude", "Max Longitude", "Min Latitude", "Max Latitude",
                "Longitude Range", "Latitude Range"),
  Value = c(
    round(min_lon, 4),
    round(max_lon, 4),
    round(min_lat, 4),
    round(max_lat, 4),
    round(max_lon - min_lon, 4),
    round(max_lat - min_lat, 4)
  )
)

kable(coord_summary, caption = "Sample Coordinate Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Save extent information
extent_df <- data.frame(
  parameter = c("min_longitude", "max_longitude", "min_latitude", "max_latitude"),
  value = c(min_lon, max_lon, min_lat, max_lat)
)
write.csv(extent_df, file.path(OUT_DIR, "data_extent.csv"), row.names = FALSE)
```

## Visualize Sample Locations

```{r plot-samples, fig.width=10, fig.height=6}
# Create simple map of sample locations
plot(coords$Longitude, coords$Latitude, 
     pch = 21, bg = "red", col = "black", cex = 1.5,
     xlab = "Longitude", ylab = "Latitude",
     main = "Sample Locations")
maps::map(add = TRUE, col = "gray70")
grid()
```

# Historical Climate Data

## Download WorldClim Bioclimatic Variables

```{r download-worldclim}
cat("Downloading/reading historical WorldClim bioclimatic variables...\n")
cat("Resolution:", RES_MINUTES, "minutes\n")

# Download WorldClim data
clim_hist_raw <- worldclim_global(
  var = "bio", 
  res = RES_MINUTES, 
  path = WORK_DIR, 
  download = TRUE
)

clim_hist <- clim_hist_raw

# Standardize layer names
names(clim_hist) <- paste0("bio", seq_len(nlyr(clim_hist)))

cat("Historical raster layers:", paste(names(clim_hist), collapse = ", "), "\n")

# Display WorldClim variables
bioclim_vars <- data.frame(
  Variable = paste0("bio", 1:19),
  Description = c(
    "Annual Mean Temperature",
    "Mean Diurnal Range",
    "Isothermality",
    "Temperature Seasonality",
    "Max Temperature of Warmest Month",
    "Min Temperature of Coldest Month",
    "Temperature Annual Range",
    "Mean Temperature of Wettest Quarter",
    "Mean Temperature of Driest Quarter",
    "Mean Temperature of Warmest Quarter",
    "Mean Temperature of Coldest Quarter",
    "Annual Precipitation",
    "Precipitation of Wettest Month",
    "Precipitation of Driest Month",
    "Precipitation Seasonality",
    "Precipitation of Wettest Quarter",
    "Precipitation of Driest Quarter",
    "Precipitation of Warmest Quarter",
    "Precipitation of Coldest Quarter"
  )
)

kable(bioclim_vars, caption = "WorldClim Bioclimatic Variables (BIO1-BIO19)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Extract Environmental Data at Sample Points

```{r extract-environment}
cat("Extracting environmental data at sample locations...\n")

# Create point data
pts <- data.frame(x = coords$Longitude, y = coords$Latitude)

# Extract values
X.env_df <- terra::extract(clim_hist, pts)

if (ncol(X.env_df) < 2) {
  stop("Historical extraction failed or has no bands.")
}

# Remove ID column
X.env <- X.env_df[, -1, drop = FALSE]

# Impute missing values with column means
X.env <- as.data.frame(lapply(X.env, function(x) {
  x[is.na(x)] <- mean(x, na.rm = TRUE)
  x
}))

# Keep only non-zero variance columns
X.env <- X.env[, sapply(X.env, function(x) sd(x, na.rm = TRUE) > 0), drop = FALSE]

if (ncol(X.env) == 0) {
  stop("No non-zero-variance historic environmental columns.")
}

# Convert to matrix and scale
X.env_mat <- as.matrix(X.env)
X.env_scaled <- scale(X.env_mat)

cat("Historical environment dimensions (samples x variables):", dim(X.env_mat), "\n")
cat("Variables with non-zero variance:", ncol(X.env), "\n")

# Display environmental data summary
env_summary <- data.frame(
  Variable = colnames(X.env),
  Mean = round(colMeans(X.env), 2),
  SD = round(apply(X.env, 2, sd), 2),
  Min = round(apply(X.env, 2, min), 2),
  Max = round(apply(X.env, 2, max), 2)
)

kable(env_summary, caption = "Environmental Data Summary at Sample Points") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# LFMM: Genotype-Environment Association

## Run LFMM Analysis

```{r run-lfmm}
cat("Running LFMM to identify genotype-environment associations...\n")
cat("This may take several minutes depending on dataset size...\n")

# Remove SNPs with complete missing data
geno_filtered <- genotype[, colSums(genotype == 9) == 0]

cat("SNPs after removing completely missing:", ncol(geno_filtered), "\n")

# Run LFMM2
mod_lfmm <- lfmm2(
  input = geno_filtered, 
  env = X.env_scaled, 
  K = K_LFMM, 
  effect.sizes = TRUE
)

# Extract effect size matrix
B_mat <- mod_lfmm@B

# Test for significance
pv <- lfmm2.test(
  mod_lfmm, 
  input = geno_filtered, 
  env = X.env_scaled, 
  full = TRUE
)

# Identify candidate SNPs
candidates <- which(-log10(pv$pvalues) > -log10(PVALUE_CUTOFF))

cat("Candidate SNPs identified:", length(candidates), "\n")
cat("Proportion of genome:", round(100 * length(candidates) / ncol(geno_filtered), 2), "%\n")

if (length(candidates) == 0) {
  stop("No candidate SNPs found. Consider:\n",
       "  1. Loosening pvalue_cutoff\n",
       "  2. Checking data quality\n",
       "  3. Increasing sample size")
}

# Create LFMM summary
lfmm_summary <- data.frame(
  Metric = c("Total SNPs tested", "Latent factors (K)", "P-value threshold",
             "Candidate SNPs", "Proportion of genome", "Environmental variables"),
  Value = c(
    ncol(geno_filtered),
    K_LFMM,
    PVALUE_CUTOFF,
    length(candidates),
    paste0(round(100 * length(candidates) / ncol(geno_filtered), 2), "%"),
    ncol(X.env_scaled)
  )
)

kable(lfmm_summary, caption = "LFMM Analysis Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Visualize LFMM Results

```{r plot-lfmm, fig.width=10, fig.height=6}
# Manhattan-style plot of -log10(p-values)
log_pvals <- -log10(pv$pvalues)

plot(1:length(log_pvals), log_pvals,
     pch = 20, cex = 0.5, col = ifelse(log_pvals > -log10(PVALUE_CUTOFF), "red", "gray50"),
     xlab = "SNP Index", ylab = "-log10(p-value)",
     main = "LFMM: Genotype-Environment Associations")
abline(h = -log10(PVALUE_CUTOFF), col = "blue", lty = 2, lwd = 2)
legend("topright", legend = c("Candidate SNPs", "Neutral SNPs", "Threshold"),
       col = c("red", "gray50", "blue"), pch = c(20, 20, NA), lty = c(NA, NA, 2),
       lwd = c(NA, NA, 2))
```

# Create Projection Grid

```{r create-grid}
cat("Creating global projection grid...\n")
cat("Grid resolution:", GRID_RESOLUTION, "cells\n")

# Create longitude and latitude vectors
long.vec <- seq(-180, 180, length = GRID_RESOLUTION)
lat.vec <- seq(-90, 90, length = GRID_RESOLUTION)

# Create coordinate grid
coord.grid <- expand.grid(Longitude = long.vec, Latitude = lat.vec)

cat("Total grid cells:", nrow(coord.grid), "\n")

# Create spatial vector for extraction
grid_pts <- vect(coord.grid, geom = c("Longitude", "Latitude"), crs = "EPSG:4326")

cat("Projection grid created successfully\n")
```

# Future Climate Projections and Genetic Offset

## Process CMIP6 Models

This section downloads future climate projections and calculates genetic offset for each model, scenario, and time period.

**Note:** This can take considerable time depending on:
- Number of models
- Number of scenarios and periods
- Internet connection speed
- Grid resolution

```{r process-cmip6, cache=TRUE}
cat("Processing CMIP6 climate models...\n")
cat("Models:", paste(CMIP6_MODELS, collapse = ", "), "\n")
cat("Scenarios:", paste(SSP_SCENARIOS, collapse = ", "), "\n")
cat("Periods:", paste(FUTURE_PERIODS, collapse = ", "), "\n\n")

# Storage for results
combo_maps <- list()         # Mean maps per SSP×period
all_per_model_maps <- list() # Individual model maps
per_model_info <- list()     # Metadata

# Initialize progress tracking
total_combinations <- length(CMIP6_MODELS) * length(SSP_SCENARIOS) * length(FUTURE_PERIODS)
current_iteration <- 0

# Loop through all combinations
for (model_name in CMIP6_MODELS) {
  for (ssp in SSP_SCENARIOS) {
    for (period in FUTURE_PERIODS) {
      
      current_iteration <- current_iteration + 1
      combo_id <- paste0("ssp", ssp, "_", period)
      run_id <- paste0(model_name, "_ssp", ssp, "_", period)
      
      cat("\n[", current_iteration, "/", total_combinations, "] Processing:", run_id, "\n")
      
      tryCatch({
        # Download future climate data
        cat("  Downloading future climate data...\n")
        clim_fut_raw <- cmip6_world(
          model = model_name,
          ssp = ssp,
          time = period,
          var = "bioc",
          res = RES_MINUTES,
          path = WORK_DIR
        )
        
        clim_fut <- to_spatraster(clim_fut_raw)
        
        # Normalize layer names
        old_names <- names(clim_fut)
        new_names <- normalize_bio_names(old_names)
        names(clim_fut) <- new_names
        
        cat("  Future layers:", paste(names(clim_fut), collapse = ", "), "\n")
        
        # Match layers between historical and future
        common_layers <- intersect(names(clim_hist), names(clim_fut))
        
        if (length(common_layers) == 0) {
          cat("  WARNING: No common layers found. Skipping.\n")
          next
        }
        
        cat("  Common layers:", paste(common_layers, collapse = ", "), "\n")
        
        # Subset to common layers
        clim_hist_subset <- clim_hist[[common_layers]]
        clim_fut_subset <- clim_fut[[common_layers]]
        
        # Extract future environment at grid points
        cat("  Extracting future environment at grid points...\n")
        X.pred_df <- terra::extract(clim_fut_subset, grid_pts)
        
        if (ncol(X.pred_df) < 2) {
          cat("  WARNING: Future extraction failed. Skipping.\n")
          next
        }
        
        X.pred <- X.pred_df[, -1, drop = FALSE]
        
        # Align columns with training data
        train_cols <- colnames(X.env_mat)
        pred_cols_available <- intersect(train_cols, colnames(X.pred))
        
        if (length(pred_cols_available) == 0) {
          cat("  WARNING: No matching environmental variables. Skipping.\n")
          next
        }
        
        X.pred_aligned <- X.pred[, pred_cols_available, drop = FALSE]
        X.env_mat_aligned <- X.env_mat[, pred_cols_available, drop = FALSE]
        
        # Scale future data using historical mean and SD
        X.pred_scaled <- scale(X.pred_aligned, 
                              center = attr(X.env_scaled, "scaled:center")[pred_cols_available],
                              scale = attr(X.env_scaled, "scaled:scale")[pred_cols_available])
        
        # Calculate genetic offset
        cat("  Calculating genetic offset...\n")
        
        offset_vec <- apply(X.pred_scaled, 1, function(x.pred) {
          if (any(is.na(x.pred))) return(NA)
          
          # Genetic turnover calculation
          pred_genetic <- B_mat[candidates, pred_cols_available, drop = FALSE] %*% x.pred
          
          # Calculate offset for each training sample
          offsets_per_sample <- sapply(1:nrow(X.env_mat_aligned), function(i) {
            x.train <- X.env_mat_aligned[i, pred_cols_available]
            x.train_scaled <- (x.train - attr(X.env_scaled, "scaled:center")[pred_cols_available]) / 
                             attr(X.env_scaled, "scaled:scale")[pred_cols_available]
            
            train_genetic <- B_mat[candidates, pred_cols_available, drop = FALSE] %*% x.train_scaled
            
            # Euclidean distance in genetic space
            sqrt(sum((pred_genetic - train_genetic)^2))
          })
          
          # Return mean offset
          mean(offsets_per_sample, na.rm = TRUE)
        })
        
        # Convert vector to matrix
        offset_mat <- matrix(offset_vec, nrow = GRID_RESOLUTION, ncol = GRID_RESOLUTION, byrow = FALSE)
        
        # Apply clipping if specified
        if (!is.na(MAX_PROJ_CLIP)) {
          offset_mat[offset_mat > MAX_PROJ_CLIP] <- MAX_PROJ_CLIP
        }
        
        # Store results
        all_per_model_maps[[run_id]] <- offset_mat
        
        # Add to combo accumulator
        if (!combo_id %in% names(combo_maps)) {
          combo_maps[[combo_id]] <- list()
        }
        combo_maps[[combo_id]][[model_name]] <- offset_mat
        
        # Store metadata
        per_model_info[[run_id]] <- list(
          model = model_name,
          ssp = ssp,
          period = period,
          common_layers = common_layers,
          mean_offset = mean(offset_vec, na.rm = TRUE),
          max_offset = max(offset_vec, na.rm = TRUE),
          na_count = sum(is.na(offset_vec))
        )
        
        cat("  Mean genetic offset:", round(mean(offset_vec, na.rm = TRUE), 4), "\n")
        cat("  Max genetic offset:", round(max(offset_vec, na.rm = TRUE), 4), "\n")
        cat("  Completed successfully!\n")
        
      }, error = function(e) {
        cat("  ERROR:", e$message, "\n")
        cat("  Skipping this combination.\n")
      })
    }
  }
}

cat("\nAll CMIP6 model processing complete!\n")
cat("Successfully processed", length(all_per_model_maps), "model combinations\n")
```

## Calculate Mean Genetic Offset

```{r calculate-means}
cat("Calculating mean genetic offset across models...\n")

# Calculate mean for each SSP×period combination
for (combo_id in names(combo_maps)) {
  model_list <- combo_maps[[combo_id]]
  
  if (length(model_list) == 0) {
    cat("Warning: No models for", combo_id, "\n")
    next
  }
  
  # Stack all model matrices
  stacked <- simplify2array(model_list)
  
  # Calculate mean across models
  mean_map <- apply(stacked, c(1, 2), mean, na.rm = TRUE)
  
  combo_maps[[combo_id]] <- mean_map
  
  cat("Combo:", combo_id, "- Mean offset:", round(mean(mean_map, na.rm = TRUE), 4), "\n")
}

# Calculate grand mean across all scenarios and periods
cat("\nCalculating grand mean across all combinations...\n")

all_combo_means <- lapply(combo_maps, function(x) if(is_valid_map(x)) x else NULL)
all_combo_means <- all_combo_means[!sapply(all_combo_means, is.null)]

if (length(all_combo_means) > 0) {
  stacked_all <- simplify2array(all_combo_means)
  grand_mean <- apply(stacked_all, c(1, 2), mean, na.rm = TRUE)
  
  cat("Grand mean offset:", round(mean(grand_mean, na.rm = TRUE), 4), "\n")
  cat("Grand max offset:", round(max(grand_mean, na.rm = TRUE), 4), "\n")
} else {
  stop("No valid combo maps to calculate grand mean")
}
```

# Visualization

## Global Genetic Offset Maps

```{r plot-global-maps, fig.width=14, fig.height=8}
# Determine global color scale
global_min <- min(sapply(all_per_model_maps, function(m) min(m, na.rm = TRUE)))
global_max <- max(sapply(all_per_model_maps, function(m) max(m, na.rm = TRUE)))

cat("Global offset range: [", global_min, ",", global_max, "]\n")

# Color palette
color_fun <- colorRampPalette(c("blue", "cyan", "yellow", "orange", "red"))(100)

# Plot grand mean
par(mfrow = c(1, 1), mar = c(5, 5, 4, 6))

fields::image.plot(
  long.vec, lat.vec, grand_mean,
  col = color_fun,
  zlim = c(global_min, global_max),
  xlab = "Longitude", 
  ylab = "Latitude",
  main = "Grand Mean Genetic Offset (All Models, SSPs, & Periods)"
)

# Add sample points
points(coords$Longitude, coords$Latitude, pch = 21, bg = "white", col = "black", cex = 1.5)

# Add country outlines
maps::map(add = TRUE, col = "gray30", lwd = 0.5)
```

## Individual Scenario Maps

```{r plot-scenarios, fig.width=14, fig.height=10}
# Plot each SSP×period combination
n_combos <- length(combo_maps)
n_cols <- 2
n_rows <- ceiling(n_combos / n_cols)

par(mfrow = c(n_rows, n_cols), mar = c(4, 4, 3, 5))

for (combo_id in names(combo_maps)) {
  mean_map <- combo_maps[[combo_id]]
  
  if (!is_valid_map(mean_map)) next
  
  fields::image.plot(
    long.vec, lat.vec, mean_map,
    col = color_fun,
    zlim = c(global_min, global_max),
    xlab = "Longitude",
    ylab = "Latitude",
    main = paste("Mean Genetic Offset -", combo_id)
  )
  
  points(coords$Longitude, coords$Latitude, pch = 21, bg = "white", col = "black", cex = 1)
  maps::map(add = TRUE, col = "gray30", lwd = 0.5)
}
```

## Regional Maps (Data Extent)

```{r plot-regional, fig.width=12, fig.height=8}
# Calculate buffered extent for regional plots
buffer_lon <- REGIONAL_BUFFER_DEGREES
buffer_lat <- REGIONAL_BUFFER_DEGREES

min_lon_buffered <- max(min_lon - buffer_lon, -180)
max_lon_buffered <- min(max_lon + buffer_lon, 180)
min_lat_buffered <- max(min_lat - buffer_lat, -90)
max_lat_buffered <- min(max_lat + buffer_lat, 90)

cat("Regional extent (with buffer):\n")
cat("  Longitude: [", min_lon_buffered, ",", max_lon_buffered, "]\n")
cat("  Latitude:  [", min_lat_buffered, ",", max_lat_buffered, "]\n")

# Find grid indices within extent
lon_idx <- which(long.vec >= min_lon_buffered & long.vec <= max_lon_buffered)
lat_idx <- which(lat.vec >= min_lat_buffered & lat.vec <= max_lat_buffered)

if (length(lon_idx) > 1 && length(lat_idx) > 1) {
  
  # Extract regional coordinates
  long.vec.regional <- long.vec[lon_idx]
  lat.vec.regional <- lat.vec[lat_idx]
  
  # Extract regional data
  grand_mean_regional <- grand_mean[lon_idx, lat_idx, drop = FALSE]
  
  # Plot regional grand mean
  par(mfrow = c(1, 1), mar = c(5, 5, 4, 6))
  
  fields::image.plot(
    long.vec.regional, lat.vec.regional, grand_mean_regional,
    col = color_fun,
    zlim = c(global_min, global_max),
    xlab = "Longitude",
    ylab = "Latitude",
    main = "Regional Grand Mean Genetic Offset"
  )
  
  # Add sample points
  points(coords$Longitude, coords$Latitude, pch = 21, bg = "white", col = "black", cex = 2)
  
  # Add map features
  maps::map(xlim = range(long.vec.regional), ylim = range(lat.vec.regional),
            add = TRUE, interior = TRUE, col = "grey30", lwd = 0.8)
  maps::map(xlim = range(long.vec.regional), ylim = range(lat.vec.regional),
            add = TRUE, interior = FALSE, col = "grey20", lwd = 1.5)
  
} else {
  cat("Regional extent too small for plotting\n")
}
```

# Save Results

```{r save-results}
cat("Saving results to disk...\n")

# Save candidate SNPs
write.csv(
  data.frame(candidate_index = candidates),
  file = file.path(OUT_DIR, "candidate_snps_index.csv"),
  row.names = FALSE
)

# Save LFMM effect size matrix
write.csv(
  as.data.frame(B_mat),
  file = file.path(OUT_DIR, "LFMM_B_matrix.csv"),
  row.names = TRUE
)

# Save per-model metadata
saveRDS(per_model_info, file = file.path(OUT_DIR, "per_model_debug_info.rds"))

# Save global mean as GeoTIFF
r_grand <- rast(t(grand_mean))
ext(r_grand) <- c(-180, 180, -90, 90)
crs(r_grand) <- "EPSG:4326"
writeRaster(r_grand, file.path(OUT_DIR, "offset_GLOBAL_MEAN.tif"), overwrite = TRUE)

cat("Grand mean raster saved to:", file.path(OUT_DIR, "offset_GLOBAL_MEAN.tif"), "\n")

# Save individual scenario rasters
for (combo_id in names(combo_maps)) {
  mean_map <- combo_maps[[combo_id]]
  if (!is_valid_map(mean_map)) next
  
  r_combo <- rast(t(mean_map))
  ext(r_combo) <- c(-180, 180, -90, 90)
  crs(r_combo) <- "EPSG:4326"
  
  tif_file <- file.path(OUT_DIR, paste0("offset_", combo_id, "_MEAN.tif"))
  writeRaster(r_combo, tif_file, overwrite = TRUE)
}

cat("\nAll scenario rasters saved\n")

# Save regional rasters if applicable
if (length(lon_idx) > 1 && length(lat_idx) > 1) {
  
  # Regional grand mean
  r_grand_regional <- rast(t(grand_mean_regional))
  ext(r_grand_regional) <- c(min(long.vec.regional), max(long.vec.regional),
                             min(lat.vec.regional), max(lat.vec.regional))
  crs(r_grand_regional) <- "EPSG:4326"
  writeRaster(r_grand_regional, 
              file.path(REGIONAL_DIR, "REGIONAL_offset_GRAND_MEAN.tif"),
              overwrite = TRUE)
  
  cat("Regional grand mean raster saved\n")
}

cat("\nAll results saved successfully!\n")
```

# Summary and Interpretation

## Results Summary

```{r results-summary}
# Create comprehensive summary table
results_summary <- data.frame(
  Metric = c(
    "Total samples",
    "Total SNPs (polymorphic)",
    "Candidate SNPs",
    "Proportion of genome",
    "Climate models processed",
    "SSP scenarios",
    "Time periods",
    "Total model combinations",
    "Mean genetic offset (grand mean)",
    "Max genetic offset (grand mean)"
  ),
  Value = c(
    nrow(coords),
    ncol(genotype),
    length(candidates),
    paste0(round(100 * length(candidates) / ncol(genotype), 2), "%"),
    length(CMIP6_MODELS),
    paste(SSP_SCENARIOS, collapse = ", "),
    paste(FUTURE_PERIODS, collapse = ", "),
    length(all_per_model_maps),
    round(mean(grand_mean, na.rm = TRUE), 4),
    round(max(grand_mean, na.rm = TRUE), 4)
  )
)

kable(results_summary, caption = "Analysis Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Interpretation Guide

### Understanding Genetic Offset Values

**Genetic offset** represents the magnitude of environmental change relative to current genotype-environment associations.

- **Low offset (< 0.5)**: Populations well-adapted to future conditions
- **Moderate offset (0.5-1.5)**: Some adaptation challenges expected
- **High offset (> 1.5)**: Significant vulnerability to climate change

**Important notes:**

- Values are relative, not absolute
- Compare across space to identify vulnerable regions
- Higher values = greater mismatch between current genetics and future environment

### Regional Patterns

Look for:

1. **Hotspots**: Areas with very high genetic offset
   - Priority for conservation intervention
   - Candidates for assisted gene flow
   
2. **Refugia**: Areas with low genetic offset
   - May serve as climate refugia
   - Source populations for translocation

3. **Gradients**: Directional patterns in offset
   - May guide translocation strategies
   - Indicate direction of climate change impact

### Management Implications

**For high offset areas:**

- **Monitor**: Increased surveillance of populations
- **Protect**: Enhance habitat protection
- **Assist**: Consider assisted migration
- **Breed**: Use in breeding programs for climate resilience

**For low offset areas:**

- **Conserve**: Protect as potential refugia
- **Source**: Use as seed sources for restoration
- **Study**: Investigate adaptive mechanisms

### Next Steps

1. **Validate predictions**: 
   - Monitor population responses to climate change
   - Compare with empirical fitness data
   
2. **Functional analysis**:
   - Annotate candidate SNPs
   - Investigate functions of associated genes
   
3. **Integrate with other data**:
   - Habitat suitability models
   - Demographic projections
   - Connectivity analyses

4. **Conservation planning**:
   - Prioritize management actions
   - Design translocation experiments
   - Develop climate adaptation strategies

## Output Files

**Data files:**

- `candidate_snps_index.csv` - Indices of candidate SNPs
- `LFMM_B_matrix.csv` - Effect size matrix from LFMM
- `per_model_debug_info.rds` - Metadata for each model run
- `data_extent.csv` - Geographic extent of data

**Raster files (GeoTIFF):**

- `offset_GLOBAL_MEAN.tif` - Grand mean across all models and scenarios
- `offset_ssp245_2041-2060_MEAN.tif` - Mean for each scenario/period combination
- `offset_ssp245_2081-2100_MEAN.tif`
- `offset_ssp585_2041-2060_MEAN.tif`
- `offset_ssp585_2081-2100_MEAN.tif`

**Regional files (if applicable):**

- `REGIONAL_offset_GRAND_MEAN.tif` - Regional grand mean
- Additional regional files in `regional_plots/` directory

# Session Information

```{r session-info}
sessionInfo()
```

# References

1. Fitzpatrick, M. C., & Keller, S. R. (2015). Ecological genomics meets community-level modelling of biodiversity: mapping the genomic landscape of current and future environmental adaptation. *Ecology Letters*, 18(1), 1-16.

2. Capblancq, T., et al. (2020). Evaluation of redundancy analysis to identify signatures of local adaptation. *Molecular Ecology Resources*, 20(6), 1625-1640.

3. Caye, K., et al. (2019). LFMM 2: Fast and accurate inference of gene-environment associations in genome-wide studies. *Molecular Biology and Evolution*, 36(4), 852-860.

4. Rellstab, C., et al. (2016). A practical guide to environmental association analysis in landscape genomics. *Molecular Ecology*, 25(17), 4095-4109.

5. Bay, R. A., et al. (2018). Genomic signals of selection predict climate-driven population declines in a migratory bird. *Science*, 359(6371), 83-86.

---

**Analysis completed:** `r Sys.time()`
