---
title: "Weighted Z-score Analysis (WZA) for Environmental Association"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6
)
```

## Overview

**Weighted Z-score Analysis (WZA)** combines genetic-environment association (GEA) signals across multiple environmental variables to identify loci under selection.

**Reference:** Booker et al. (2021) - [https://github.com/TBooker/WZA](https://github.com/TBooker/WZA)

### Method

1. For each SNP: calculate z-scores from GEA tests with each environmental variable
2. Weight z-scores by R² (strength of association)  
3. **WZA score** = sum(|z| × R²) / sum(R²)

### Requirements

- **VCF file** with genotypes
- **Metadata file** with:
  - Sample IDs matching VCF
  - Population/group labels
  - Environmental variables (continuous numeric)

---

## Configuration

**Modify these parameters for your dataset:**

```{r config}
################################################################################
# USER CONFIGURATION
################################################################################

# Input files
VCF_FILE <- "your_genotypes.vcf"  # VCF or VCF.gz format
METADATA_FILE <- "your_metadata.csv"

# Column names in metadata
SAMPLE_COL <- "Sample_ID"      # Column with sample identifiers
POP_COL <- "Population"        # Column with population/group labels

# Environmental variables
# Provide a vector of column names from your metadata file
ENV_VARS <- c("bio_01", "bio_02", "bio_05", "bio_12", "bio_15")
# Examples:
# - Bioclimatic: c("bio_01", "bio_12", "bio_15")
# - Custom: c("temperature", "precipitation", "elevation")
# - All matching pattern: grep("^bio_", colnames(metadata), value = TRUE)

# Analysis parameters
MIN_MAF <- 0.05              # Minimum minor allele frequency (0.05 = 5%)
MIN_SAMPLES <- 20            # Minimum samples required for regression
OUTLIER_PERCENTILE <- 0.99   # Top X% threshold for outliers (0.99 = top 1%)

# Output prefix for all files
OUTPUT_PREFIX <- "WZA_results"

# Number of top SNPs to display in report
N_TOP_SNPS <- 10
```

---

## Load Required Libraries

```{r libraries}
################################################################################
# LOAD LIBRARIES
################################################################################

# Check and install missing packages
required_packages <- c("vcfR", "dplyr", "tidyr", "ggplot2")

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}

cat("All required packages loaded successfully\n")
```

---

## Define Helper Functions

```{r functions}
################################################################################
# HELPER FUNCTIONS
################################################################################

# Calculate minor allele frequency
calc_maf <- function(geno) {
  # geno should be 0, 1, 2 (or NA/9 for missing)
  geno_clean <- geno[!is.na(geno) & geno != 9]
  if (length(geno_clean) == 0) return(NA)
  
  # Allele frequency
  p <- sum(geno_clean) / (2 * length(geno_clean))
  # Minor allele frequency
  min(p, 1 - p)
}

# Calculate z-score from linear regression
calc_z_score <- function(genotypes, environment) {
  # Remove missing data
  keep <- !is.na(genotypes) & !is.na(environment)
  
  if (sum(keep) < MIN_SAMPLES) {
    return(list(z = NA, r2 = NA, p = NA))
  }
  
  g <- genotypes[keep]
  e <- environment[keep]
  
  # Linear regression: environment ~ genotype
  fit <- tryCatch({
    lm(e ~ g)
  }, error = function(err) {
    return(NULL)
  })
  
  if (is.null(fit)) {
    return(list(z = NA, r2 = NA, p = NA))
  }
  
  # Extract results
  coef_summary <- summary(fit)$coefficients
  
  if (nrow(coef_summary) < 2) {
    return(list(z = NA, r2 = NA, p = NA))
  }
  
  # Z-score = coefficient / standard error
  beta <- coef_summary[2, "Estimate"]
  se <- coef_summary[2, "Std. Error"]
  z <- beta / se
  
  # P-value and R-squared
  p <- coef_summary[2, "Pr(>|t|)"]
  r2 <- summary(fit)$r.squared
  
  list(z = z, r2 = r2, p = p)
}

# Calculate WZA score for a single SNP across all environments
calc_wza <- function(genotypes, env_matrix) {
  # env_matrix: samples × environmental variables
  
  results <- lapply(1:ncol(env_matrix), function(i) {
    calc_z_score(genotypes, env_matrix[, i])
  })
  
  # Extract z-scores and R² values
  z_scores <- sapply(results, function(x) x$z)
  r2_values <- sapply(results, function(x) x$r2)
  p_values <- sapply(results, function(x) x$p)
  
  # Remove NA values
  valid <- !is.na(z_scores) & !is.na(r2_values)
  
  if (sum(valid) == 0) {
    return(list(
      wza = NA,
      mean_abs_z = NA,
      max_abs_z = NA,
      sum_r2 = NA,
      n_sig = NA
    ))
  }
  
  z_valid <- z_scores[valid]
  r2_valid <- r2_values[valid]
  p_valid <- p_values[valid]
  
  # WZA score: weighted mean of absolute z-scores
  abs_z <- abs(z_valid)
  
  if (sum(r2_valid) > 0) {
    wza <- sum(abs_z * r2_valid) / sum(r2_valid)
  } else {
    wza <- mean(abs_z)
  }
  
  # Additional metrics
  mean_abs_z <- mean(abs_z)
  max_abs_z <- max(abs_z)
  sum_r2 <- sum(r2_valid)
  n_sig <- sum(p_valid < 0.05, na.rm = TRUE)
  
  list(
    wza = wza,
    mean_abs_z = mean_abs_z,
    max_abs_z = max_abs_z,
    sum_r2 = sum_r2,
    n_sig = n_sig
  )
}

cat("Helper functions defined\n")
```

---

## Load Data

### Load VCF File

```{r load_vcf}
################################################################################
# LOAD VCF
################################################################################

cat("Loading VCF:", VCF_FILE, "\n")

if (!file.exists(VCF_FILE)) {
  stop("VCF file not found: ", VCF_FILE)
}

vcf <- read.vcfR(VCF_FILE, verbose = FALSE)

cat(sprintf("Successfully loaded: %s SNPs, %s samples\n", 
            format(nrow(vcf@fix), big.mark = ","),
            ncol(vcf@gt) - 1))

# Extract genotype matrix (as numeric: 0, 1, 2)
gt <- extract.gt(vcf, element = "GT", as.numeric = TRUE)
gt <- t(gt)  # Transpose to samples × SNPs

# Get SNP information
snp_data <- data.frame(
  SNP_index = 1:ncol(gt),
  CHROM = vcf@fix[, "CHROM"],
  POS = as.numeric(vcf@fix[, "POS"]),
  stringsAsFactors = FALSE
)

cat("\nFirst few samples:", paste(head(rownames(gt), 5), collapse = ", "), "\n")
```

### Load Metadata

```{r load_metadata}
################################################################################
# LOAD METADATA
################################################################################

cat("\nLoading metadata:", METADATA_FILE, "\n")

if (!file.exists(METADATA_FILE)) {
  stop("Metadata file not found: ", METADATA_FILE)
}

meta <- read.csv(METADATA_FILE, stringsAsFactors = FALSE, check.names = FALSE)

cat(sprintf("Successfully loaded: %s samples\n", format(nrow(meta), big.mark = ",")))

# Check required columns
if (!SAMPLE_COL %in% colnames(meta)) {
  stop("Sample column '", SAMPLE_COL, "' not found in metadata")
}

if (!POP_COL %in% colnames(meta)) {
  stop("Population column '", POP_COL, "' not found in metadata")
}

# Check environmental variables
env_available <- ENV_VARS[ENV_VARS %in% colnames(meta)]

if (length(env_available) == 0) {
  stop("No environmental variables found in metadata. Check ENV_VARS configuration.")
}

cat(sprintf("\nFound %s/%s specified environmental variables:\n", 
            length(env_available), length(ENV_VARS)))
cat(paste("  -", env_available, collapse = "\n"), "\n")

if (length(env_available) < length(ENV_VARS)) {
  missing <- setdiff(ENV_VARS, env_available)
  cat("\nMissing variables:", paste(missing, collapse = ", "), "\n")
}

# Display first few rows
cat("\nFirst few rows of metadata:\n")
print(head(meta[, c(SAMPLE_COL, POP_COL, env_available)], 3))
```

---

## Match and Filter Samples

```{r match_samples}
################################################################################
# MATCH SAMPLES BETWEEN VCF AND METADATA
################################################################################

vcf_samples <- rownames(gt)
meta_samples <- meta[[SAMPLE_COL]]

cat("Matching samples between VCF and metadata...\n\n")
cat("VCF samples (first 5):\n  ", paste(head(vcf_samples, 5), collapse = ", "), "\n")
cat("Metadata samples (first 5):\n  ", paste(head(meta_samples, 5), collapse = ", "), "\n\n")

# Find common samples
common <- intersect(vcf_samples, meta_samples)

cat(sprintf("Matched samples: %s\n", length(common)))
cat(sprintf("  - In VCF only: %s\n", length(setdiff(vcf_samples, meta_samples))))
cat(sprintf("  - In metadata only: %s\n", length(setdiff(meta_samples, vcf_samples))))

if (length(common) == 0) {
  stop("No matching samples found! Check that sample IDs match between VCF and metadata.")
}

# Subset and align
gt_matched <- gt[common, ]
meta_matched <- meta[match(common, meta[[SAMPLE_COL]]), ]

# Verify alignment
if (!all(rownames(gt_matched) == meta_matched[[SAMPLE_COL]])) {
  stop("Sample alignment failed!")
}

cat("\nSamples successfully aligned\n")

# Extract environmental data
env_data <- meta_matched[, env_available, drop = FALSE]

# Remove samples with missing environmental data
complete_idx <- complete.cases(env_data)

cat(sprintf("\nFiltering samples with complete environmental data:\n"))
cat(sprintf("  Keeping: %s/%s samples (%.1f%%)\n", 
            sum(complete_idx), 
            length(complete_idx),
            100 * sum(complete_idx) / length(complete_idx)))

gt_final <- gt_matched[complete_idx, ]
env_final <- env_data[complete_idx, ]
meta_final <- meta_matched[complete_idx, ]

# Standardize environmental variables (mean = 0, SD = 1)
cat("\nStandardizing environmental variables (Z-scores)...\n")
env_scaled <- scale(env_final)

# Show scaling statistics
cat("\nEnvironmental variable statistics after scaling:\n")
env_stats <- data.frame(
  Variable = colnames(env_scaled),
  Mean = round(colMeans(env_scaled), 6),
  SD = round(apply(env_scaled, 2, sd), 6),
  Min = round(apply(env_scaled, 2, min), 2),
  Max = round(apply(env_scaled, 2, max), 2)
)
print(env_stats, row.names = FALSE)
```

---

## Filter SNPs by MAF

```{r filter_maf}
################################################################################
# FILTER SNPS BY MINOR ALLELE FREQUENCY
################################################################################

cat(sprintf("\nCalculating MAF (minimum threshold = %.3f)...\n", MIN_MAF))

# Calculate MAF for each SNP
maf <- apply(gt_final, 2, calc_maf)

# Filter SNPs
keep_snps <- which(!is.na(maf) & maf >= MIN_MAF)

cat(sprintf("\nKeeping: %s/%s SNPs (%.1f%%)\n", 
            format(length(keep_snps), big.mark = ","),
            format(ncol(gt_final), big.mark = ","),
            100 * length(keep_snps) / ncol(gt_final)))

gt_filtered <- gt_final[, keep_snps]
snp_filtered <- snp_data[keep_snps, ]
maf_filtered <- maf[keep_snps]

# MAF distribution
cat("\nMAF distribution:\n")
cat(sprintf("  Min:    %.4f\n", min(maf_filtered, na.rm = TRUE)))
cat(sprintf("  Median: %.4f\n", median(maf_filtered, na.rm = TRUE)))
cat(sprintf("  Mean:   %.4f\n", mean(maf_filtered, na.rm = TRUE)))
cat(sprintf("  Max:    %.4f\n", max(maf_filtered, na.rm = TRUE)))
```

---

## Run WZA Analysis

```{r run_wza}
################################################################################
# COMPUTE WZA SCORES
################################################################################

cat(sprintf("\nComputing WZA scores for %s SNPs...\n", 
            format(ncol(gt_filtered), big.mark = ",")))
cat(sprintf("Using %s environmental variables\n", ncol(env_scaled)))
cat("This may take several minutes depending on dataset size...\n\n")

# Calculate WZA for each SNP
n_snps <- ncol(gt_filtered)
progress_interval <- max(1, floor(n_snps / 20))

wza_list <- vector("list", n_snps)

start_time <- Sys.time()

for (i in 1:n_snps) {
  if (i %% progress_interval == 0) {
    elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
    rate <- i / elapsed
    eta <- (n_snps - i) / rate / 60  # minutes
    
    cat(sprintf("  Progress: %s/%s (%.1f%%) - ETA: %.1f min\n", 
                format(i, big.mark = ","),
                format(n_snps, big.mark = ","),
                100 * i / n_snps,
                eta))
  }
  
  wza_list[[i]] <- calc_wza(gt_filtered[, i], env_scaled)
}

end_time <- Sys.time()
total_time <- difftime(end_time, start_time, units = "mins")

cat(sprintf("\nAnalysis complete! Total time: %.2f minutes\n", as.numeric(total_time)))

# Compile results
wza_results <- data.frame(
  snp_filtered,
  MAF = maf_filtered,
  WZA = sapply(wza_list, function(x) x$wza),
  Mean_abs_Z = sapply(wza_list, function(x) x$mean_abs_z),
  Max_abs_Z = sapply(wza_list, function(x) x$max_abs_z),
  Sum_R2 = sapply(wza_list, function(x) x$sum_r2),
  N_sig_envs = sapply(wza_list, function(x) x$n_sig),
  stringsAsFactors = FALSE
)

# Remove SNPs with NA WZA
wza_results <- wza_results[!is.na(wza_results$WZA), ]

cat(sprintf("\nValid WZA scores: %s SNPs\n", 
            format(nrow(wza_results), big.mark = ",")))
```

---

## Identify Outliers

```{r identify_outliers}
################################################################################
# IDENTIFY OUTLIER SNPS
################################################################################

# Calculate outlier threshold
threshold <- quantile(wza_results$WZA, OUTLIER_PERCENTILE)

wza_results <- wza_results %>%
  mutate(
    WZA_rank = rank(-WZA),
    WZA_percentile = 1 - (WZA_rank / n()),
    Outlier = WZA >= threshold
  )

cat(sprintf("\nOutlier threshold (top %.1f%%): %.4f\n", 
            (1 - OUTLIER_PERCENTILE) * 100, threshold))
cat(sprintf("Outlier SNPs identified: %s (%.2f%%)\n\n", 
            sum(wza_results$Outlier),
            100 * sum(wza_results$Outlier) / nrow(wza_results)))

# WZA score statistics
cat("WZA score distribution:\n")
cat(sprintf("  Min:    %.4f\n", min(wza_results$WZA)))
cat(sprintf("  Q1:     %.4f\n", quantile(wza_results$WZA, 0.25)))
cat(sprintf("  Median: %.4f\n", median(wza_results$WZA)))
cat(sprintf("  Mean:   %.4f\n", mean(wza_results$WZA)))
cat(sprintf("  Q3:     %.4f\n", quantile(wza_results$WZA, 0.75)))
cat(sprintf("  Max:    %.4f\n", max(wza_results$WZA)))
cat(sprintf("  SD:     %.4f\n", sd(wza_results$WZA)))
```

### Top SNPs by WZA Score

```{r top_snps}
# Display top SNPs
top_snps <- head(wza_results %>% 
                   arrange(desc(WZA)) %>%
                   select(SNP_index, CHROM, POS, WZA, MAF, 
                          N_sig_envs, Sum_R2, WZA_percentile), 
                 N_TOP_SNPS)

cat(sprintf("\nTop %d SNPs by WZA score:\n", N_TOP_SNPS))
print(top_snps, row.names = FALSE)
```

---

## Visualizations

### Manhattan Plot

```{r manhattan_plot, fig.width=14, fig.height=6}
################################################################################
# MANHATTAN PLOT
################################################################################

# Calculate cumulative positions for plotting
chrom_summary <- wza_results %>%
  group_by(CHROM) %>%
  summarise(
    max_pos = max(POS),
    .groups = 'drop'
  ) %>%
  arrange(CHROM) %>%
  mutate(
    cum_start = cumsum(c(0, head(max_pos, -1))),
    mid_pos = cum_start + max_pos / 2
  )

plot_data <- wza_results %>%
  left_join(chrom_summary %>% select(CHROM, cum_start), by = "CHROM") %>%
  mutate(
    pos_plot = cum_start + POS,
    chrom_color = as.numeric(factor(CHROM)) %% 2
  )

# Create Manhattan plot
ggplot(plot_data, aes(x = pos_plot, y = WZA)) +
  geom_point(aes(color = factor(chrom_color)), alpha = 0.6, size = 1) +
  geom_hline(yintercept = threshold, color = "red", 
             linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = c("0" = "#2c3e50", "1" = "#95a5a6"), 
                     guide = "none") +
  scale_x_continuous(
    breaks = chrom_summary$mid_pos,
    labels = chrom_summary$CHROM,
    expand = c(0.01, 0.01)
  ) +
  labs(
    title = "WZA Manhattan Plot",
    subtitle = sprintf("Red line = top %.1f%% threshold (WZA ≥ %.3f)", 
                       (1 - OUTLIER_PERCENTILE) * 100, threshold),
    x = "Chromosome",
    y = "WZA Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Distribution of WZA Scores

```{r distribution_plot, fig.width=10, fig.height=6}
################################################################################
# DISTRIBUTION PLOT
################################################################################

ggplot(wza_results, aes(x = WZA)) +
  geom_histogram(bins = 100, fill = "#3498db", color = "black", alpha = 0.7) +
  geom_vline(xintercept = threshold, color = "red", 
             linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median(wza_results$WZA), 
             color = "black", linewidth = 1) +
  annotate("text", x = threshold, y = Inf, 
           label = sprintf("  Outlier threshold\n  (%.3f)", threshold),
           hjust = 0, vjust = 1.5, color = "red", size = 3.5) +
  annotate("text", x = median(wza_results$WZA), y = Inf,
           label = sprintf("  Median\n  (%.3f)", median(wza_results$WZA)),
           hjust = 1.1, vjust = 1.5, color = "black", size = 3.5) +
  labs(
    title = "Distribution of WZA Scores",
    subtitle = "Black = median, Red = outlier threshold",
    x = "WZA Score",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5)
  )
```

### Outliers by Chromosome

```{r chromosome_plot, fig.width=12, fig.height=6}
################################################################################
# CHROMOSOME SUMMARY PLOT
################################################################################

chrom_outliers <- wza_results %>%
  group_by(CHROM) %>%
  summarise(
    Total_SNPs = n(),
    Outliers = sum(Outlier),
    Proportion = Outliers / Total_SNPs,
    .groups = 'drop'
  )

ggplot(chrom_outliers, aes(x = CHROM, y = Outliers)) +
  geom_col(fill = "#e74c3c", alpha = 0.8) +
  geom_text(aes(label = Outliers), vjust = -0.5, size = 3) +
  labs(
    title = "WZA Outlier SNPs by Chromosome",
    subtitle = sprintf("Total outliers: %s", sum(chrom_outliers$Outliers)),
    x = "Chromosome",
    y = "Number of Outlier SNPs"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### WZA vs MAF Relationship

```{r wza_maf_plot, fig.width=10, fig.height=6}
################################################################################
# WZA VS MAF PLOT
################################################################################

ggplot(wza_results, aes(x = MAF, y = WZA)) +
  geom_point(aes(color = Outlier), alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = threshold, color = "red", 
             linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = c("FALSE" = "#95a5a6", "TRUE" = "#e74c3c"),
                     labels = c("Non-outlier", "Outlier")) +
  labs(
    title = "WZA Score vs Minor Allele Frequency",
    x = "Minor Allele Frequency (MAF)",
    y = "WZA Score",
    color = "Classification"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "right"
  )
```

---

## Export Results

```{r export_results}
################################################################################
# SAVE RESULTS
################################################################################

# 1. Complete results
out_complete <- paste0(OUTPUT_PREFIX, "_results.csv")
write.csv(wza_results, out_complete, row.names = FALSE)
cat("Saved complete results:", out_complete, "\n")

# 2. Outliers only
outliers <- wza_results %>%
  filter(Outlier) %>%
  arrange(desc(WZA))

out_outliers <- paste0(OUTPUT_PREFIX, "_outliers.csv")
write.csv(outliers, out_outliers, row.names = FALSE)
cat(sprintf("Saved outliers: %s (%s SNPs)\n", out_outliers, nrow(outliers)))

# 3. Chromosome summary
out_chroms <- paste0(OUTPUT_PREFIX, "_chromosome_summary.csv")
write.csv(chrom_outliers, out_chroms, row.names = FALSE)
cat("Saved chromosome summary:", out_chroms, "\n")

# 4. Save plots
ggsave(paste0(OUTPUT_PREFIX, "_manhattan.png"), 
       width = 14, height = 6, dpi = 300)
cat("Saved Manhattan plot:", paste0(OUTPUT_PREFIX, "_manhattan.png\n"))

ggsave(paste0(OUTPUT_PREFIX, "_distribution.png"), 
       width = 10, height = 6, dpi = 300)
cat("Saved distribution plot:", paste0(OUTPUT_PREFIX, "_distribution.png\n"))

ggsave(paste0(OUTPUT_PREFIX, "_chromosomes.png"), 
       width = 12, height = 6, dpi = 300)
cat("Saved chromosome plot:", paste0(OUTPUT_PREFIX, "_chromosomes.png\n"))

ggsave(paste0(OUTPUT_PREFIX, "_wza_vs_maf.png"),
       width = 10, height = 6, dpi = 300)
cat("Saved WZA vs MAF plot:", paste0(OUTPUT_PREFIX, "_wza_vs_maf.png\n"))
```

---

## Summary Statistics

### Overall Summary

```{r summary_stats}
################################################################################
# SUMMARY STATISTICS
################################################################################

cat("\n")
cat("================================================================================\n")
cat("ANALYSIS SUMMARY\n")
cat("================================================================================\n\n")

cat("Dataset:\n")
cat(sprintf("  Total SNPs analyzed: %s\n", 
            format(nrow(wza_results), big.mark = ",")))
cat(sprintf("  Total samples: %s\n", nrow(gt_final)))
cat(sprintf("  Environmental variables: %s\n", ncol(env_scaled)))
cat("\n")

cat("Outlier detection:\n")
cat(sprintf("  Threshold (top %.1f%%): %.4f\n", 
            (1 - OUTLIER_PERCENTILE) * 100, threshold))
cat(sprintf("  Outlier SNPs: %s (%.2f%%)\n", 
            sum(wza_results$Outlier),
            100 * sum(wza_results$Outlier) / nrow(wza_results)))
cat("\n")

cat("WZA score range:\n")
cat(sprintf("  Minimum: %.4f\n", min(wza_results$WZA)))
cat(sprintf("  Maximum: %.4f\n", max(wza_results$WZA)))
cat(sprintf("  Mean: %.4f\n", mean(wza_results$WZA)))
cat(sprintf("  Median: %.4f\n", median(wza_results$WZA)))
```

### Chromosomes with Most Outliers

```{r top_chromosomes}
top_chroms <- chrom_outliers %>%
  arrange(desc(Outliers)) %>%
  head(10)

cat("\nTop 10 chromosomes by outlier count:\n")
print(top_chroms, row.names = FALSE)
```

---

## Session Information

```{r session_info}
sessionInfo()
```

---

## Interpretation Guide

### Understanding WZA Scores

- **Higher WZA scores** indicate SNPs with strong, consistent associations across multiple environmental variables
- **Outlier SNPs** (top 1% by default) are candidates for being under environmental selection
- **N_sig_envs** shows how many environmental variables had significant associations (p < 0.05)
- **Sum_R2** indicates the total explanatory power across all environmental variables

### Next Steps

1. **Annotate candidate SNPs**: Identify genes near outlier SNPs
2. **Validate results**: Test top candidates in independent datasets
3. **Gene ontology analysis**: Look for enriched biological pathways
4. **Environmental correlations**: Examine which specific environments drive signals
5. **Compare methods**: Cross-validate with other GEA approaches (RDA, LFMM, BayPass)

### References

Booker TR, Yeaman S, Whitlock MC (2021) Variation in recombination rate affects detection of outliers in genome scans under neutrality. *Molecular Ecology* 30:4274-4279. [doi:10.1111/mec.16024](https://doi.org/10.1111/mec.16024)
