---
title: "LFMM: Genotype-Environment Association Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 14,
  fig.height = 8,
  fig.align = 'center'
)
```

# Introduction

## What is LFMM?

**LFMM (Latent Factor Mixed Models)** is a statistical method for identifying genotype-environment associations (GEA) while controlling for population structure and neutral genetic variation.

**Key features:**

- **Controls for population structure**: Uses latent factors (similar to PCA)
- **Multiple testing correction**: FDR and Bonferroni adjustments
- **Scalable**: Works with genome-wide SNP data
- **Flexible**: Tests multiple environmental variables simultaneously

**Applications:**

- Identify loci under environmental selection
- Discover candidate genes for adaptation
- Predict genetic vulnerability to climate change
- Guide conservation and breeding programs

## How LFMM Works

1. **Input**: Genotype matrix (SNPs × samples) + Environmental variables
2. **Latent factors (K)**: Capture population structure and confounding effects
3. **Association testing**: Tests each SNP for correlation with environment
4. **Correction**: Adjusts p-values for multiple testing
5. **Output**: Candidate SNPs associated with environmental variables

# Setup

## Load Required Libraries

```{r load-libraries}
# Load required packages
library(lfmm)          # LFMM analysis
library(vcfR)          # VCF file handling
library(dplyr)         # Data manipulation
library(ggplot2)       # Plotting
library(tidyr)         # Data reshaping
library(knitr)         # Table formatting
library(kableExtra)    # Enhanced tables

# Optional: Color palettes for visualization
library(RColorBrewer)

cat("All required packages loaded successfully\n")
```

## Set File Paths and Parameters

```{r set-parameters}
# ==================== INPUT FILES ====================
VCF_FILE <- "your_genotypes.vcf"                    # Path to VCF file
ENV_CSV <- "your_environmental_data.csv"            # Path to environmental data CSV

# ==================== METADATA COLUMNS ====================
SAMPLE_ID_COLUMN <- "SampleID"                      # Column name with sample IDs
LATITUDE_COLUMN <- "Latitude"                       # Latitude column (optional)
LONGITUDE_COLUMN <- "Longitude"                     # Longitude column (optional)

# Columns to exclude from environmental analysis
EXCLUDE_COLUMNS <- c("SampleID", "Latitude", "Longitude", "Population", "Site")

# ==================== LFMM PARAMETERS ====================
K_VALUE <- 6                                        # Number of latent factors
                                                    # Typically 1-10, based on population structure

# ==================== SIGNIFICANCE THRESHOLDS ====================
P_VALUE_THRESHOLD <- 0.001                          # Nominal p-value threshold
FDR_THRESHOLD <- 0.01                               # FDR threshold
BONFERRONI_ALPHA <- 0.05                            # Alpha for Bonferroni correction

# ==================== OUTPUT OPTIONS ====================
OUTPUT_PREFIX <- "lfmm"                             # Prefix for output files
OUTPUT_DIR <- "lfmm_results"                        # Output directory

# Create output directory
if (!dir.exists(OUTPUT_DIR)) {
  dir.create(OUTPUT_DIR, recursive = TRUE)
}

# Display parameters
param_table <- data.frame(
  Parameter = c("VCF File", "Environmental CSV", "Sample ID Column", 
                "Latent Factors (K)", "P-value Threshold", "FDR Threshold"),
  Value = c(VCF_FILE, ENV_CSV, SAMPLE_ID_COLUMN, 
            K_VALUE, P_VALUE_THRESHOLD, FDR_THRESHOLD)
)

kable(param_table, caption = "Analysis Parameters") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Data Import and Preparation

## Read VCF File

```{r read-vcf}
cat("Reading VCF file:", VCF_FILE, "\n")

# Check if file exists
if (!file.exists(VCF_FILE)) {
  stop("VCF file not found: ", VCF_FILE)
}

# Read VCF
vcf <- read.vcfR(VCF_FILE)

cat("VCF loaded successfully\n")
cat("  SNPs:", nrow(vcf@fix), "\n")
cat("  Samples:", ncol(vcf@gt) - 1, "\n")

# Extract genotypes as numeric matrix (0, 1, 2 for diploid genotypes)
gt_numeric <- extract.gt(vcf, element = "GT", as.numeric = TRUE)

cat("  Genotype matrix dimensions:", dim(gt_numeric), "\n")

# Create VCF summary
vcf_summary <- data.frame(
  Metric = c("Total SNPs", "Total Samples", "Missing Genotypes"),
  Value = c(
    nrow(gt_numeric),
    ncol(gt_numeric),
    paste0(round(100 * sum(is.na(gt_numeric)) / length(gt_numeric), 2), "%")
  )
)

kable(vcf_summary, caption = "VCF Data Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Read Environmental Data

```{r read-environment}
cat("Reading environmental data:", ENV_CSV, "\n")

# Check if file exists
if (!file.exists(ENV_CSV)) {
  stop("Environmental CSV file not found: ", ENV_CSV)
}

# Read CSV
env_data <- read.csv(ENV_CSV, header = TRUE, stringsAsFactors = FALSE)

cat("Environmental data loaded\n")
cat("  Samples:", nrow(env_data), "\n")
cat("  Columns:", ncol(env_data), "\n")

# Display first few rows
cat("\nFirst 6 rows of environmental data:\n")
kable(head(env_data), caption = "Environmental Data Preview") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%")

# Check for missing coordinates (if applicable)
if (LATITUDE_COLUMN %in% colnames(env_data) && LONGITUDE_COLUMN %in% colnames(env_data)) {
  coords <- env_data[, c(LATITUDE_COLUMN, LONGITUDE_COLUMN)]
  missing_coords <- which(!complete.cases(coords))
  
  if (length(missing_coords) > 0) {
    cat("\nWARNING:", length(missing_coords), "samples have missing coordinates\n")
    cat("Row indices:", paste(head(missing_coords, 20), collapse = ", "), 
        ifelse(length(missing_coords) > 20, "...", ""), "\n")
    
    # Option to remove samples with missing coordinates
    env_data_clean <- env_data[-missing_coords, ]
    cat("Removed", length(missing_coords), "samples with missing coordinates\n")
  } else {
    env_data_clean <- env_data
    cat("All samples have complete coordinates\n")
  }
} else {
  env_data_clean <- env_data
}

cat("Clean environmental data:", nrow(env_data_clean), "samples\n")
```

## Match Samples Between VCF and Environmental Data

```{r match-samples}
cat("Matching samples between VCF and environmental data...\n")

# Get sample IDs from both datasets
vcf_samples <- colnames(gt_numeric)
csv_samples <- env_data_clean[[SAMPLE_ID_COLUMN]]

cat("VCF samples:", length(vcf_samples), "\n")
cat("CSV samples:", length(csv_samples), "\n")

# Find common samples
common_samples <- intersect(vcf_samples, csv_samples)

cat("Common samples:", length(common_samples), "\n")

if (length(common_samples) == 0) {
  stop("ERROR: No matching samples found!\n",
       "  VCF samples (first 5): ", paste(head(vcf_samples, 5), collapse = ", "), "\n",
       "  CSV samples (first 5): ", paste(head(csv_samples, 5), collapse = ", "))
}

# Create matching summary
match_summary <- data.frame(
  Category = c("VCF samples", "CSV samples", "Matched samples", 
               "VCF-only", "CSV-only"),
  Count = c(
    length(vcf_samples),
    length(csv_samples),
    length(common_samples),
    length(vcf_samples) - length(common_samples),
    length(csv_samples) - length(common_samples)
  )
)

kable(match_summary, caption = "Sample Matching Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Subset both datasets to common samples
gt_matched <- gt_numeric[, colnames(gt_numeric) %in% common_samples, drop = FALSE]
csv_matched <- env_data_clean[env_data_clean[[SAMPLE_ID_COLUMN]] %in% common_samples, ]

# Ensure same sample order
csv_matched <- csv_matched[match(colnames(gt_matched), csv_matched[[SAMPLE_ID_COLUMN]]), ]

cat("\nMatched datasets created\n")
cat("  Genotype matrix:", dim(gt_matched), "\n")
cat("  Environmental data:", dim(csv_matched), "\n")
```

## Prepare Genotype Matrix for LFMM

```{r prepare-genotype}
cat("Preparing genotype matrix for LFMM...\n")

# Impute missing genotypes with column (SNP) means
genotype_data <- apply(gt_matched, 2, function(x) {
  ifelse(is.na(x), mean(x, na.rm = TRUE), x)
})

# Transpose: LFMM expects individuals as rows, SNPs as columns
genotype_data <- t(genotype_data)

cat("Final genotype matrix dimensions (individuals × SNPs):", dim(genotype_data), "\n")
cat("Missing values after imputation:", sum(is.na(genotype_data)), "\n")
```

# Multiple Environmental Variables Analysis

## Loop Through All Environmental Variables

```{r loop-environmental-variables}
cat("Identifying environmental columns...\n")

# Identify environmental columns (exclude metadata)
env_cols <- setdiff(colnames(csv_matched), EXCLUDE_COLUMNS)

cat("Environmental variables to analyze:", length(env_cols), "\n")
cat("Variables:", paste(env_cols, collapse = ", "), "\n\n")

# Initialize list to store results
all_lfmm_results <- list()

# Loop through all environmental variables
for (env_var in env_cols) {
  cat("Processing environmental variable:", env_var, "\n")
  
  # Extract current environmental variable
  env_matrix <- as.matrix(csv_matched[, env_var, drop = FALSE])
  
  # Handle missing values - impute with mean
  env_matrix[is.na(env_matrix)] <- mean(env_matrix, na.rm = TRUE)
  
  # Check if there's variation in the environmental variable
  if (var(env_matrix, na.rm = TRUE) == 0) {
    cat("  Skipping", env_var, "- no variation in data\n")
    next
  }
  
  # Run LFMM
  tryCatch({
    mod.lfmm <- lfmm_ridge(Y = genotype_data, X = env_matrix, K = K_VALUE)
    pv <- lfmm_test(Y = genotype_data, X = env_matrix, lfmm = mod.lfmm)
    
    # Store results with chromosome and position information
    lfmm_results <- data.frame(
      SNP = 1:ncol(genotype_data), 
      p_value = pv$pvalue[, 1], 
      log_p = -log10(pv$pvalue[, 1]),
      env_variable = env_var
    )
    
    # Add chromosome and position information
    lfmm_results <- cbind(
      lfmm_results, 
      chromosome = as.factor(vcf@fix[, "CHROM"]), 
      pos = as.numeric(vcf@fix[, "POS"])
    )
    
    # Add results to list
    all_lfmm_results[[env_var]] <- lfmm_results
    
    cat("  Completed LFMM for", env_var, "\n")
    cat("    Significant SNPs (p < 0.05):", 
        sum(lfmm_results$p_value < 0.05, na.rm = TRUE), "\n")
    
  }, error = function(e) {
    cat("  Error processing", env_var, ":", e$message, "\n")
  })
}

cat("\nAnalysis complete!\n")
cat("Total environmental variables processed:", length(all_lfmm_results), "\n")
```

## Combine Results

```{r combine-results}
if (length(all_lfmm_results) > 0) {
  # Combine all results into one data frame
  combined_results <- do.call(rbind, all_lfmm_results)
  
  cat("Combined results dimensions:", dim(combined_results), "\n")
  
  # Calculate FDR and other statistics per environmental variable
  combined_results <- combined_results %>%
    group_by(env_variable) %>%
    mutate(
      bonferroni_threshold = -log10(BONFERRONI_ALPHA / n()),
      fdr_p_value = p.adjust(p_value, method = "BH"),
      log_fdr_p = -log10(fdr_p_value)
    ) %>%
    ungroup()
  
  # Add chromosome numeric mapping for plotting
  unique_chroms <- unique(combined_results$chromosome)
  chromosome_mapping <- setNames(1:length(unique_chroms), unique_chroms)
  
  combined_results <- combined_results %>%
    mutate(chromosome_numeric = chromosome_mapping[as.character(chromosome)])
  
  cat("Combined results prepared for visualization\n")
  
} else {
  stop("No environmental variables were successfully processed.")
}
```

## Summary Statistics

```{r summary-statistics}
# Calculate summary statistics for each environmental variable
summary_stats <- combined_results %>%
  group_by(env_variable) %>%
  summarise(
    total_snps = n(),
    bonferroni_significant = sum(p_value < (BONFERRONI_ALPHA / n()), na.rm = TRUE),
    fdr_significant = sum(fdr_p_value < FDR_THRESHOLD, na.rm = TRUE),
    p_001_significant = sum(p_value < P_VALUE_THRESHOLD, na.rm = TRUE),
    mean_log_p = round(mean(log_p, na.rm = TRUE), 3),
    max_log_p = round(max(log_p, na.rm = TRUE), 3),
    .groups = 'drop'
  )

kable(summary_stats, 
      caption = "Summary of Significant SNPs per Environmental Variable") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Visualization

## Faceted Manhattan Plots

```{r faceted-manhattan, fig.width=16, fig.height=12}
# Create faceted Manhattan plot for all environmental variables
p_faceted <- ggplot(combined_results, 
                    aes(x = SNP, y = log_p, color = as.factor(chromosome_numeric))) +
  geom_point(alpha = 0.6, size = 0.5) +
  geom_hline(data = combined_results %>% 
               group_by(env_variable) %>% 
               slice(1) %>% 
               ungroup(), 
             aes(yintercept = bonferroni_threshold), 
             color = "red", linetype = "dashed", alpha = 0.7) +
  geom_hline(yintercept = -log10(FDR_THRESHOLD), 
             color = "purple", linetype = "dashed", alpha = 0.7) +
  facet_wrap(~ env_variable, scales = "free_y", ncol = 3) +
  labs(
    title = "Manhattan Plots for All Environmental Variables",
    x = "SNP Index",
    y = "-log10(p-value)",
    color = "Chromosome",
    caption = "Red: Bonferroni threshold (per variable); Purple: FDR threshold"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 9, face = "bold"),
    plot.caption = element_text(size = 9, hjust = 0),
    panel.spacing = unit(0.5, "lines")
  ) +
  scale_color_brewer(palette = "Set3", guide = guide_legend(nrow = 2))

print(p_faceted)
```

## Summary Bar Plot

```{r summary-barplot, fig.width=12, fig.height=6}
# Create summary plot showing number of significant SNPs per environmental variable
summary_plot <- summary_stats %>%
  select(env_variable, bonferroni_significant, fdr_significant, p_001_significant) %>%
  pivot_longer(cols = -env_variable, names_to = "threshold_type", values_to = "count") %>%
  mutate(threshold_type = case_when(
    threshold_type == "bonferroni_significant" ~ "Bonferroni",
    threshold_type == "fdr_significant" ~ paste0("FDR < ", FDR_THRESHOLD),
    threshold_type == "p_001_significant" ~ paste0("p < ", P_VALUE_THRESHOLD)
  )) %>%
  ggplot(aes(x = reorder(env_variable, -count), y = count, fill = threshold_type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(
    title = "Number of Significant SNPs by Environmental Variable",
    x = "Environmental Variable",
    y = "Number of Significant SNPs",
    fill = "Significance Threshold"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set2")

print(summary_plot)
```

# Save Results

```{r save-results}
cat("Saving results to disk...\n")

# Save combined results
write.csv(combined_results, 
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_all_variables.csv")), 
          row.names = FALSE)
cat("Saved:", file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_all_variables.csv")), "\n")

# Save summary statistics
write.csv(summary_stats, 
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_summary.csv")), 
          row.names = FALSE)
cat("Saved:", file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_summary.csv")), "\n")

# Save individual result files for each environmental variable
for (env_var in names(all_lfmm_results)) {
  filename <- file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_", env_var, ".csv"))
  write.csv(all_lfmm_results[[env_var]], filename, row.names = FALSE)
}
cat("Saved individual files for each environmental variable\n")

# Save plots
ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_manhattan_faceted.png")), 
       plot = p_faceted, 
       width = 16, height = 12, dpi = 300, units = "in")
cat("Saved faceted Manhattan plot\n")

ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_summary_barplot.png")), 
       plot = summary_plot, 
       width = 12, height = 6, dpi = 300, units = "in")
cat("Saved summary bar plot\n")

cat("\nAll results saved successfully!\n")
cat("Output directory:", OUTPUT_DIR, "\n")
```

# Session Information

```{r session-info}
sessionInfo()
```

# References

1. Caye, K., et al. (2019). LFMM 2: Fast and accurate inference of gene-environment associations in genome-wide studies. *Molecular Biology and Evolution*, 36(4), 852-860.

2. Frichot, E., & François, O. (2015). LEA: An R package for landscape and ecological association studies. *Methods in Ecology and Evolution*, 6(8), 925-929.

---

**Analysis completed:** `r Sys.time()`
