---
title: "Redundancy Analysis (RDA) for Genomic Data"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  fig.align = 'center'
)
```

# Introduction

## What is RDA?

Redundancy Analysis (RDA) is a constrained ordination method that identifies associations between genomic variation (from SNP data) and environmental or climate variables. 

**Key applications:**

- Identifies relationships between genomic variation and environmental predictors
- Detects candidate loci potentially under environmental selection
- Provides visualization of genotype-environment associations
- Tests statistical significance through permutation tests

## When to Use RDA

Use RDA when you have:

- Genomic data (SNPs) from multiple individuals/populations
- Environmental or climate data for sample collection locations
- Interest in identifying adaptive genetic variation
- Questions about genotype-environment associations

# Setup

## Install and Load Required Packages

```{r load-packages}
# Install and load required packages
required_packages <- c("vcfR", "vegan", "ggplot2", "dplyr", "corrplot", "knitr", "kableExtra")

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

**Package purposes:**

- `vcfR`: Reading and handling VCF files
- `vegan`: Performing RDA and multivariate analyses
- `ggplot2`: Creating publication-quality plots
- `dplyr`: Data manipulation
- `corrplot`: Visualizing correlation matrices
- `knitr` & `kableExtra`: Creating formatted tables

## Set Analysis Parameters

```{r set-parameters}
# === FILE INPUTS ===
VCF_FILE <- "your_genotypes.vcf"
CLIMATE_CSV <- "your_environmental_data.csv"
SAMPLE_ID_COLUMN <- "SampleID"
EXCLUDE_COLUMNS <- c("SampleID", "Longitude", "Latitude")

# === ANALYSIS PARAMETERS ===
CORRELATION_CUTOFF <- 0.9        # Remove highly correlated variables
OUTLIER_QUANTILE <- 0.999        # 99.9th percentile for candidate SNPs
N_PERMUTATIONS <- 100            # Permutations for significance testing
MAX_MISSING_DATA <- 0.30         # Maximum 30% missing data per SNP

# === OUTPUT OPTIONS ===
OUTPUT_PREFIX <- "rda"
OUTPUT_DIR <- "rda_results"

# Create output directory
if (!dir.exists(OUTPUT_DIR)) {
  dir.create(OUTPUT_DIR)
}
```

# Input Data Requirements

## VCF File (Genotype Data)

Your VCF file should contain:

- SNP genotype calls for all samples
- Chromosome and position information
- Quality-filtered SNPs

**Example VCF structure:**
```
#CHROM  POS     ID      REF     ALT     QUAL    FILTER  INFO    FORMAT  Sample1 Sample2 Sample3
Chr1    1000    .       A       G       99      PASS    .       GT      0/0     0/1     1/1
Chr1    2000    .       C       T       99      PASS    .       GT      0/1     0/1     0/0
```

## Environmental/Climate CSV File

The CSV file must include:

- **First column**: Sample IDs (must match VCF sample names)
- **Optional columns**: Longitude, Latitude (will be excluded from analysis)
- **Remaining columns**: Environmental/climate variables (continuous)

**Example CSV structure:**
```
SampleID,Longitude,Latitude,Temperature,Precipitation,Elevation
Sample1,-120.5,45.2,15.3,800,1200
Sample2,-118.3,44.8,16.1,750,950
Sample3,-122.1,46.5,14.2,900,1450
```

# Helper Functions

```{r helper-functions}
# Function to identify highly correlated variables
identify_highly_correlated <- function(cor_matrix, cutoff = 0.9) {
  upper_tri <- upper.tri(cor_matrix)
  high_cor_pairs <- which(abs(cor_matrix) > cutoff & upper_tri, arr.ind = TRUE)
  
  if (nrow(high_cor_pairs) == 0) {
    return(integer(0))
  }
  
  vars_to_remove <- c()
  
  for (i in 1:nrow(high_cor_pairs)) {
    var1 <- high_cor_pairs[i, 1]
    var2 <- high_cor_pairs[i, 2]
    
    mean_cor_var1 <- mean(abs(cor_matrix[var1, -var1]), na.rm = TRUE)
    mean_cor_var2 <- mean(abs(cor_matrix[var2, -var2]), na.rm = TRUE)
    
    if (mean_cor_var1 > mean_cor_var2) {
      vars_to_remove <- c(vars_to_remove, var1)
    } else {
      vars_to_remove <- c(vars_to_remove, var2)
    }
  }
  
  return(unique(vars_to_remove))
}
```

# Data Import and Processing

## Read VCF File

```{r read-vcf}
cat("Step 1: Reading VCF file...\n")

# Check if file exists
if (!file.exists(VCF_FILE)) {
  stop("Error: VCF file not found at: ", VCF_FILE)
}

# Read VCF
vcf <- read.vcfR(VCF_FILE)

cat("  - Loaded", nrow(vcf@fix), "SNPs from", ncol(vcf@gt) - 1, "samples\n")
```

## Extract Genotype Matrix

```{r extract-genotypes}
cat("Step 2: Extracting genotype data...\n")

# Extract numeric genotype matrix
gt_numeric <- extract.gt(vcf, element = "GT", as.numeric = TRUE)

cat("  - Genotype matrix dimensions:", nrow(gt_numeric), "SNPs x", 
    ncol(gt_numeric), "samples\n")
```

## Filter SNPs by Missing Data

```{r filter-snps}
cat("Step 3: Filtering SNPs by missing data threshold...\n")
cat("  - Original SNPs:", nrow(gt_numeric), "\n")

# Calculate missing data proportion per SNP
missing_per_snp <- rowSums(is.na(gt_numeric))
missing_proportion <- missing_per_snp / ncol(gt_numeric)

# Filter SNPs
snps_pass_filter <- missing_proportion <= MAX_MISSING_DATA

gt_numeric <- gt_numeric[snps_pass_filter, , drop = FALSE]
vcf@fix <- vcf@fix[snps_pass_filter, , drop = FALSE]
vcf@gt <- vcf@gt[snps_pass_filter, , drop = FALSE]

cat("  - Maximum missing data allowed:", MAX_MISSING_DATA * 100, "%\n")
cat("  - SNPs passing filter (≤", MAX_MISSING_DATA * 100, "% missing):", 
    sum(snps_pass_filter), "\n")
cat("  - SNPs removed (>", MAX_MISSING_DATA * 100, "% missing):", 
    sum(!snps_pass_filter), "\n")
cat("  - Retained", nrow(gt_numeric), "SNPs\n")
cat("  - Average missing data in retained SNPs:", 
    round(mean(rowSums(is.na(gt_numeric)) / ncol(gt_numeric)) * 100, 2), "%\n")
```

## Read Environmental Data

```{r read-environmental}
cat("Step 4: Reading climate/environmental data...\n")

# Check if file exists
if (!file.exists(CLIMATE_CSV)) {
  stop("Error: Climate CSV file not found at: ", CLIMATE_CSV)
}

# Read environmental data
env_data <- read.csv(CLIMATE_CSV, header = TRUE, stringsAsFactors = FALSE)

cat("  - Loaded data for", nrow(env_data), "samples with", ncol(env_data), "columns\n")

# Validate that sample ID column exists
if (!SAMPLE_ID_COLUMN %in% colnames(env_data)) {
  stop("Error: Sample ID column '", SAMPLE_ID_COLUMN, "' not found in CSV file")
}

# Display first few rows
kable(head(env_data, 5), caption = "First 5 rows of environmental data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Match Samples Between VCF and Environmental Data

```{r match-samples}
cat("Step 5: Matching samples between VCF and climate data...\n")

vcf_samples <- colnames(gt_numeric)
csv_samples <- env_data[[SAMPLE_ID_COLUMN]]

cat("  - VCF samples (first 5):", paste(head(vcf_samples, 5), collapse = ", "), "\n")
cat("  - CSV samples (first 5):", paste(head(csv_samples, 5), collapse = ", "), "\n\n")

# Find common samples
common_samples <- intersect(vcf_samples, csv_samples)

cat("  - Found", length(common_samples), "samples with both genetic and environmental data\n")
cat("  - VCF-only samples:", length(vcf_samples) - length(common_samples), "\n")
cat("  - CSV-only samples:", length(csv_samples) - length(common_samples), "\n")

# Subset to common samples
gt_matched <- gt_numeric[, common_samples, drop = FALSE]
env_matched <- env_data[match(common_samples, csv_samples), ]

cat("  - Final dataset:", length(common_samples), "samples x", nrow(gt_matched), "SNPs\n")
```

## Prepare Environmental Variables

```{r prepare-env}
cat("Step 6: Preparing environmental variables...\n")

# Extract environmental variables (exclude metadata columns)
env_vars <- env_matched[, !colnames(env_matched) %in% EXCLUDE_COLUMNS, drop = FALSE]

# Convert to numeric
env_vars <- as.data.frame(lapply(env_vars, as.numeric))
rownames(env_vars) <- common_samples

cat("  - Initial environmental variables:", ncol(env_vars), "\n")

# Check for missing data
missing_by_var <- colSums(is.na(env_vars)) / nrow(env_vars)
cat("  - Variables with missing data:\n")
if (any(missing_by_var > 0)) {
  print(round(missing_by_var[missing_by_var > 0] * 100, 2))
}

# Remove variables with too much missing data
missing_threshold <- 0.1  # 10%
env_vars <- env_vars[, missing_by_var <= missing_threshold, drop = FALSE]

cat("  - Retained", ncol(env_vars), "variables after removing those with >10% missing data\n")

# Display summary statistics
kable(
  data.frame(
    Variable = colnames(env_vars),
    Mean = round(colMeans(env_vars, na.rm = TRUE), 2),
    SD = round(apply(env_vars, 2, sd, na.rm = TRUE), 2),
    Min = round(apply(env_vars, 2, min, na.rm = TRUE), 2),
    Max = round(apply(env_vars, 2, max, na.rm = TRUE), 2),
    Missing_pct = round(colSums(is.na(env_vars)) / nrow(env_vars) * 100, 2)
  ),
  caption = "Summary statistics for environmental variables"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Check for Multicollinearity

```{r check-correlation}
cat("Step 7: Checking for multicollinearity...\n")

# Calculate correlation matrix
env_cor <- cor(env_vars, use = "complete.obs")

# Visualize correlations
corrplot(env_cor, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, 
         title = "Environmental Variable Correlations (Before Filtering)",
         mar = c(0, 0, 2, 0), addCoef.col = "black", number.cex = 0.7)

# Identify highly correlated variables
vars_to_remove <- identify_highly_correlated(env_cor, CORRELATION_CUTOFF)

if (length(vars_to_remove) > 0) {
  cat("  - Found", length(vars_to_remove), "highly correlated variables to remove:\n")
  cat("   ", colnames(env_vars)[vars_to_remove], "\n")
  env_vars_final <- env_vars[, -vars_to_remove, drop = FALSE]
} else {
  cat("  - No highly correlated variables found\n")
  env_vars_final <- env_vars
}

cat("  - Retained", ncol(env_vars_final), "environmental variables for RDA\n")

# Visualize final correlations
if (ncol(env_vars_final) > 1) {
  env_cor_final <- cor(env_vars_final, use = "complete.obs")
  corrplot(env_cor_final, method = "color", type = "upper", 
           tl.col = "black", tl.srt = 45, 
           title = "Environmental Variable Correlations (After Filtering)",
           mar = c(0, 0, 2, 0), addCoef.col = "black", number.cex = 0.7)
}
```

## Scale Data for RDA

```{r scale-data}
cat("Step 8: Scaling data for RDA...\n")

# Scale environmental variables (mean = 0, sd = 1)
env_vars_scaled <- scale(env_vars_final, center = TRUE, scale = TRUE)

# Prepare genotype data
genotype_data <- t(gt_matched)  # Transpose: samples as rows, SNPs as columns

# Impute missing genotypes with mean
genotype_data[is.na(genotype_data)] <- mean(genotype_data, na.rm = TRUE)

# Ensure sample order matches between genotype and environmental data
genotype_data <- genotype_data[match(rownames(env_vars_scaled), rownames(genotype_data)), ]

cat("  - Genotype matrix:", nrow(genotype_data), "samples x", 
    ncol(genotype_data), "SNPs\n")
cat("  - Environmental matrix:", nrow(env_vars_scaled), "samples x", 
    ncol(env_vars_scaled), "variables\n")
```

# Redundancy Analysis (RDA)

## Run RDA

```{r run-rda}
cat("Step 9: Running RDA...\n")

# Perform RDA
rda_result <- rda(genotype_data ~ ., data = as.data.frame(env_vars_scaled))

# Print summary
cat("\n=== RDA Summary ===\n")
summary(rda_result)
```

## Test Significance

```{r test-significance}
cat("Step 10: Testing significance with permutation tests...\n")

# Overall model significance
anova_overall <- anova.cca(rda_result, permutations = N_PERMUTATIONS)

# Axis significance
anova_axes <- anova.cca(rda_result, by = "axis", permutations = N_PERMUTATIONS)

# Individual variable significance
anova_terms <- anova.cca(rda_result, by = "terms", permutations = N_PERMUTATIONS)
```

### Overall Model Significance

```{r show-overall}
kable(anova_overall, caption = "Overall RDA Model Significance") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Axis Significance

```{r show-axes}
kable(anova_axes, caption = "RDA Axis Significance") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Environmental Variable Significance

```{r show-terms}
kable(anova_terms, caption = "Environmental Variable Significance") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Calculate Variance Explained

```{r variance-explained}
cat("Step 11: Calculating variance components...\n")

# Extract variance components
total_var <- rda_result$tot.chi
constrained_var <- rda_result$CCA$tot.chi
unconstrained_var <- rda_result$CA$tot.chi
var_explained <- constrained_var / total_var

# Variance per axis
eigenvalues <- rda_result$CCA$eig
axis_var <- (eigenvalues / total_var) * 100

# Create summary table
variance_table <- data.frame(
  Component = c("Total Variance", "Constrained Variance", 
                "Unconstrained Variance", "Proportion Explained"),
  Value = c(
    round(total_var, 2),
    round(constrained_var, 2), 
    round(unconstrained_var, 2),
    paste0(round(var_explained * 100, 2), "%")
  )
)

kable(variance_table, caption = "RDA Variance Components") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Axis variance
axis_variance_table <- data.frame(
  Axis = paste0("RDA", 1:min(5, length(axis_var))),
  Variance_pct = round(axis_var[1:min(5, length(axis_var))], 2)
)

kable(axis_variance_table, caption = "Variance Explained by RDA Axes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Visualization

## RDA Biplot

```{r rda-biplot, fig.width=10, fig.height=8}
cat("Step 12: Creating RDA biplot...\n")

# Extract sample scores
sample_scores <- scores(rda_result, choices = 1:2, display = "sites", scaling = 1)
rda_df <- data.frame(
  Sample = rownames(sample_scores),
  RDA1 = sample_scores[, 1],
  RDA2 = sample_scores[, 2]
)

# Extract environmental variable loadings
env_scores <- scores(rda_result, choices = 1:2, display = "bp", scaling = 1)
env_arrows <- data.frame(
  Variable = rownames(env_scores),
  RDA1 = env_scores[, 1],
  RDA2 = env_scores[, 2]
)

# Create biplot
rda_plot <- ggplot(rda_df, aes(x = RDA1, y = RDA2)) +
  geom_point(size = 2.5, alpha = 0.7, color = "steelblue") +
  geom_segment(data = env_arrows, 
               aes(x = 0, y = 0, xend = RDA1 * 3, yend = RDA2 * 3),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "red", size = 1, alpha = 0.8) +
  geom_text(data = env_arrows, 
            aes(x = RDA1 * 3.2, y = RDA2 * 3.2, label = Variable),
            color = "black", size = 3.5, fontface = "bold") +
  labs(
    title = "RDA Biplot: Genomic Variation Constrained by Environment",
    subtitle = paste0("Total variance explained: ", round(var_explained * 100, 2), "%"),
    x = paste0("RDA1 (", round(axis_var[1], 2), "%)"),
    y = paste0("RDA2 (", round(axis_var[2], 2), "%)")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  coord_fixed()

print(rda_plot)
```

## RDA Biplot Colored by Environmental Variable

```{r rda-biplot-colored, fig.width=10, fig.height=8}
# Create biplot colored by first environmental variable
if (ncol(env_vars_final) > 0) {
  first_env_var <- colnames(env_vars_final)[1]
  rda_df_colored <- cbind(rda_df, 
                          env_vars_final[match(rda_df$Sample, rownames(env_vars_final)), , drop = FALSE])
  
  rda_plot_colored <- ggplot(rda_df_colored, aes(x = RDA1, y = RDA2)) +
    geom_point(aes_string(color = first_env_var), size = 2.5, alpha = 0.8) +
    geom_segment(data = env_arrows, 
                 aes(x = 0, y = 0, xend = RDA1 * 3, yend = RDA2 * 3),
                 arrow = arrow(length = unit(0.3, "cm")), 
                 color = "black", size = 1, alpha = 0.8) +
    geom_text(data = env_arrows, 
              aes(x = RDA1 * 3.2, y = RDA2 * 3.2, label = Variable),
              color = "black", size = 3.5, fontface = "bold") +
    scale_color_gradient(low = "blue", high = "red") +
    labs(
      title = paste("RDA Biplot Colored by", first_env_var),
      subtitle = paste0("Total variance explained: ", round(var_explained * 100, 2), "%"),
      x = paste0("RDA1 (", round(axis_var[1], 2), "%)"),
      y = paste0("RDA2 (", round(axis_var[2], 2), "%)")
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 14),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      legend.position = "bottom"
    ) +
    coord_fixed()
  
  print(rda_plot_colored)
}
```

# Candidate SNP Detection

## Identify Outlier SNPs

```{r identify-outliers}
cat("Step 13: Identifying candidate SNPs using Mahalanobis distance...\n")

# Extract SNP loadings
species_scores <- scores(rda_result, choices = 1:2, display = "species", scaling = 1)
loadings_subset <- species_scores[, 1:2]

# Calculate Mahalanobis distances
maha_dist <- mahalanobis(loadings_subset, 
                         center = colMeans(loadings_subset), 
                         cov = cov(loadings_subset))

# Identify outliers
outlier_threshold <- quantile(maha_dist, OUTLIER_QUANTILE)
outlier_snps <- which(maha_dist > outlier_threshold)

cat("  - Identified", length(outlier_snps), "candidate SNPs\n")
cat("  - Outlier threshold (", OUTLIER_QUANTILE * 100, "th percentile): ", 
    round(outlier_threshold, 3), "\n", sep = "")

# Create outlier data frame
outlier_df <- data.frame(
  SNP_index = outlier_snps,
  RDA1_loading = species_scores[outlier_snps, 1],
  RDA2_loading = species_scores[outlier_snps, 2],
  Mahalanobis_distance = maha_dist[outlier_snps],
  chromosome = vcf@fix[outlier_snps, "CHROM"],
  position = as.numeric(vcf@fix[outlier_snps, "POS"])
)

# Display top candidates
kable(head(outlier_df[order(-outlier_df$Mahalanobis_distance), ], 10),
      caption = "Top 10 Candidate SNPs by Mahalanobis Distance",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## SNP Loading Plot

```{r snp-loadings, fig.width=10, fig.height=8}
# Create loading data frame
loading_df <- data.frame(
  SNP = 1:nrow(species_scores),
  RDA1 = species_scores[, 1],
  RDA2 = species_scores[, 2],
  Outlier = 1:nrow(species_scores) %in% outlier_snps
)

# Create loading plot
loading_plot <- ggplot(loading_df, aes(x = RDA1, y = RDA2)) +
  geom_point(aes(color = Outlier), size = 1, alpha = 0.6) +
  scale_color_manual(values = c("gray70", "red"), 
                     labels = c("Background SNPs", "Candidate SNPs")) +
  labs(
    title = "SNP Loadings on RDA Axes",
    subtitle = paste("Candidate SNPs (n =", length(outlier_snps), ") highlighted in red"),
    x = paste0("RDA1 (", round(axis_var[1], 2), "%)"),
    y = paste0("RDA2 (", round(axis_var[2], 2), "%)")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  ) +
  coord_fixed()

print(loading_plot)
```

## Manhattan-style Plot of Candidate SNPs

```{r manhattan-plot, fig.width=12, fig.height=6}
# Prepare data for Manhattan plot
manhattan_df <- data.frame(
  SNP_index = 1:nrow(species_scores),
  chromosome = vcf@fix[, "CHROM"],
  position = as.numeric(vcf@fix[, "POS"]),
  Mahalanobis_distance = maha_dist,
  Outlier = 1:nrow(species_scores) %in% outlier_snps
)

# Convert chromosome to numeric if needed
manhattan_df$chromosome <- as.factor(manhattan_df$chromosome)
manhattan_df$chr_numeric <- as.numeric(manhattan_df$chromosome)

# Create Manhattan plot
manhattan_plot <- ggplot(manhattan_df, aes(x = SNP_index, y = Mahalanobis_distance)) +
  geom_point(aes(color = Outlier), size = 1.5, alpha = 0.7) +
  geom_hline(yintercept = outlier_threshold, linetype = "dashed", 
             color = "red", size = 1) +
  scale_color_manual(values = c("gray50", "red"),
                     labels = c("Background SNPs", "Candidate SNPs")) +
  labs(
    title = "Manhattan Plot: Mahalanobis Distance for All SNPs",
    subtitle = paste("Threshold =", round(outlier_threshold, 2), 
                     "| Candidate SNPs =", length(outlier_snps)),
    x = "SNP Index",
    y = "Mahalanobis Distance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "bottom",
    legend.title = element_blank()
  )

print(manhattan_plot)
```

# Save Results

```{r save-results}
cat("Step 14: Saving results...\n")

# Save RData object
save(rda_result, file = file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_results.RData")))
cat("  - Saved RData object\n")

# Save plots
ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_biplot.png")), 
       rda_plot, width = 10, height = 8, dpi = 300)
cat("  - Saved RDA biplot\n")

if (exists("rda_plot_colored")) {
  ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_biplot_colored.png")), 
         rda_plot_colored, width = 10, height = 8, dpi = 300)
  cat("  - Saved colored RDA biplot\n")
}

ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_snp_loadings.png")), 
       loading_plot, width = 10, height = 8, dpi = 300)
cat("  - Saved SNP loadings plot\n")

ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_manhattan.png")), 
       manhattan_plot, width = 12, height = 6, dpi = 300)
cat("  - Saved Manhattan plot\n")

# Save candidate SNPs table
write.csv(outlier_df, 
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_outlier_snps.csv")), 
          row.names = FALSE)
cat("  - Saved candidate SNPs table\n")

# Save environmental correlation matrix
if (ncol(env_vars_final) > 1) {
  png(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_environmental_correlations.png")), 
      width = 8, height = 8, units = "in", res = 300)
  corrplot(env_cor_final, method = "color", type = "upper", 
           tl.col = "black", tl.srt = 45, 
           title = "Environmental Variable Correlations",
           mar = c(0, 0, 2, 0))
  dev.off()
  cat("  - Saved environmental correlation plot\n")
}

# Create and save summary table
rda_summary_table <- data.frame(
  Metric = c("Total Variance", "Constrained Variance", "Unconstrained Variance", 
             "Proportion Explained", "Number of Environmental Variables", 
             "Number of Samples", "Number of SNPs", "Candidate SNPs"),
  Value = c(
    round(total_var, 2),
    round(constrained_var, 2), 
    round(unconstrained_var, 2),
    paste0(round(var_explained * 100, 2), "%"),
    ncol(env_vars_scaled),
    nrow(genotype_data),
    ncol(genotype_data),
    length(outlier_snps)
  )
)

write.csv(rda_summary_table, 
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_summary_table.csv")), 
          row.names = FALSE)
cat("  - Saved summary table\n")
```

# Summary and Interpretation

## Summary Table

```{r summary-table}
kable(rda_summary_table, caption = "RDA Analysis Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Interpretation Guidelines

### Variance Explained

The **proportion of constrained variance** indicates how much of the total genomic variation is explained by environmental variables.

- **< 5%**: Weak environmental association
- **5-15%**: Moderate environmental association  
- **> 15%**: Strong environmental association

**Your result:** `r round(var_explained * 100, 2)`%

### Significance

```{r interpretation}
if (anova_overall$`Pr(>F)`[1] < 0.001) {
  sig_interpretation <- "highly significant (p < 0.001)"
} else if (anova_overall$`Pr(>F)`[1] < 0.01) {
  sig_interpretation <- "very significant (p < 0.01)"
} else if (anova_overall$`Pr(>F)`[1] < 0.05) {
  sig_interpretation <- "significant (p < 0.05)"
} else {
  sig_interpretation <- "not significant (p ≥ 0.05)"
}
```

The overall RDA model is **`r sig_interpretation`**, indicating that environmental variables `r if(anova_overall$'Pr(>F)'[1] < 0.05) "do" else "do not"` significantly explain genomic variation.

### Candidate SNPs

**`r length(outlier_snps)` candidate SNPs** were identified using the `r OUTLIER_QUANTILE * 100`th percentile threshold.

**Next steps for candidate SNPs:**

1. Annotate SNPs to identify genes and functional regions
2. Validate findings with independent datasets
3. Test for signatures of selection (e.g., Fst, iHS)
4. Investigate biological functions and pathways
5. Consider environmental associations for each candidate

### Reading the RDA Biplot

- **Points (samples)**: Positioned by their genomic variation constrained by environment
- **Arrows (environmental variables)**: 
  - Length indicates strength of relationship
  - Direction shows environmental gradient
  - Samples in arrow direction have higher values for that variable
- **Arrow angles**: Show correlations between variables
  - Acute angles (< 90°): Positive correlation
  - Obtuse angles (> 90°): Negative correlation  
  - Right angles (90°): No correlation

# Session Information

```{r session-info}
sessionInfo()
```

# References

1. Forester, B. R., et al. (2018). Comparing methods for detecting multilocus adaptation with multivariate genotype–environment associations. *Molecular Ecology*, 27(9), 2215-2233.

2. Capblancq, T., & Forester, B. R. (2021). Redundancy analysis: A Swiss Army Knife for landscape genomics. *Methods in Ecology and Evolution*, 12(12), 2298-2309.

3. Legendre, P., & Legendre, L. (2012). *Numerical ecology* (3rd ed.). Elsevier.

4. Rellstab, C., et al. (2015). A practical guide to environmental association analysis in landscape genomics. *Molecular Ecology*, 24(17), 4348-4370.

---

**Analysis completed:** `r Sys.time()`
