---
title: "Calculate BLUEs for GWAS from Multi-Environment Trials"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Overview

This analysis calculates **Best Linear Unbiased Estimates (BLUEs)** for multiple traits from multi-environment trial data, typically for use in Genome-Wide Association Studies (GWAS).

**Key Features:**

- BLUEs calculated across all environments (primary GWAS input)
- Environment-specific BLUEs for each location
- Broad-sense heritability estimates for trait quality assessment
- Variance component decomposition
- Automatic filtering of high-heritability traits

## Methods

### BLUE Calculation Model

**Across all environments:**

$$Y_{ijk} = \mu + G_i + E_j + R_{jk} + \epsilon_{ijk}$$

Where:
- $Y_{ijk}$ = observed phenotype
- $\mu$ = overall mean
- $G_i$ = **fixed** genotype effect (what we want to estimate)
- $E_j$ = random environment effect
- $R_{jk}$ = random replicate effect within environment
- $\epsilon_{ijk}$ = residual error

**For individual environments:**

$$Y_{ik} = \mu + G_i + R_k + \epsilon_{ik}$$

### Heritability Calculation

Broad-sense heritability across environments:

$$H^2 = \frac{\sigma^2_G}{\sigma^2_G + \frac{\sigma^2_E}{n_E} + \frac{\sigma^2_{Rep}}{n_E \times n_{Rep}} + \frac{\sigma^2_e}{n_E \times n_{Rep}}}$$

## Configuration

**Edit these parameters for your data:**

```{r config}
# ============================================
# INPUT DATA CONFIGURATION
# ============================================

# Input CSV file with phenotype data
input_file <- "your_data.csv"

# Column names for experimental design
accession_col <- "accession"      # Genotype/line/accession identifier
environment_col <- "environment"  # Environment/location/trial name
rep_col <- "rep"                  # Replicate/block number

# ============================================
# OUTPUT CONFIGURATION
# ============================================

# Output file prefix
output_prefix <- "BLUEs"

# Heritability threshold for high-quality traits
h2_threshold <- 0.3

# ============================================
# ANALYSIS OPTIONS
# ============================================

# Calculate environment-specific BLUEs?
calc_env_specific <- TRUE

# Calculate heritability?
calc_heritability <- TRUE
```

## Load Required Libraries

```{r libraries}
# Load required packages
required_packages <- c("lme4", "emmeans", "dplyr", "purrr", "ggplot2", "tidyr", "knitr")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Required package '", pkg, "' is not installed.\n",
         "Please install it with: install.packages('", pkg, "')")
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

cat("All libraries loaded successfully.\n")
```

## Load and Validate Data

```{r load_data}
# Check if input file exists
if (!file.exists(input_file)) {
  stop("Error: Input file '", input_file, "' not found!")
}

# Read data
data <- read.csv(input_file, na.strings = c("NA", ".", ""))

cat("Data loaded successfully!\n")
cat("Dimensions:", nrow(data), "rows x", ncol(data), "columns\n\n")

# Verify required columns exist
required_cols <- c(accession_col, environment_col, rep_col)
missing_cols <- setdiff(required_cols, names(data))

if (length(missing_cols) > 0) {
  stop("Error: Missing required columns: ", paste(missing_cols, collapse = ", "))
}

# Get trait names (all columns EXCEPT the design columns)
trait_names <- setdiff(names(data), required_cols)

cat("Found", length(trait_names), "trait columns\n")
cat("Traits:", paste(head(trait_names, 10), collapse = ", "), 
    if(length(trait_names) > 10) "..." else "", "\n\n")

# Convert design factors to factors
data[[accession_col]] <- as.factor(data[[accession_col]])
data[[environment_col]] <- as.factor(data[[environment_col]])
data[[rep_col]] <- as.factor(data[[rep_col]])

# Display experimental design summary
cat("Experimental Design Summary:\n")
cat("  Accessions:  ", nlevels(data[[accession_col]]), "\n")
cat("  Environments:", nlevels(data[[environment_col]]), "\n")
cat("  Replicates:  ", nlevels(data[[rep_col]]), "\n")
cat("  Total plots: ", nrow(data), "\n\n")

# Show first few rows
cat("Data preview:\n")
print(head(data))
```

## Calculate BLUEs Across All Environments

BLUEs calculated with genotype as fixed effect and environment/replicate as random effects.

```{r calculate_blues}
# Function to calculate BLUEs for one trait
calc_blues <- function(trait_name) {
  
  # Remove missing values for this trait
  data_clean <- data %>% filter(!is.na(!!sym(trait_name)))
  
  # Build formula: genotype FIXED, environment and rep RANDOM
  formula_str <- paste0(trait_name, " ~ ", accession_col, 
                       " + (1|", environment_col, ") + (1|", 
                       environment_col, ":", rep_col, ")")
  formula <- as.formula(formula_str)
  
  # Fit mixed model
  model <- lmer(formula, data = data_clean, REML = TRUE)
  
  # Extract BLUEs (Best Linear Unbiased Estimates)
  blues <- emmeans(model, specs = accession_col)
  blues_df <- as.data.frame(blues)
  
  # Create clean output
  blues_clean <- data.frame(
    accession = blues_df[[accession_col]],
    value = blues_df$emmean
  )
  names(blues_clean) <- c(accession_col, trait_name)
  
  return(blues_clean)
}

# Calculate BLUEs for all traits
cat("Calculating BLUEs across all environments for", length(trait_names), "traits...\n\n")
blues_list <- list()

# Progress bar
pb <- txtProgressBar(min = 0, max = length(trait_names), style = 3)

for (i in seq_along(trait_names)) {
  trait <- trait_names[i]
  
  tryCatch({
    blues_list[[i]] <- calc_blues(trait)
  }, error = function(e) {
    cat("\n  ERROR with", trait, ":", e$message, "\n")
    blues_list[[i]] <- NULL
  })
  
  setTxtProgressBar(pb, i)
}

close(pb)

# Remove NULL entries (failed traits)
blues_list <- blues_list[!sapply(blues_list, is.null)]

# Combine all traits into one dataframe
blues_all <- reduce(blues_list, left_join, by = accession_col)

cat("\n\nBLUEs calculation complete!\n")
cat("Successfully calculated BLUEs for:", ncol(blues_all) - 1, "traits\n")
cat("Accessions:", nrow(blues_all), "\n")
```

### BLUEs Summary

```{r blues_summary}
# Summary statistics for BLUEs
cat("\nBLUEs Data Structure:\n")
cat("  Accessions: ", nrow(blues_all), "\n")
cat("  Traits:     ", ncol(blues_all) - 1, "\n\n")

# Show first few BLUEs
cat("First few BLUEs:\n")
print(head(blues_all))
```

## Calculate Heritability

```{r calculate_heritability, eval=calc_heritability}
if (calc_heritability) {
  
  # Function to calculate heritability for one trait
  calc_h2 <- function(trait_name) {
    
    data_clean <- data %>% filter(!is.na(!!sym(trait_name)))
    
    # Build formula with accession as RANDOM for variance components
    formula_str <- paste0(trait_name, " ~ (1|", accession_col, 
                         ") + (1|", environment_col, ") + (1|", 
                         environment_col, ":", rep_col, ")")
    formula <- as.formula(formula_str)
    
    # Fit model
    model <- lmer(formula, data = data_clean, REML = TRUE)
    
    # Extract variance components
    vc <- as.data.frame(VarCorr(model))
    var_g <- vc$vcov[vc$grp == accession_col]
    var_env <- vc$vcov[vc$grp == environment_col]
    var_rep <- vc$vcov[grepl(":", vc$grp) & grepl(environment_col, vc$grp)]
    var_e <- vc$vcov[vc$grp == "Residual"]
    
    # Handle missing components
    if (length(var_g) == 0) var_g <- 0
    if (length(var_env) == 0) var_env <- 0
    if (length(var_rep) == 0) var_rep <- 0
    if (length(var_e) == 0) var_e <- 0
    
    # Calculate broad-sense heritability
    n_env <- length(unique(data_clean[[environment_col]]))
    n_rep <- length(unique(data_clean[[rep_col]]))
    h2 <- var_g / (var_g + var_env/n_env + var_rep/(n_env*n_rep) + var_e/(n_env*n_rep))
    
    return(data.frame(
      trait = trait_name,
      h2 = round(h2, 3),
      var_genotype = round(var_g, 4),
      var_environment = round(var_env, 4),
      var_rep = round(var_rep, 4),
      var_residual = round(var_e, 4)
    ))
  }
  
  # Calculate heritability for all traits
  cat("Calculating heritabilities...\n\n")
  h2_list <- list()
  
  # Progress bar
  pb <- txtProgressBar(min = 0, max = length(trait_names), style = 3)
  
  for (i in seq_along(trait_names)) {
    trait <- trait_names[i]
    
    tryCatch({
      h2_list[[i]] <- calc_h2(trait)
    }, error = function(e) {
      cat("\n  ERROR with", trait, ":", e$message, "\n")
      h2_list[[i]] <- NULL
    })
    
    setTxtProgressBar(pb, i)
  }
  
  close(pb)
  
  # Combine results
  h2_all <- bind_rows(h2_list)
  
  cat("\n\nHeritability calculation complete!\n")
  cat("Traits analyzed:", nrow(h2_all), "\n")
}
```

### Heritability Summary

```{r heritability_summary, eval=calc_heritability}
if (calc_heritability && exists("h2_all") && nrow(h2_all) > 0) {
  
  cat("\nHeritability Statistics:\n")
  cat("==========================================\n")
  cat("Mean H²:   ", round(mean(h2_all$h2, na.rm = TRUE), 3), "\n")
  cat("Median H²: ", round(median(h2_all$h2, na.rm = TRUE), 3), "\n")
  cat("Range:     ", round(min(h2_all$h2, na.rm = TRUE), 3), "-", 
      round(max(h2_all$h2, na.rm = TRUE), 3), "\n")
  cat("Traits with H² >", h2_threshold, ":", 
      sum(!is.na(h2_all$h2) & h2_all$h2 > h2_threshold), "\n")
  cat("==========================================\n\n")
  
  # Top 10 traits by heritability
  cat("Top 10 Traits by Heritability:\n")
  print(h2_all %>% arrange(desc(h2)) %>% head(10))
}
```

### Heritability Distribution

```{r plot_heritability, eval=calc_heritability, fig.width=10, fig.height=6}
if (calc_heritability && exists("h2_all") && nrow(h2_all) > 0) {
  
  ggplot(h2_all, aes(x = h2)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
    geom_vline(xintercept = h2_threshold, color = "red", linetype = "dashed", size = 1) +
    labs(
      title = "Distribution of Broad-Sense Heritability",
      subtitle = paste0("Red line indicates H² threshold of ", h2_threshold),
      x = "Broad-Sense Heritability (H²)",
      y = "Number of Traits"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 11, color = "gray50"),
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 12)
    )
}
```

### Variance Component Decomposition

```{r plot_variance_components, eval=calc_heritability, fig.width=12, fig.height=8}
if (calc_heritability && exists("h2_all") && nrow(h2_all) > 0) {
  
  # Prepare data for stacked bar plot
  var_data <- h2_all %>%
    select(trait, var_genotype, var_environment, var_rep, var_residual) %>%
    pivot_longer(cols = starts_with("var_"), 
                 names_to = "component", 
                 values_to = "variance") %>%
    mutate(
      component = factor(
        component,
        levels = c("var_genotype", "var_environment", "var_rep", "var_residual"),
        labels = c("Genotype", "Environment", "Replicate", "Residual")
      )
    )
  
  # Calculate total variance for percentage
  total_var <- h2_all %>%
    mutate(total = var_genotype + var_environment + var_rep + var_residual) %>%
    select(trait, total)
  
  var_data <- var_data %>%
    left_join(total_var, by = "trait") %>%
    mutate(percentage = variance / total * 100)
  
  # Sort traits by heritability and take top 20
  h2_sorted <- h2_all %>% arrange(desc(h2))
  top_20_traits <- h2_sorted$trait[1:min(20, nrow(h2_sorted))]
  
  var_data_top20 <- var_data %>%
    filter(trait %in% top_20_traits) %>%
    mutate(trait = factor(trait, levels = rev(top_20_traits)))
  
  ggplot(var_data_top20, aes(x = trait, y = percentage, fill = component)) +
    geom_bar(stat = "identity", position = "stack") +
    coord_flip() +
    scale_fill_manual(
      values = c(
        "Genotype" = "#2ecc71",
        "Environment" = "#e74c3c",
        "Replicate" = "#f39c12",
        "Residual" = "#95a5a6"
      )
    ) +
    labs(
      title = "Variance Component Decomposition (Top 20 Traits by H²)",
      x = "Trait",
      y = "Percentage of Total Variance (%)",
      fill = "Component"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      legend.position = "top"
    )
}
```

## Calculate Environment-Specific BLUEs

```{r calculate_env_blues, eval=calc_env_specific}
if (calc_env_specific) {
  
  # Function to calculate BLUEs for one environment
  calc_blues_by_env <- function(trait_name, env) {
    
    # Filter data for this environment and trait
    data_env <- data %>% 
      filter(!!sym(environment_col) == env) %>%
      filter(!is.na(!!sym(trait_name)))
    
    # Build formula - accession fixed, rep random
    formula_str <- paste0(trait_name, " ~ ", accession_col, " + (1|", rep_col, ")")
    formula <- as.formula(formula_str)
    
    # Fit mixed model
    model <- lmer(formula, data = data_env, REML = TRUE)
    
    # Extract BLUEs
    blues <- emmeans(model, specs = accession_col)
    blues_df <- as.data.frame(blues)
    
    # Create clean output
    blues_clean <- data.frame(
      accession = blues_df[[accession_col]],
      value = blues_df$emmean
    )
    names(blues_clean) <- c(accession_col, trait_name)
    
    return(blues_clean)
  }
  
  # Get unique environments
  environments <- levels(data[[environment_col]])
  cat("\nCalculating BLUEs for each environment separately...\n")
  cat("Environments found:", paste(environments, collapse = ", "), "\n\n")
  
  # Calculate BLUEs for each environment
  blues_by_env <- list()
  
  for (env in environments) {
    cat("Processing environment:", env, "\n")
    
    env_blues_list <- list()
    
    for (i in seq_along(trait_names)) {
      trait <- trait_names[i]
      
      tryCatch({
        env_blues_list[[i]] <- calc_blues_by_env(trait, env)
      }, error = function(e) {
        # Silent error handling for individual traits
        env_blues_list[[i]] <- NULL
      })
    }
    
    # Remove NULL entries
    env_blues_list <- env_blues_list[!sapply(env_blues_list, is.null)]
    
    # Combine all traits for this environment
    if (length(env_blues_list) > 0) {
      blues_by_env[[env]] <- reduce(env_blues_list, left_join, by = accession_col)
      cat("  Completed:", env, "-", ncol(blues_by_env[[env]]) - 1, "traits\n")
    }
  }
  
  cat("\nEnvironment-specific BLUEs calculation complete!\n")
}
```

### Environment-Specific Summary

```{r env_blues_summary, eval=calc_env_specific}
if (calc_env_specific && exists("blues_by_env") && length(blues_by_env) > 0) {
  
  cat("\nEnvironment-Specific BLUEs Summary:\n")
  cat("==========================================\n")
  
  for (env in names(blues_by_env)) {
    cat(env, ":\n")
    cat("  Accessions: ", nrow(blues_by_env[[env]]), "\n")
    cat("  Traits:     ", ncol(blues_by_env[[env]]) - 1, "\n\n")
  }
}
```

## High Heritability Traits

Traits with H² > `r h2_threshold` are considered high-quality for GWAS.

```{r high_h2_traits, eval=calc_heritability}
if (calc_heritability && exists("h2_all") && nrow(h2_all) > 0) {
  
  high_h2_traits <- h2_all %>% 
    filter(h2 > h2_threshold) %>% 
    arrange(desc(h2))
  
  cat("\nTraits with H² >", h2_threshold, ":\n")
  cat("==========================================\n")
  
  if (nrow(high_h2_traits) > 0) {
    print(high_h2_traits)
    
    # Extract BLUEs for high heritability traits only
    blues_high_h2 <- blues_all[, c(accession_col, high_h2_traits$trait)]
    
    cat("\n", nrow(high_h2_traits), "high-heritability traits identified\n")
  } else {
    cat("No traits with H² >", h2_threshold, "\n")
  }
}
```

## Correlation Analysis

Correlation between BLUEs across traits.

```{r correlation_analysis, fig.width=12, fig.height=10}
# Calculate correlation matrix for BLUEs (exclude accession column)
blues_numeric <- blues_all %>% select(-all_of(accession_col))

if (ncol(blues_numeric) > 1 && ncol(blues_numeric) <= 50) {
  
  cor_matrix <- cor(blues_numeric, use = "pairwise.complete.obs")
  
  # Convert to long format for ggplot
  cor_long <- as.data.frame(as.table(cor_matrix))
  names(cor_long) <- c("Trait1", "Trait2", "Correlation")
  
  ggplot(cor_long, aes(x = Trait1, y = Trait2, fill = Correlation)) +
    geom_tile() +
    scale_fill_gradient2(
      low = "blue", 
      mid = "white", 
      high = "red",
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    labs(
      title = "Correlation Heatmap of BLUEs",
      x = "",
      y = ""
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
  
} else if (ncol(blues_numeric) > 50) {
  cat("Too many traits (", ncol(blues_numeric), ") for correlation heatmap visualization.\n")
  cat("Correlation matrix calculated but not displayed.\n")
} else {
  cat("Not enough traits for correlation analysis.\n")
}
```

## Save Results

```{r save_results}
# Create output directory if needed
output_dir <- dirname(output_prefix)
if (output_dir != "." && !dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Export main BLUEs file (for GWAS)
main_blues_file <- paste0(output_prefix, "_for_GWAS.csv")
write.csv(blues_all, main_blues_file, row.names = FALSE)
cat("Saved:", main_blues_file, "\n")

# Export environment-specific BLUEs
if (calc_env_specific && exists("blues_by_env") && length(blues_by_env) > 0) {
  for (env in names(blues_by_env)) {
    env_file <- paste0(output_prefix, "_", env, ".csv")
    write.csv(blues_by_env[[env]], env_file, row.names = FALSE)
    cat("Saved:", env_file, "\n")
  }
}

# Export heritabilities
if (calc_heritability && exists("h2_all") && nrow(h2_all) > 0) {
  h2_file <- paste0(output_prefix, "_heritabilities.csv")
  write.csv(h2_all, h2_file, row.names = FALSE)
  cat("Saved:", h2_file, "\n")
}

# Export high heritability BLUEs
if (calc_heritability && exists("high_h2_traits") && nrow(high_h2_traits) > 0) {
  high_h2_file <- paste0(output_prefix, "_high_heritability.csv")
  write.csv(blues_high_h2, high_h2_file, row.names = FALSE)
  cat("Saved:", high_h2_file, "\n")
}
```

## Final Summary

```{r final_summary}
cat("\n================================================================================\n")
cat("BLUE CALCULATION COMPLETE!\n")
cat("================================================================================\n")
cat("Input file:              ", input_file, "\n")
cat("Accessions analyzed:     ", nrow(blues_all), "\n")
cat("Total traits processed:  ", length(trait_names), "\n")
cat("Traits with BLUEs:       ", ncol(blues_all) - 1, "\n")

if (calc_heritability && exists("h2_all") && nrow(h2_all) > 0) {
  cat("\nHeritability Statistics:\n")
  cat("  Mean H²:               ", round(mean(h2_all$h2, na.rm = TRUE), 3), "\n")
  cat("  Traits with H² >", h2_threshold, ":", 
      sum(!is.na(h2_all$h2) & h2_all$h2 > h2_threshold), "\n")
}

if (calc_env_specific && exists("blues_by_env") && length(blues_by_env) > 0) {
  cat("\nEnvironments analyzed:   ", length(blues_by_env), "\n")
  for (env in names(blues_by_env)) {
    cat("  ", env, ": ", ncol(blues_by_env[[env]]) - 1, " traits\n", sep = "")
  }
}

cat("\nOutput files created:\n")
cat("  - ", main_blues_file, "\n", sep = "")

if (calc_env_specific && exists("blues_by_env") && length(blues_by_env) > 0) {
  for (env in names(blues_by_env)) {
    cat("  - ", output_prefix, "_", env, ".csv\n", sep = "")
  }
}

if (calc_heritability && exists("h2_all") && nrow(h2_all) > 0) {
  cat("  - ", output_prefix, "_heritabilities.csv\n", sep = "")
}

if (calc_heritability && exists("high_h2_traits") && nrow(high_h2_traits) > 0) {
  cat("  - ", output_prefix, "_high_heritability.csv\n", sep = "")
}

cat("================================================================================\n")
```

## Session Information

```{r session_info}
sessionInfo()
```

---

## Notes for GWAS

**Using BLUEs in GWAS:**

1. **Primary file**: `BLUEs_for_GWAS.csv` contains BLUEs across all environments
   - This is the standard input for GWAS software
   - Rows = accessions, Columns = traits

2. **Quality filtering**: Consider using only high-heritability traits (H² > 0.3)
   - These traits have better genetic signal
   - Reduces multiple testing burden
   - File: `BLUEs_high_heritability.csv`

3. **Environment-specific GWAS**: Individual environment files allow detection of:
   - Environment-specific QTL
   - Genotype × Environment interactions
   - Local adaptation signals

**Recommended GWAS workflow:**

1. Start with high-heritability traits from `BLUEs_high_heritability.csv`
2. Run genome-wide scan with appropriate population structure correction
3. For significant associations, validate across individual environments
4. Check if variance components suggest strong G×E interactions

**File format compatibility:**

- Most GWAS software (GAPIT, TASSEL, GEMMA, etc.) accepts CSV format
- Ensure accession IDs match exactly with genotype file
- Missing data (NA) is acceptable in trait columns
