---
title: "Environmental Genome-Wide Association Study (eGWAS)"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)
```

## Overview

This analysis performs **environmental Genome-Wide Association Study (eGWAS)** to identify genetic variants associated with environmental variables. The approach uses linear mixed models (LMM) with various population structure corrections.

**Key Features:**

- VCF file processing and quality control
- Kinship matrix calculation for relatedness
- Multiple population structure correction strategies
- Genomic inflation factor (λ) calculation
- FDR-corrected significance testing
- Manhattan and QQ plots for visualization
- Variance partitioning analysis

## Methods

### Association Model

Linear mixed model for environmental association:

$$y = X\beta + Zu + \epsilon$$

Where:
- $y$ = environmental variable values
- $X$ = fixed effects matrix (population structure: PCs)
- $\beta$ = fixed effect coefficients
- $Z$ = genotype design matrix
- $u \sim N(0, K\sigma^2_g)$ = random polygenic effects
- $K$ = kinship matrix
- $\epsilon$ = residual error

**Population structure correction strategies:**

1. **Model 1**: Kinship only (K)
2. **Model 2**: Kinship + 3 PCs (K + PC1-3)
3. **Model 3**: Kinship + 10 PCs (K + PC1-10)

The best model is selected based on genomic inflation factor (λ ≈ 1).

## Configuration

**Edit these parameters for your analysis:**

```{r config}
# ============================================
# INPUT FILES
# ============================================

# VCF file with genotype data
vcf_file <- "path/to/your/genotypes.vcf"

# CSV file with environmental data
# Required columns: sample_id (matching VCF), environmental variables
# Optional columns: latitude, longitude (excluded from analysis)
env_file <- "path/to/your/environment.csv"

# ============================================
# OUTPUT DIRECTORIES
# ============================================

output_dir <- "egwas_output"
figures_dir <- "egwas_figures"

# ============================================
# ANALYSIS PARAMETERS
# ============================================

# Quality control thresholds
min_maf <- 0.05              # Minimum minor allele frequency
missing_thresh <- 0.10       # Maximum proportion missing data per marker

# Significance thresholds
fdr_threshold <- 0.20        # FDR threshold for significance

# Population structure
n_pcs_model2 <- 3            # Number of PCs for model 2
n_pcs_model3 <- 10           # Number of PCs for model 3
```

## Load Required Libraries

```{r libraries}
# CRITICAL NOTE: Package conflicts
# Several packages mask dplyr functions (especially select, filter, rename)
# Functions are called with dplyr:: prefix throughout this script
# Do NOT remove the dplyr:: prefixes or you will get errors

# Load required packages
required_packages <- c(
  "vcfR",      # VCF file reading
  "sommer",    # Mixed models and FDR
  "tidyverse", # Data manipulation
  "rrBLUP",    # GWAS and kinship
  "qvalue",    # FDR correction
  "QCEWAS",    # Genomic inflation
  "qqman"      # Manhattan/QQ plots
)

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Required package '", pkg, "' is not installed.\n",
         "Please install it with: install.packages('", pkg, "')")
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# Create output directories
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

cat("All libraries loaded successfully.\n")
cat("Output directories created.\n")
```

## Read and Process VCF File

```{r read_vcf}
cat("Reading VCF file...\n")

# Check if file exists
if (!file.exists(vcf_file)) {
  stop("VCF file not found: ", vcf_file)
}

vcf <- read.vcfR(vcf_file, verbose = FALSE)

cat("VCF file loaded:\n")
cat("  Variants:", nrow(vcf@fix), "\n")
cat("  Samples: ", ncol(vcf@gt) - 1, "\n\n")
```

### Convert Genotypes to Numeric Format

```{r convert_genotypes}
cat("Converting genotypes to numeric format...\n")

# Extract genotype matrix
geno_mat <- extract.gt(vcf, element = "GT")

# Convert VCF genotypes to numeric (-1, 0, 1)
# 0/0 or 0|0 → -1 (homozygous reference)
# 0/1 or 0|1 or 1|0 → 0 (heterozygous)
# 1/1 or 1|1 → 1 (homozygous alternate)

geno_numeric <- matrix(NA, nrow = nrow(geno_mat), ncol = ncol(geno_mat))
rownames(geno_numeric) <- rownames(geno_mat)
colnames(geno_numeric) <- colnames(geno_mat)

pb <- txtProgressBar(min = 0, max = nrow(geno_mat), style = 3)

for (i in 1:nrow(geno_mat)) {
  for (j in 1:ncol(geno_mat)) {
    gt <- geno_mat[i, j]
    if (is.na(gt)) {
      geno_numeric[i, j] <- NA
    } else if (gt == "0/0" || gt == "0|0") {
      geno_numeric[i, j] <- -1
    } else if (gt == "0/1" || gt == "0|1" || gt == "1|0") {
      geno_numeric[i, j] <- 0
    } else if (gt == "1/1" || gt == "1|1") {
      geno_numeric[i, j] <- 1
    } else {
      # Handle multiallelic or other cases
      alleles <- as.numeric(unlist(strsplit(gt, "[/|]")))
      geno_numeric[i, j] <- sum(alleles) - 1
    }
  }
  
  if (i %% 1000 == 0) setTxtProgressBar(pb, i)
}

close(pb)

# Transpose: samples as rows, markers as columns
geno_mat_t <- t(geno_numeric)

cat("\nGenotype matrix transposed:", nrow(geno_mat_t), "samples x", 
    ncol(geno_mat_t), "markers\n")
```

### Remove Monomorphic Markers

```{r remove_monomorphic}
cat("\nChecking for monomorphic markers...\n")

is_polymorphic <- apply(geno_mat_t, 2, function(x) {
  x_clean <- x[!is.na(x)]
  length(unique(x_clean)) > 1
})

n_monomorphic <- sum(!is_polymorphic)

if (n_monomorphic > 0) {
  cat("Removing", n_monomorphic, "monomorphic markers\n")
  geno_mat_t <- geno_mat_t[, is_polymorphic]
  geno_numeric <- geno_numeric[is_polymorphic, ]
}

# Create SNP information data frame
snp_info <- data.frame(
  marker = rownames(geno_numeric),
  chrom = getCHROM(vcf)[is_polymorphic],
  pos = getPOS(vcf)[is_polymorphic],
  stringsAsFactors = FALSE
)

cat("After removing monomorphic markers:\n")
cat("  Markers:", nrow(snp_info), "\n")
cat("  Samples:", ncol(geno_mat_t), "\n")
```

## Read Environmental Data

```{r read_env}
cat("\nReading environmental data...\n")

# Check if file exists
if (!file.exists(env_file)) {
  stop("Environmental data file not found: ", env_file)
}

env_data <- read.csv(env_file, stringsAsFactors = FALSE)

cat("Environmental data loaded:\n")
cat("  Samples:", nrow(env_data), "\n")
cat("  Variables:", ncol(env_data), "\n\n")

# Check for sample_id column
if (!"sample_id" %in% colnames(env_data)) {
  colnames(env_data)[1] <- "sample_id"
  cat("Note: Assuming first column contains sample IDs\n")
}

# Identify environmental variables
env_vars <- setdiff(colnames(env_data), c("sample_id", "latitude", "longitude"))

cat("Found", length(env_vars), "environmental variables:\n")
cat(paste(head(env_vars, 10), collapse = ", "), 
    if(length(env_vars) > 10) "..." else "", "\n\n")

# Display preview
cat("Environmental data preview:\n")
print(head(env_data))
```

## Match Samples

```{r match_samples}
# Find common samples between genotype and environmental data
common_samples <- intersect(rownames(geno_mat_t), env_data$sample_id)

cat("\nSample matching:\n")
cat("  Genotype samples:     ", nrow(geno_mat_t), "\n")
cat("  Environment samples:  ", nrow(env_data), "\n")
cat("  Common samples:       ", length(common_samples), "\n\n")

if (length(common_samples) == 0) {
  stop("No common samples found! Check that sample IDs match between VCF and CSV files.")
}

# Subset to common samples
geno_mat_t <- geno_mat_t[common_samples, ]
env_data <- env_data[match(common_samples, env_data$sample_id), ]

cat("Datasets matched to", length(common_samples), "common samples\n")
```

## Quality Control and Filtering

```{r qc_filtering}
cat("\nPerforming quality control...\n")

# Calculate missing data per marker
missing_per_marker <- colMeans(is.na(geno_mat_t))
markers_to_keep <- missing_per_marker <= missing_thresh

cat("Removed", sum(!markers_to_keep), "markers with >", 
    missing_thresh * 100, "% missing data\n")

geno_mat_filtered <- geno_mat_t[, markers_to_keep]
snp_info_filtered <- snp_info[markers_to_keep, ]

# Calculate minor allele frequency
maf <- apply(geno_mat_filtered, 2, function(x) {
  x <- x[!is.na(x)]
  p <- (sum(x == 1) + 2 * sum(x == 0)) / (2 * length(x))
  return(min(p, 1 - p))
})

# Filter by MAF
markers_pass_maf <- maf >= min_maf
cat("Removed", sum(!markers_pass_maf), "markers with MAF <", min_maf, "\n")

geno_mat_final <- geno_mat_filtered[, markers_pass_maf]
snp_info_final <- snp_info_filtered[markers_pass_maf, ]

cat("\nFinal filtered dataset:\n")
cat("  Samples: ", nrow(geno_mat_final), "\n")
cat("  Markers: ", ncol(geno_mat_final), "\n\n")

# Impute missing genotypes with column means
cat("Imputing missing genotypes...\n")
for (j in 1:ncol(geno_mat_final)) {
  missing_idx <- is.na(geno_mat_final[, j])
  if (any(missing_idx)) {
    geno_mat_final[missing_idx, j] <- mean(geno_mat_final[, j], na.rm = TRUE)
  }
}
cat("Imputation complete\n")
```

### QC Summary Plot

```{r plot_qc, fig.width=12, fig.height=5}
# Create QC summary plots
par(mfrow = c(1, 2))

# MAF distribution
hist(maf[markers_pass_maf], breaks = 50, 
     main = "Minor Allele Frequency Distribution",
     xlab = "MAF", col = "steelblue", border = "white")
abline(v = min_maf, col = "red", lwd = 2, lty = 2)

# Missing data distribution
hist(missing_per_marker[markers_to_keep], breaks = 50,
     main = "Missing Data per Marker",
     xlab = "Proportion Missing", col = "coral", border = "white")
abline(v = missing_thresh, col = "red", lwd = 2, lty = 2)
```

## Calculate Kinship Matrix

```{r kinship}
cat("\nCalculating kinship matrix...\n")

K <- A.mat(geno_mat_final, min.MAF = 0, max.missing = 1)

cat("Kinship matrix calculated:", nrow(K), "x", ncol(K), "\n")

# Summary statistics
cat("Kinship matrix summary:\n")
cat("  Mean relatedness:  ", round(mean(K[lower.tri(K)]), 4), "\n")
cat("  Median relatedness:", round(median(K[lower.tri(K)]), 4), "\n")
cat("  Range:             ", round(min(K), 4), "to", round(max(K), 4), "\n")
```

### Visualize Kinship Matrix

```{r plot_kinship, fig.width=10, fig.height=9}
# Heatmap of kinship matrix
heatmap(K, 
        col = colorRampPalette(c("blue", "white", "red"))(100),
        main = "Kinship Matrix Heatmap",
        margins = c(8, 8),
        cexRow = 0.5,
        cexCol = 0.5)
```

## Calculate Principal Components

```{r pca}
cat("\nCalculating principal components from kinship matrix...\n")

K_pca <- prcomp(K)

# Extract PCs
K_eigens <- K_pca$x[, 1:max(n_pcs_model2, n_pcs_model3)] %>%
  as.data.frame() %>%
  rownames_to_column("sample_id")

# Variance explained
var_exp <- summary(K_pca)$importance[2, ]

cat("Variance explained by top PCs:\n")
for (i in 1:min(10, length(var_exp))) {
  cat("  PC", i, ": ", round(var_exp[i] * 100, 2), "%\n", sep = "")
}
```

### PCA Plots

```{r plot_pca, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))

# PC1 vs PC2
plot(K_pca$x[, 1], K_pca$x[, 2],
     xlab = paste0("PC1 (", round(var_exp[1] * 100, 1), "%)"),
     ylab = paste0("PC2 (", round(var_exp[2] * 100, 1), "%)"),
     main = "PCA of Kinship Matrix",
     pch = 16, col = alpha("steelblue", 0.6))

# Scree plot
plot(1:min(20, length(var_exp)), var_exp[1:min(20, length(var_exp))] * 100,
     type = "b", pch = 16,
     xlab = "Principal Component",
     ylab = "Variance Explained (%)",
     main = "Scree Plot",
     col = "darkblue", lwd = 2)
grid()
```

## Prepare Data for GWAS

```{r prepare_gwas}
cat("\nPreparing data for GWAS...\n")

# Format genotype data for rrBLUP
geno_df <- t(geno_mat_final) %>%
  as.data.frame() %>%
  rownames_to_column("marker") %>%
  dplyr::inner_join(snp_info_final, ., by = "marker") %>%
  as.data.frame()

# Ensure correct column order: marker, chrom, pos, then individuals
col_order <- c("marker", "chrom", "pos", rownames(geno_mat_final))
geno_df <- geno_df[, col_order]

cat("Genotype data formatted:\n")
cat("  Markers:    ", nrow(geno_df), "\n")
cat("  Individuals:", ncol(geno_df) - 3, "\n\n")

# Prepare phenotype (environmental) data
pheno <- env_data %>%
  dplyr::select(sample_id, all_of(env_vars)) %>%
  dplyr::rename(individual = sample_id) %>%
  as.data.frame()

cat("Phenotype data prepared:\n")
cat("  Individuals:", nrow(pheno), "\n")
cat("  Variables:  ", ncol(pheno) - 1, "\n")
```

## Run eGWAS with Multiple Models

```{r run_egwas}
cat("\nRunning environmental GWAS...\n")
cat("Testing", length(env_vars), "environmental variables\n")
cat("Using 3 population structure correction models\n\n")

# Initialize storage for results
egwas_models <- list()

# Define models
models <- list(
  model1 = list(name = "K only", n_pcs = 0),
  model2 = list(name = paste0("K + ", n_pcs_model2, " PCs"), n_pcs = n_pcs_model2),
  model3 = list(name = paste0("K + ", n_pcs_model3, " PCs"), n_pcs = n_pcs_model3)
)

# Run GWAS for each model
for (model_name in names(models)) {
  model_info <- models[[model_name]]
  
  cat("Running", model_info$name, "...\n")
  
  # Prepare fixed effects
  if (model_info$n_pcs > 0) {
    pc_cols <- paste0("PC", 1:model_info$n_pcs)
    pheno_with_pcs <- pheno %>%
      dplyr::left_join(K_eigens[, c("sample_id", pc_cols)], 
                       by = c("individual" = "sample_id"))
    fixed <- as.matrix(pheno_with_pcs[, pc_cols])
  } else {
    pheno_with_pcs <- pheno
    fixed <- NULL
  }
  
  # Run GWAS for each environmental variable
  model_results <- list()
  
  pb <- txtProgressBar(min = 0, max = length(env_vars), style = 3)
  
  for (i in seq_along(env_vars)) {
    var <- env_vars[i]
    
    tryCatch({
      # Prepare phenotype for this variable
      y <- pheno_with_pcs[[var]]
      
      # Run GWAS
      gwas_result <- GWAS(
        pheno = data.frame(individual = pheno_with_pcs$individual, trait = y),
        geno = geno_df,
        K = K,
        fixed = fixed,
        n.PC = 0,  # PCs already in fixed
        min.MAF = 0,
        n.core = 1,
        P3D = TRUE,
        plot = FALSE
      )
      
      # Store results
      model_results[[var]] <- gwas_result %>%
        mutate(variable = var, model = model_name)
      
    }, error = function(e) {
      cat("\n  Error with", var, ":", e$message, "\n")
      model_results[[var]] <- NULL
    })
    
    setTxtProgressBar(pb, i)
  }
  
  close(pb)
  
  # Combine results for this model
  egwas_models[[model_name]] <- bind_rows(model_results)
  
  cat("\n", model_info$name, "complete:", nrow(egwas_models[[model_name]]), "tests\n\n")
}

# Combine all models
gwas_combined <- bind_rows(egwas_models)

cat("Total GWAS tests completed:", nrow(gwas_combined), "\n")
```

## Calculate Genomic Inflation Factors

```{r lambda}
cat("\nCalculating genomic inflation factors (λ)...\n")

# Calculate lambda for each variable and model
p_lambda <- gwas_combined %>%
  group_by(variable, model) %>%
  summarise(
    n_markers = n(),
    lambda = median(qchisq(1 - trait, 1), na.rm = TRUE) / qchisq(0.5, 1),
    .groups = "drop"
  ) %>%
  arrange(variable, lambda)

# Identify best model for each variable (λ closest to 1)
p_lambda <- p_lambda %>%
  group_by(variable) %>%
  mutate(
    best_model = model[which.min(abs(lambda - 1))],
    is_best = model == best_model
  ) %>%
  ungroup()

cat("\nGenomic inflation factors:\n")
print(p_lambda)
```

### Lambda Summary Plot

```{r plot_lambda, fig.width=10, fig.height=6}
ggplot(p_lambda, aes(x = variable, y = lambda, fill = model)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Genomic Inflation Factor by Model",
    subtitle = "Values close to 1 indicate good population structure correction",
    x = "Environmental Variable",
    y = "Genomic Inflation Factor (λ)",
    fill = "Model"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11, color = "gray50")
  ) +
  scale_fill_brewer(palette = "Set2")
```

## Process GWAS Results

```{r process_results}
cat("\nProcessing GWAS results...\n")

# Add -log10(p-value) and uniform score
gwas_analyzed <- gwas_combined %>%
  mutate(
    p_value = trait,  # GWAS function returns p-values in 'trait' column
    log10p = -log10(p_value),
    unif_score = rank(p_value) / length(p_value)
  )

cat("GWAS results processed:", nrow(gwas_analyzed), "tests\n")
```

## QQ Plots

```{r qq_plots, fig.width=12, fig.height=ceiling(length(env_vars)/3)*4}
cat("\nGenerating QQ plots...\n")

# Calculate grid dimensions
n_vars <- length(env_vars)
n_cols <- 3
n_rows <- ceiling(n_vars / n_cols)

par(mfrow = c(n_rows, n_cols), mar = c(4, 4, 3, 1))

for (var in env_vars) {
  for (model_name in names(models)) {
    
    # Get data for this variable and model
    var_data <- gwas_analyzed %>%
      dplyr::filter(variable == var, model == model_name, !is.na(p_value))
    
    if (nrow(var_data) > 0) {
      # Create QQ plot
      qq(var_data$p_value, 
         main = paste0(var, " (", model_name, ")"))
    }
  }
}

cat("QQ plots generated\n")
```

## Identify Significant Markers

```{r significant_markers}
cat("\nIdentifying significant markers...\n")

# Use model2 (K + 3 PCs) as default best model
# Modify based on your p_lambda results
sig_markers <- gwas_analyzed %>%
  dplyr::filter(model == "model2") %>%
  group_by(variable) %>%
  group_split() %>%
  map_dfr(function(df) {
    var_name <- df$variable[1]
    p_vals <- df$p_value[!is.na(df$p_value)]
    
    if (length(p_vals) == 0) return(tibble())
    
    # Calculate FDR
    fdr_result <- sommer:::fdr(p_vals, fdr.level = fdr_threshold)
    threshold <- fdr_result$fdr.10
    
    sig <- df %>%
      dplyr::filter(p_value <= threshold) %>%
      arrange(p_value)
    
    if (nrow(sig) > 0) {
      cat("  ", var_name, ":", nrow(sig), "significant markers\n")
    }
    
    return(sig)
  })

cat("\nTotal significant markers:", nrow(sig_markers), "\n")

if (nrow(sig_markers) > 0) {
  cat("\nTop significant markers:\n")
  print(head(sig_markers %>% 
               dplyr::select(variable, marker, chrom, pos, p_value) %>%
               arrange(p_value), 20))
}
```

## Manhattan Plots

```{r manhattan_plots, fig.width=14, fig.height=ceiling(length(env_vars)/2)*4}
if (nrow(sig_markers) > 0) {
  cat("\nGenerating Manhattan plots...\n")
  
  par(mfrow = c(ceiling(length(env_vars)/2), 2), mar = c(4, 4, 3, 1))
  
  for (var in env_vars) {
    
    # Get best model for this variable
    best_model_var <- p_lambda %>%
      dplyr::filter(variable == var, is_best) %>%
      pull(model)
    
    if (length(best_model_var) == 0) {
      best_model_var <- "model2"
    }
    
    # Get data
    var_data <- gwas_analyzed %>%
      dplyr::filter(variable == var, model == best_model_var) %>%
      dplyr::filter(!is.na(p_value), p_value > 0)
    
    if (nrow(var_data) > 0) {
      # Prepare for manhattan plot
      manhattan_data <- var_data %>%
        dplyr::select(marker, chrom, pos, p_value) %>%
        dplyr::rename(SNP = marker, CHR = chrom, BP = pos, P = p_value)
      
      # Convert chromosome to numeric if needed
      if (!is.numeric(manhattan_data$CHR)) {
        manhattan_data$CHR <- as.numeric(factor(manhattan_data$CHR))
      }
      
      # Create plot
      tryCatch({
        manhattan(manhattan_data,
                 chr = "CHR", bp = "BP", p = "P", snp = "SNP",
                 main = paste0(var, " (", best_model_var, ")"),
                 col = c("blue4", "orange3"),
                 suggestiveline = -log10(0.05),
                 genomewideline = -log10(fdr_threshold / nrow(manhattan_data)),
                 cex = 0.5)
      }, error = function(e) {
        plot.new()
        text(0.5, 0.5, paste("Error:", var), cex = 1)
      })
    } else {
      plot.new()
      text(0.5, 0.5, paste("No data:", var), cex = 1.5)
    }
  }
  
  cat("Manhattan plots generated\n")
}
```

## Variance Partitioning

Calculate variance explained by significant markers.

```{r variance_explained}
cat("\nCalculating variance explained by significant markers...\n")

variance_explained <- list()

if (nrow(sig_markers) > 0) {
  
  # Merge phenotype with PCs
  pheno_with_pcs <- pheno %>%
    dplyr::left_join(K_eigens[, c("sample_id", paste0("PC", 1:n_pcs_model2))], 
                     by = c("individual" = "sample_id"))
  
  for (var_name in unique(sig_markers$variable)) {
    
    cat("  Processing", var_name, "...\n")
    
    # Get significant markers
    sig_mar_var <- sig_markers %>%
      dplyr::filter(variable == var_name) %>%
      pull(marker)
    
    if (length(sig_mar_var) == 0) next
    
    # Prepare model components
    y <- pheno_with_pcs[[var_name]]
    X <- model.matrix(~ PC1 + PC2 + PC3, data = pheno_with_pcs)
    Z <- model.matrix(~ -1 + individual, data = pheno_with_pcs)
    
    # Fit mixed model
    mm_fit <- mixed.solve(y = y, Z = Z, K = K, X = X, method = "REML")
    
    # Calculate adjusted values
    y_adj <- y - ((X %*% mm_fit$beta) + (Z %*% mm_fit$u))
    
    # Get genotypes for significant markers
    geno_sig <- geno_mat_final[, sig_mar_var, drop = FALSE]
    
    # Fit marker model
    model_data <- data.frame(y_adj = as.numeric(y_adj), geno_sig)
    
    if (ncol(geno_sig) > 1) {
      marker_model <- lm(y_adj ~ ., data = model_data)
      marker_model_step <- step(marker_model, trace = 0)
    } else {
      marker_model_step <- lm(y_adj ~ ., data = model_data)
    }
    
    # Calculate variance components
    pop_structure_r2 <- var(X %*% mm_fit$beta) / var(y)
    polygenic_r2 <- mm_fit$Vu / var(y)
    residual_r2 <- 1 - (pop_structure_r2 + polygenic_r2)
    marker_r2 <- summary(marker_model_step)$adj.r.squared
    
    variance_explained[[var_name]] <- tibble(
      variable = var_name,
      n_sig_markers = length(sig_mar_var),
      n_retained_markers = length(coef(marker_model_step)) - 1,
      pop_structure_var_exp = pop_structure_r2,
      polygenic_var_exp = polygenic_r2,
      markers_var_exp = marker_r2 * residual_r2,
      total_var_exp = pop_structure_r2 + polygenic_r2 + (marker_r2 * residual_r2)
    )
  }
  
  variance_df <- bind_rows(variance_explained)
  
  cat("\nVariance explained summary:\n")
  print(variance_df)
  
} else {
  cat("  No significant markers found\n")
  variance_df <- tibble()
}
```

### Variance Partitioning Plot

```{r plot_variance, fig.width=10, fig.height=6}
if (nrow(variance_df) > 0) {
  
  # Reshape for plotting
  var_plot_data <- variance_df %>%
    dplyr::select(variable, pop_structure_var_exp, polygenic_var_exp, markers_var_exp) %>%
    pivot_longer(cols = -variable, names_to = "component", values_to = "variance") %>%
    mutate(
      component = factor(
        component,
        levels = c("pop_structure_var_exp", "polygenic_var_exp", "markers_var_exp"),
        labels = c("Population Structure", "Polygenic Background", "Significant Markers")
      )
    )
  
  ggplot(var_plot_data, aes(x = variable, y = variance * 100, fill = component)) +
    geom_col(position = "stack") +
    labs(
      title = "Variance Partitioning",
      x = "Environmental Variable",
      y = "Variance Explained (%)",
      fill = "Component"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      legend.position = "top"
    ) +
    scale_fill_manual(values = c(
      "Population Structure" = "#3498db",
      "Polygenic Background" = "#95a5a6",
      "Significant Markers" = "#e74c3c"
    ))
}
```

## Save Results

```{r save_results}
cat("\nSaving results...\n")

# Save R workspace
save(
  gwas_combined,
  gwas_analyzed,
  p_lambda,
  sig_markers,
  variance_df,
  egwas_models,
  K,
  snp_info_final,
  geno_mat_final,
  env_data,
  file = file.path(output_dir, "egwas_results.RData")
)
cat("  Workspace saved to:", file.path(output_dir, "egwas_results.RData"), "\n")

# Save significant markers
if (nrow(sig_markers) > 0) {
  sig_markers_clean <- sig_markers %>%
    dplyr::select(any_of(c("marker", "chrom", "pos", "model", "variable", 
                           "p_value", "log10p", "unif_score"))) %>%
    as.data.frame()
  
  write.csv(sig_markers_clean,
            file = file.path(output_dir, "significant_markers.csv"),
            row.names = FALSE)
  cat("  Significant markers saved to:", 
      file.path(output_dir, "significant_markers.csv"), "\n")
}

# Save variance explained
if (nrow(variance_df) > 0) {
  write.csv(variance_df,
            file = file.path(output_dir, "variance_explained.csv"),
            row.names = FALSE)
  cat("  Variance components saved to:", 
      file.path(output_dir, "variance_explained.csv"), "\n")
}

# Save lambda values
p_lambda_clean <- p_lambda %>%
  dplyr::select(any_of(c("variable", "model", "lambda", "best_model", "is_best"))) %>%
  as.data.frame()

write.csv(p_lambda_clean,
          file = file.path(output_dir, "genomic_inflation_factors.csv"),
          row.names = FALSE)
cat("  Lambda values saved to:", 
    file.path(output_dir, "genomic_inflation_factors.csv"), "\n")
```

## Final Summary

```{r final_summary}
cat("\n================================================================================\n")
cat("ENVIRONMENTAL GWAS COMPLETE!\n")
cat("================================================================================\n")
cat("Input files:\n")
cat("  VCF file:       ", vcf_file, "\n")
cat("  Environment:    ", env_file, "\n\n")

cat("Final dataset:\n")
cat("  Samples:        ", nrow(geno_mat_final), "\n")
cat("  Markers:        ", ncol(geno_mat_final), "\n")
cat("  Environmental variables:", length(env_vars), "\n\n")

cat("Analysis results:\n")
cat("  Total GWAS tests:", nrow(gwas_combined), "\n")

if (nrow(sig_markers) > 0) {
  cat("  Significant markers (FDR <", fdr_threshold, "):", nrow(sig_markers), "\n")
  
  sig_summary <- sig_markers %>%
    group_by(variable) %>%
    summarise(n_sig = n(), .groups = "drop")
  
  cat("\n  Significant markers by variable:\n")
  for (i in 1:nrow(sig_summary)) {
    cat("    ", sig_summary$variable[i], ":", sig_summary$n_sig[i], "\n")
  }
} else {
  cat("  No significant markers found at FDR <", fdr_threshold, "\n")
}

cat("\nOutput directories:\n")
cat("  Results:", output_dir, "\n")
cat("  Figures:", figures_dir, "\n")
cat("================================================================================\n")
```

## Session Information

```{r session_info}
sessionInfo()
```

---

## Interpretation Guide

**Genomic Inflation Factor (λ):**

- **λ ≈ 1**: Good population structure correction
- **λ > 1**: Inflation (too many false positives) - need more stringent correction
- **λ < 1**: Deflation (too conservative) - may be losing power

**Model Selection:**

- Choose the model with λ closest to 1 for each environmental variable
- Model 2 (K + 3 PCs) is typically a good default

**Variance Partitioning:**

- Shows how much environmental variation is explained by:
  - Population structure (PCs)
  - Polygenic background (kinship)
  - Identified significant markers

**Significant Markers:**

- FDR-corrected significance threshold
- Markers associated with environmental adaptation
- Can indicate local adaptation or selection

**Next Steps:**

1. Validate significant markers in independent populations
2. Check for overlap with functional genes
3. Investigate allele frequency clines across environments
4. Consider gene ontology enrichment analysis
