---
title: "Broad-Sense Heritability (H²) Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Overview

This analysis calculates **broad-sense heritability (H²)** for multiple traits using linear mixed models. Heritability estimates are calculated both across all environments (overall) and for individual environments.

**Key Outputs:**

- Overall heritability across all environments
- Environment-specific heritability estimates
- Variance component decomposition (genetic, environment, replicate, residual)
- Summary statistics and visualizations

## Methods

**Heritability Calculation:**

Broad-sense heritability is calculated as:

$$H^2 = \frac{\sigma^2_G}{\sigma^2_G + \frac{\sigma^2_E}{n_E} + \frac{\sigma^2_{Rep}}{n_E \times n_{Rep}} + \frac{\sigma^2_e}{n_E \times n_{Rep}}}$$

Where:
- $\sigma^2_G$ = genetic variance
- $\sigma^2_E$ = environment variance
- $\sigma^2_{Rep}$ = replicate variance
- $\sigma^2_e$ = residual variance
- $n_E$ = number of environments
- $n_{Rep}$ = number of replicates

For single environments:

$$H^2 = \frac{\sigma^2_G}{\sigma^2_G + \frac{\sigma^2_{Rep}}{n_{Rep}} + \frac{\sigma^2_e}{n_{Rep}}}$$

## Configuration

**Edit these parameters for your data:**

```{r config}
# ============================================
# INPUT DATA CONFIGURATION
# ============================================

# Input CSV file with phenotype data
input_file <- "your_data.csv"

# Column names for experimental design
accession_col <- "accession"    # Genotype/accession identifier
environment_col <- "environment"  # Environment/location/trial
rep_col <- "rep"                 # Replicate/block

# Minimum observations required per trait
min_observations <- 10

# ============================================
# OUTPUT CONFIGURATION
# ============================================

# Prefix for output files
output_prefix <- "heritability"

# Heritability threshold for "high heritability" designation
h2_threshold <- 0.3
```

## Load Required Libraries

```{r libraries}
# Load required packages
required_packages <- c("lme4", "ggplot2", "dplyr", "tidyr", "knitr")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Required package '", pkg, "' is not installed.\n",
         "Please install it with: install.packages('", pkg, "')")
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

cat("All libraries loaded successfully.\n")
```

## Load and Prepare Data

```{r load_data}
# Check if input file exists
if (!file.exists(input_file)) {
  stop("Error: Input file '", input_file, "' not found!")
}

# Read data
data <- read.csv(input_file, na.strings = c("NA", ".", ""))

cat("Data loaded successfully!\n")
cat("Dimensions:", nrow(data), "rows x", ncol(data), "columns\n\n")

# Verify required columns exist
required_cols <- c(accession_col, environment_col, rep_col)
missing_cols <- setdiff(required_cols, names(data))

if (length(missing_cols) > 0) {
  stop("Error: Missing required columns: ", paste(missing_cols, collapse = ", "))
}

# Get trait names (all columns EXCEPT the design columns)
trait_names <- setdiff(names(data), required_cols)

cat("Found", length(trait_names), "traits to analyze:\n")
cat(paste(trait_names, collapse = ", "), "\n\n")

# Convert design factors to factors
data[[accession_col]] <- as.factor(data[[accession_col]])
data[[environment_col]] <- as.factor(data[[environment_col]])
data[[rep_col]] <- as.factor(data[[rep_col]])

# Convert trait columns to numeric
for (trait in trait_names) {
  data[[trait]] <- as.numeric(as.character(data[[trait]]))
}

# Display experimental design summary
cat("Experimental Design:\n")
cat("  Accessions:  ", nlevels(data[[accession_col]]), "\n")
cat("  Environments:", nlevels(data[[environment_col]]), "\n")
cat("  Replicates:  ", nlevels(data[[rep_col]]), "\n")
cat("  Total plots: ", nrow(data), "\n\n")

# Show first few rows
cat("First rows of data:\n")
print(head(data))
```

## Calculate Overall Heritability

Heritability across all environments combined.

```{r calculate_overall_h2}
cat("Calculating heritabilities across all environments...\n\n")

# Create empty list to store results
h2_overall_list <- list()

# Progress bar setup
pb <- txtProgressBar(min = 0, max = length(trait_names), style = 3)

for (i in seq_along(trait_names)) {
  trait <- trait_names[i]
  
  tryCatch({
    # Get clean data for this trait
    df <- data[!is.na(data[[trait]]), ]
    
    # Skip if too little data
    if (nrow(df) < min_observations) {
      cat("  [", i, "/", length(trait_names), "] ", trait, 
          " - Skipped (insufficient data)\n", sep = "")
      next
    }
    
    # Fit model with ALL effects as RANDOM
    formula_str <- paste0(trait, " ~ (1|", accession_col, ") + (1|", 
                          environment_col, ") + (1|", environment_col, ":", rep_col, ")")
    model <- lmer(as.formula(formula_str), data = df, REML = TRUE)
    
    # Extract variance components
    vc <- as.data.frame(VarCorr(model))
    
    var_g <- vc$vcov[vc$grp == accession_col]
    var_env <- vc$vcov[vc$grp == environment_col]
    var_rep <- vc$vcov[grepl(":", vc$grp) & grepl(environment_col, vc$grp)]
    var_e <- vc$vcov[vc$grp == "Residual"]
    
    # Handle cases where variance components might be missing
    if (length(var_g) == 0) var_g <- 0
    if (length(var_env) == 0) var_env <- 0
    if (length(var_rep) == 0) var_rep <- 0
    if (length(var_e) == 0) var_e <- 0
    
    # Calculate broad-sense heritability
    n_env <- length(unique(df[[environment_col]]))
    n_rep <- length(unique(df[[rep_col]]))
    
    h2 <- var_g / (var_g + var_env/n_env + var_rep/(n_env*n_rep) + var_e/(n_env*n_rep))
    
    # Store results
    h2_overall_list[[i]] <- data.frame(
      trait = trait,
      h2 = round(h2, 3),
      var_g = round(var_g, 4),
      var_env = round(var_env, 4),
      var_rep = round(var_rep, 4),
      var_res = round(var_e, 4),
      n_obs = nrow(df),
      n_env = n_env,
      n_rep = n_rep,
      stringsAsFactors = FALSE
    )
    
  }, error = function(e) {
    cat("  [", i, "/", length(trait_names), "] ", trait, 
        " - ERROR: ", e$message, "\n", sep = "")
  })
  
  setTxtProgressBar(pb, i)
}

close(pb)

# Combine all results
h2_overall_list <- h2_overall_list[!sapply(h2_overall_list, is.null)]

if (length(h2_overall_list) > 0) {
  h2_overall <- do.call(rbind, h2_overall_list)
  rownames(h2_overall) <- NULL
} else {
  h2_overall <- data.frame(
    trait = character(),
    h2 = numeric(),
    var_g = numeric(),
    var_env = numeric(),
    var_rep = numeric(),
    var_res = numeric(),
    n_obs = integer(),
    n_env = integer(),
    n_rep = integer(),
    stringsAsFactors = FALSE
  )
}

cat("\n\nOverall heritability calculation complete!\n")
cat("Successfully analyzed:", nrow(h2_overall), "traits\n")
```

### Overall Heritability Summary

```{r overall_summary}
if (nrow(h2_overall) > 0) {
  cat("\nSummary Statistics:\n")
  cat("==========================================\n")
  cat("Mean H²:   ", round(mean(h2_overall$h2, na.rm = TRUE), 3), "\n")
  cat("Median H²: ", round(median(h2_overall$h2, na.rm = TRUE), 3), "\n")
  cat("Range:     ", round(min(h2_overall$h2, na.rm = TRUE), 3), "-", 
      round(max(h2_overall$h2, na.rm = TRUE), 3), "\n")
  cat("Traits with H² >", h2_threshold, ":", 
      sum(!is.na(h2_overall$h2) & h2_overall$h2 > h2_threshold), "\n")
  cat("==========================================\n\n")
  
  # Display top traits
  h2_sorted <- h2_overall[order(-h2_overall$h2), ]
  
  cat("Top 10 Traits by Overall Heritability:\n")
  print(head(h2_sorted[, c("trait", "h2", "var_g", "var_env", "var_res")], 10))
} else {
  cat("No traits were successfully analyzed.\n")
}
```

### Overall Heritability Distribution

```{r plot_overall_h2, fig.width=10, fig.height=6}
if (nrow(h2_overall) > 0) {
  # Histogram
  ggplot(h2_overall, aes(x = h2)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
    geom_vline(xintercept = h2_threshold, color = "red", linetype = "dashed", size = 1) +
    labs(
      title = "Distribution of Overall Heritability (Across All Environments)",
      x = "Broad-Sense Heritability (H²)",
      y = "Number of Traits"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 12)
    )
}
```

### Variance Component Decomposition

```{r plot_variance_components, fig.width=12, fig.height=8}
if (nrow(h2_overall) > 0) {
  # Prepare data for stacked bar plot
  var_data <- h2_overall %>%
    select(trait, var_g, var_env, var_rep, var_res) %>%
    pivot_longer(cols = starts_with("var_"), 
                 names_to = "component", 
                 values_to = "variance") %>%
    mutate(
      component = factor(
        component,
        levels = c("var_g", "var_env", "var_rep", "var_res"),
        labels = c("Genetic", "Environment", "Replicate", "Residual")
      )
    )
  
  # Calculate total variance for percentage
  total_var <- h2_overall %>%
    mutate(total = var_g + var_env + var_rep + var_res) %>%
    select(trait, total)
  
  var_data <- var_data %>%
    left_join(total_var, by = "trait") %>%
    mutate(percentage = variance / total * 100)
  
  # Plot top 20 traits by heritability
  top_20_traits <- h2_sorted$trait[1:min(20, nrow(h2_sorted))]
  
  var_data_top20 <- var_data %>%
    filter(trait %in% top_20_traits) %>%
    mutate(trait = factor(trait, levels = rev(top_20_traits)))
  
  ggplot(var_data_top20, aes(x = trait, y = percentage, fill = component)) +
    geom_bar(stat = "identity", position = "stack") +
    coord_flip() +
    scale_fill_manual(
      values = c(
        "Genetic" = "#2ecc71",
        "Environment" = "#e74c3c",
        "Replicate" = "#f39c12",
        "Residual" = "#95a5a6"
      )
    ) +
    labs(
      title = "Variance Component Decomposition (Top 20 Traits by H²)",
      x = "Trait",
      y = "Percentage of Total Variance (%)",
      fill = "Component"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      legend.position = "top"
    )
}
```

## Calculate Environment-Specific Heritability

Heritability for each environment separately.

```{r calculate_env_specific_h2}
cat("Calculating heritabilities for individual environments...\n\n")

environments <- levels(data[[environment_col]])
h2_by_env <- list()

for (env in environments) {
  cat("Processing environment:", env, "\n")
  
  env_h2_list <- list()
  
  for (trait in trait_names) {
    tryCatch({
      # Get data for this environment and trait
      df <- data[data[[environment_col]] == env & !is.na(data[[trait]]), ]
      
      if (nrow(df) < min_observations) next
      
      # Fit model: accession and rep as RANDOM
      formula_str <- paste0(trait, " ~ (1|", accession_col, ") + (1|", rep_col, ")")
      model <- lmer(as.formula(formula_str), data = df, REML = TRUE)
      
      # Extract variance components
      vc <- as.data.frame(VarCorr(model))
      
      var_g <- vc$vcov[vc$grp == accession_col]
      var_rep <- vc$vcov[vc$grp == rep_col]
      var_e <- vc$vcov[vc$grp == "Residual"]
      
      # Handle missing components
      if (length(var_g) == 0) var_g <- 0
      if (length(var_rep) == 0) var_rep <- 0
      if (length(var_e) == 0) var_e <- 0
      
      # Calculate heritability for single environment
      n_rep <- length(unique(df[[rep_col]]))
      h2 <- var_g / (var_g + var_rep/n_rep + var_e/n_rep)
      
      # Store results
      env_h2_list[[trait]] <- data.frame(
        trait = trait,
        h2 = round(h2, 3),
        var_g = round(var_g, 4),
        var_rep = round(var_rep, 4),
        var_res = round(var_e, 4),
        n_obs = nrow(df),
        n_rep = n_rep,
        stringsAsFactors = FALSE
      )
      
    }, error = function(e) {})
  }
  
  # Combine results for this environment
  env_h2_list <- env_h2_list[!sapply(env_h2_list, is.null)]
  
  if (length(env_h2_list) > 0) {
    env_h2 <- do.call(rbind, env_h2_list)
    rownames(env_h2) <- NULL
    h2_by_env[[as.character(env)]] <- env_h2
    cat("  Completed:", nrow(env_h2), "traits\n")
  } else {
    cat("  No traits successfully analyzed\n")
  }
}

cat("\nEnvironment-specific heritability calculation complete!\n")
```

### Environment-Specific Summary

```{r env_specific_summary}
if (length(h2_by_env) > 0) {
  cat("\nEnvironment-Specific Heritability Summary:\n")
  cat("==========================================\n")
  
  env_summary <- data.frame()
  
  for (env in names(h2_by_env)) {
    env_data <- h2_by_env[[env]]
    env_summary <- rbind(
      env_summary,
      data.frame(
        environment = env,
        n_traits = nrow(env_data),
        mean_h2 = round(mean(env_data$h2, na.rm = TRUE), 3),
        median_h2 = round(median(env_data$h2, na.rm = TRUE), 3),
        high_h2_count = sum(!is.na(env_data$h2) & env_data$h2 > h2_threshold)
      )
    )
  }
  
  print(env_summary)
}
```

### Environment Comparison

```{r plot_env_comparison, fig.width=12, fig.height=6}
if (length(h2_by_env) > 0) {
  # Combine all environment data
  all_env_data <- data.frame()
  
  for (env in names(h2_by_env)) {
    env_data <- h2_by_env[[env]]
    env_data$environment <- env
    all_env_data <- rbind(all_env_data, env_data)
  }
  
  # Boxplot comparison
  ggplot(all_env_data, aes(x = environment, y = h2, fill = environment)) +
    geom_boxplot(alpha = 0.7) +
    geom_hline(yintercept = h2_threshold, color = "red", linetype = "dashed", size = 1) +
    labs(
      title = "Heritability Distribution by Environment",
      x = "Environment",
      y = "Broad-Sense Heritability (H²)"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 12),
      legend.position = "none"
    )
}
```

### Heritability Comparison: Overall vs. Environment-Specific

```{r plot_h2_comparison, fig.width=12, fig.height=8}
if (length(h2_by_env) > 0 && nrow(h2_overall) > 0) {
  # Reshape data for comparison
  comparison_data <- data.frame()
  
  for (env in names(h2_by_env)) {
    env_data <- h2_by_env[[env]] %>%
      select(trait, h2) %>%
      rename(env_h2 = h2) %>%
      mutate(environment = env)
    
    comparison_data <- rbind(comparison_data, env_data)
  }
  
  # Join with overall h2
  comparison_data <- comparison_data %>%
    left_join(h2_overall %>% select(trait, h2) %>% rename(overall_h2 = h2), 
              by = "trait")
  
  # Plot
  ggplot(comparison_data, aes(x = overall_h2, y = env_h2, color = environment)) +
    geom_point(size = 2, alpha = 0.6) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
    labs(
      title = "Environment-Specific vs. Overall Heritability",
      x = "Overall H² (Across Environments)",
      y = "Environment-Specific H²",
      color = "Environment"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 12),
      legend.position = "right"
    )
}
```

## High Heritability Traits

Traits with H² > `r h2_threshold` across all environments.

```{r high_h2_traits}
if (nrow(h2_overall) > 0) {
  high_h2 <- h2_overall %>%
    filter(!is.na(h2) & h2 > h2_threshold) %>%
    arrange(desc(h2))
  
  if (nrow(high_h2) > 0) {
    cat("Traits with Overall H² >", h2_threshold, ":\n")
    print(high_h2)
  } else {
    cat("No traits with H² >", h2_threshold, "\n")
  }
}
```

## Save Results

```{r save_results}
# Export overall heritabilities
overall_file <- paste0(output_prefix, "_overall.csv")
write.csv(h2_overall, overall_file, row.names = FALSE)
cat("Saved:", overall_file, "\n")

# Export environment-specific heritabilities
if (length(h2_by_env) > 0) {
  for (env in names(h2_by_env)) {
    env_file <- paste0(output_prefix, "_", env, ".csv")
    write.csv(h2_by_env[[env]], env_file, row.names = FALSE)
    cat("Saved:", env_file, "\n")
  }
}

# Export combined environment-specific results
if (length(h2_by_env) > 0) {
  all_env_file <- paste0(output_prefix, "_by_environment_combined.csv")
  all_env_combined <- data.frame()
  
  for (env in names(h2_by_env)) {
    env_data <- h2_by_env[[env]]
    env_data$environment <- env
    all_env_combined <- rbind(all_env_combined, env_data)
  }
  
  write.csv(all_env_combined, all_env_file, row.names = FALSE)
  cat("Saved:", all_env_file, "\n")
}
```

## Final Summary

```{r final_summary}
cat("\n================================================================================\n")
cat("HERITABILITY ANALYSIS COMPLETE!\n")
cat("================================================================================\n")
cat("Input file:              ", input_file, "\n")
cat("Total traits processed:  ", length(trait_names), "\n")
cat("Traits successfully analyzed (overall):", nrow(h2_overall), "\n")

if (nrow(h2_overall) > 0) {
  cat("\nOverall Heritability Statistics:\n")
  cat("  Mean H²:               ", round(mean(h2_overall$h2, na.rm = TRUE), 3), "\n")
  cat("  Traits with H² >", h2_threshold, ":", 
      sum(!is.na(h2_overall$h2) & h2_overall$h2 > h2_threshold), "\n")
}

if (length(h2_by_env) > 0) {
  cat("\nEnvironments analyzed:   ", length(h2_by_env), "\n")
  for (env in names(h2_by_env)) {
    env_data <- h2_by_env[[env]]
    cat("  ", env, ": ", nrow(env_data), " traits (mean H² = ", 
        round(mean(env_data$h2, na.rm = TRUE), 3), ")\n", sep = "")
  }
}

cat("\nOutput files created:\n")
cat("  - ", overall_file, "\n", sep = "")
if (length(h2_by_env) > 0) {
  for (env in names(h2_by_env)) {
    cat("  - ", output_prefix, "_", env, ".csv\n", sep = "")
  }
  cat("  - ", all_env_file, "\n", sep = "")
}
cat("================================================================================\n")
```

## Session Information

```{r session_info}
sessionInfo()
```

---

## Interpretation Guide

**Heritability (H²) Ranges:**

- **H² < 0.2**: Low heritability - trait heavily influenced by environment
- **0.2 ≤ H² < 0.5**: Moderate heritability - mixed genetic and environmental control
- **H² ≥ 0.5**: High heritability - trait strongly controlled by genetics

**Variance Components:**

- **Genetic variance (σ²_G)**: Variation due to genetic differences between accessions
- **Environment variance (σ²_E)**: Variation due to differences between environments
- **Replicate variance (σ²_Rep)**: Variation due to replicate/block effects within environments
- **Residual variance (σ²_e)**: Unexplained variation (measurement error, micro-environment, etc.)

**Applications:**

- Traits with high H² are good candidates for marker-assisted selection
- Traits with low H² may benefit from improved experimental design or environmental control
- Comparing overall vs. environment-specific H² reveals genotype × environment interactions
