---
title: "Environmental Association Analysis"
subtitle: "Generic workflow for analyzing environmental GWAS results"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 300
)
```

# Overview

This workflow analyzes environmental association (eGWAS) results to identify genomic regions associated with environmental adaptation. The analysis includes:

- Comparison of significant eGWAS markers with population genetic statistics (F~ST~, SPA)
- Gene annotation proximity analysis
- Permutation testing to assess enrichment
- Visualization of results

**Key outputs:**

1. Summary statistics of significant environmental associations
2. Comparison of eGWAS hits with F~ST~ and SPA outliers
3. Gene proximity analysis for significant markers
4. Figures comparing eGWAS markers to random background sets

---

# Configuration

## Species-Specific Parameters

**IMPORTANT:** Update these parameters for your species before running the analysis.

```{r config}
# ==============================================================================
# SPECIES CONFIGURATION - UPDATE THESE FOR YOUR DATASET
# ==============================================================================

# Project identifier (used in file naming)
species_name <- "YourSpecies"  # e.g., "Cranberry", "Pepper", "Lentil"

# FDR threshold for significance
fdr_threshold <- 0.20  # Adjust based on your study design

# Linkage disequilibrium decay distance (bp)
# This is the distance at which LD decays to r^2 = 0.2 in your species
# Estimate from LD decay analysis or use literature values
ld_decay_distance <- 17000  # bp; UPDATE THIS FOR YOUR SPECIES

# Model selection approach
# Options: "best_lambda", "manual", "all_models"
model_selection <- "best_lambda"  
# If "manual", specify which model to use:
manual_model <- "model2"  # e.g., "model1", "model2", "model3"

# Population category for wild/native germplasm
# This should match the category name in your metadata
wild_category <- "Wild"  # e.g., "Wild", "Landrace", "Native"

# Outlier percentile threshold (for SPA and FST)
outlier_percentile <- 0.999  # Top 0.1% of markers

# Number of permutations for significance testing
n_permutations <- 10000

# GFF annotation file (set to NULL if not available)
gff_file_name <- "species_annotations.gff"  # or NULL

# ==============================================================================
```

## Load Required Packages

```{r load-packages}
# Core analysis packages
library(tidyverse)
library(sommer)

# Genomic data handling
library(raster)
library(readxl)

# Statistical analysis
library(qvalue)
library(slider)

# Visualization
library(patchwork)
library(cowplot)
library(ggplot2)

# Optional packages (comment out if not available)
# library(neyhart)  # For custom color palettes
# library(snps)     # For SNP-specific operations
# library(LDheatmap) # For LD heatmaps

# Custom functions
source("startup.R")  # Load your custom startup script if available
```

## Define File Paths

```{r file-paths}
# ==============================================================================
# FILE PATH CONFIGURATION
# ==============================================================================

# Base directories
if (!exists("data_dir")) {
  data_dir <- "data"  # Directory with input data
}
if (!exists("result_dir")) {
  result_dir <- "results"  # Directory for saving results
}
if (!exists("fig_dir")) {
  fig_dir <- "figures"  # Directory for saving figures
}

# Create output directories if they don't exist
dir.create(result_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

# Input files
# These should be RData files containing the required objects
# Adjust the file names to match your naming convention
metadata_file <- file.path(data_dir, "population_metadata_and_genotypes.RData")
climate_file <- file.path(data_dir, "germplasm_origin_bioclim_data.RData")
gwas_results_file <- file.path(result_dir, "eaa_gwas_results.RData")
popgen_results_file <- file.path(result_dir, "population_genetics_stats.RData")

# GFF annotation file (if available)
if (!is.null(gff_file_name)) {
  gff_file <- file.path(data_dir, gff_file_name)
}

# ==============================================================================
```

---

# Data Loading

## Load Genomic and Environmental Data

```{r load-data}
# Load population metadata and genotypes
# Expected objects: pop_metadata, geno_mat_*, geno_hmp_*, snp_info, K_wild
load(metadata_file)

# Load environmental/climate data
# Expected objects: eaa_environmental_data, eaa_environmental_vars
load(climate_file)

# Load eGWAS results
# Expected objects: gwas_out, p_lambda, eaa_gwas_sigmar, eaa_environmental_vars
load(gwas_results_file)

# Load population genetics results
# Expected objects: global_marker_fst, spa_out
load(popgen_results_file)

cat("Data loaded successfully!\n")
```

## Data Structure Check

```{r check-data}
# Verify that required objects exist
required_objects <- c(
  "pop_metadata", "snp_info", "gwas_out", 
  "global_marker_fst", "spa_out"
)

missing_objects <- required_objects[!sapply(required_objects, exists)]

if (length(missing_objects) > 0) {
  stop("Missing required objects: ", paste(missing_objects, collapse = ", "))
}

cat("All required data objects present.\n")
cat("\nDataset dimensions:\n")
cat("  SNPs:", nrow(snp_info), "\n")
cat("  Individuals:", nrow(pop_metadata), "\n")
cat("  GWAS results:", nrow(gwas_out), "\n")
cat("  FST values:", nrow(global_marker_fst), "\n")
cat("  SPA scores:", nrow(spa_out), "\n")
```

---

# Data Preparation

## Prepare Genotype Matrices

```{r prep-genotypes}
# Select the appropriate filtered genotype matrices
# Adjust these variable names to match your naming convention
# Common patterns: geno_mat_wild_filter3, geno_mat_wild_final, etc.

# Find available genotype matrix objects
geno_mat_objects <- ls(pattern = "^geno_mat_.*wild.*filter")
geno_hmp_objects <- ls(pattern = "^geno_hmp_.*wild.*filter")

if (length(geno_mat_objects) > 0) {
  # Use the last (most filtered) version
  geno_mat_wild <- get(sort(geno_mat_objects, decreasing = TRUE)[1])
  cat("Using genotype matrix:", sort(geno_mat_objects, decreasing = TRUE)[1], "\n")
} else {
  stop("No filtered wild genotype matrix found. Expected pattern: geno_mat_*wild*filter*")
}

if (length(geno_hmp_objects) > 0) {
  geno_hmp_wild <- get(sort(geno_hmp_objects, decreasing = TRUE)[1])
  cat("Using hapmap format:", sort(geno_hmp_objects, decreasing = TRUE)[1], "\n")
} else {
  stop("No filtered wild hapmap found. Expected pattern: geno_hmp_*wild*filter*")
}

# Also prepare all-individuals matrix if available
geno_mat_all_objects <- ls(pattern = "^geno_mat_all.*filter")
if (length(geno_mat_all_objects) > 0) {
  geno_mat_all <- get(sort(geno_mat_all_objects, decreasing = TRUE)[1])
}
```

## Prepare Environmental Data

```{r prep-environmental}
# Center and scale environmental variables if not already done
if (exists("eaa_environmental_vars") && exists("eaa_environmental_data")) {
  
  eaa_environmental_data1 <- eaa_environmental_data %>%
    select(location_abbr, all_of(eaa_environmental_vars$variable))
  
  cat("Environmental variables prepared:", ncol(eaa_environmental_data1) - 1, "\n")
  
} else {
  cat("Warning: Environmental data not found or already prepared.\n")
}
```

## Prepare Bioclimate Data for Analysis

```{r prep-bioclim}
# Prepare bioclimate data matched to genetic samples
germplasm_bioclim_data <- pop_metadata %>%
  filter(category == wild_category, !is.na(latitude)) %>%
  select(individual, latitude, longitude) %>%
  left_join(., eaa_environmental_data1, by = c("individual" = "location_abbr")) %>%
  as.data.frame() %>%
  filter(individual %in% colnames(K_wild))

cat("Individuals with both genetic and environmental data:", nrow(germplasm_bioclim_data), "\n")
```

## Prepare Marker Data

```{r prep-markers}
# Subset genotype matrix to individuals with environmental data
geno_mat_wild1 <- geno_mat_wild[germplasm_bioclim_data$individual, ]

# Subset hapmap format
geno_hmp_wild1 <- geno_hmp_wild %>%
  select(marker:alleles, row.names(geno_mat_wild1))

# Prepare SNP info in standard format
snp_info1 <- snp_info %>%
  rename(Locus = marker, LG = chrom, Position = pos) %>%
  as.data.frame()

# Split marker matrix by chromosome
geno_mat_list <- geno_hmp_wild %>%
  select(marker, chrom) %>%
  split(.$chrom) %>% 
  map("marker") %>%
  map(~geno_mat_wild1[, .x, drop = FALSE])

# Prepare genotype data for downstream analysis
geno <- geno_hmp_wild %>%
  select(marker, chrom, pos, all_of(colnames(K_wild))) %>%
  mutate(across(all_of(colnames(K_wild)), ~. - 1)) %>%
  as.data.frame()

cat("Marker data prepared.\n")
cat("  Chromosomes:", length(unique(snp_info1$LG)), "\n")
cat("  Markers:", nrow(geno_hmp_wild1), "\n")
```

---

# Model Selection and Significant Markers

## Select Best GWAS Model

```{r model-selection}
if (model_selection == "best_lambda") {
  
  # Use the model with best genomic inflation factor (lambda)
  if (exists("p_lambda") && "best_model" %in% names(p_lambda)) {
    
    gwas_out_best_model <- gwas_out %>%
      inner_join(., subset(p_lambda, best_model == "*", c(model, variable), drop = TRUE))
    
    # Show model selection
    cat("Best models by environmental variable:\n")
    print(xtabs(~ model, p_lambda, subset = best_model == "*"))
    
  } else {
    warning("p_lambda object not found or doesn't have best_model column. Using all models.")
    gwas_out_best_model <- gwas_out
  }
  
} else if (model_selection == "manual") {
  
  # Use manually specified model
  gwas_out_best_model <- gwas_out %>% 
    filter(model == manual_model)
  
  cat("Using manually selected model:", manual_model, "\n")
  
} else if (model_selection == "all_models") {
  
  # Use all models
  gwas_out_best_model <- gwas_out
  cat("Using all models.\n")
  
}

cat("GWAS results after model selection:", nrow(gwas_out_best_model), "\n")
```

## Extract Significant Markers

```{r extract-sigmar}
# Extract significant markers based on FDR threshold
# This should have been generated in the GWAS analysis
if (exists("eaa_gwas_sigmar")) {
  egwas_sigmar <- eaa_gwas_sigmar
} else {
  # If not pre-computed, calculate from GWAS results
  egwas_sigmar <- gwas_out_best_model %>%
    group_by(variable) %>%
    mutate(qvalue = qvalue(pvalue)$qvalues) %>%
    filter(qvalue <= fdr_threshold) %>%
    ungroup()
}

# Summary statistics
n_total_assoc <- nrow(egwas_sigmar)
n_unique_markers <- n_distinct(egwas_sigmar$marker)
n_unique_vars <- n_distinct(egwas_sigmar$variable)

cat("\n=== Significant Association Summary ===\n")
cat("Total associations (marker-variable pairs):", n_total_assoc, "\n")
cat("Unique markers:", n_unique_markers, "\n")
cat("Environmental variables with associations:", n_unique_vars, "\n")
```

## Significant Markers by Environmental Variable

```{r sigmar-by-variable}
# Number of unique markers per environmental variable
sigmar_per_var <- egwas_sigmar %>%
  left_join(., select(eaa_environmental_vars, -class)) %>%
  group_by(full_name, variable) %>%
  summarize(n_markers = n_distinct(marker), .groups = "drop") %>%
  arrange(desc(n_markers))

cat("\nTop environmental variables by number of significant markers:\n")
print(head(sigmar_per_var, 10))

# Create a bar plot
if (nrow(sigmar_per_var) > 0) {
  ggplot(sigmar_per_var, aes(x = reorder(full_name, n_markers), y = n_markers)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(
      title = "Significant Markers per Environmental Variable",
      x = "Environmental Variable",
      y = "Number of Significant Markers"
    ) +
    theme_minimal()
}
```

## Markers by Environmental Variable Class

```{r sigmar-by-class}
# Check if environmental variable classes are available
if ("class" %in% names(eaa_environmental_vars)) {
  
  # Identify variables of interest (if pre-defined)
  if (exists("env_variables_of_interest")) {
    egwas_sigmar_cat <- egwas_sigmar %>%
      left_join(., eaa_environmental_vars) %>%
      filter(variable %in% env_variables_of_interest)
  } else {
    egwas_sigmar_cat <- egwas_sigmar %>%
      left_join(., eaa_environmental_vars)
  }
  
  # Summarize by class
  sigmar_by_class <- egwas_sigmar_cat %>%
    group_by(class, variable) %>%
    summarize(nSigMar = n(), .groups = "drop_last") %>%
    summarize(
      nVariable = n_distinct(variable), 
      nSigMar = sum(nSigMar),
      .groups = "drop"
    )
  
  cat("\nSignificant markers by environmental variable class:\n")
  print(sigmar_by_class)
  
  # Visualize
  ggplot(sigmar_by_class, aes(x = reorder(class, nSigMar), y = nSigMar)) +
    geom_col(aes(fill = class), show.legend = FALSE) +
    coord_flip() +
    labs(
      title = "Significant Markers by Environmental Class",
      x = "Environmental Class",
      y = "Number of Marker-Variable Associations"
    ) +
    theme_minimal()
  
} else {
  cat("\nEnvironmental variable classes not available.\n")
}
```

## Characterize Significant Markers

```{r characterize-sigmar}
# How many environmental variables does each marker associate with?
egwas_sigmar_unique <- egwas_sigmar %>%
  select(marker, variable) %>%
  group_by(marker) %>%
  summarize(
    nVar = n(),
    variables = paste(variable, collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(nVar))

cat("\nMarkers with most environmental associations:\n")
print(head(egwas_sigmar_unique, 10))

# Distribution of associations per marker
ggplot(egwas_sigmar_unique, aes(x = nVar)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Environmental Associations per Marker",
    x = "Number of Environmental Variables",
    y = "Number of Markers"
  ) +
  theme_minimal()
```

---

# Comparison with Population Genetic Outliers

## Identify SPA and F~ST~ Outliers

```{r identify-outliers}
# Calculate quantile thresholds
spa_threshold <- quantile(spa_out$spa_score, outlier_percentile)
fst_threshold <- quantile(global_marker_fst$Fst, outlier_percentile)

# Identify outliers
spa_outliers <- spa_out %>%
  filter(spa_score >= spa_threshold)

fst_outliers <- global_marker_fst %>%
  filter(Fst >= fst_threshold)

cat("\n=== Population Genetic Outlier Summary ===\n")
cat("SPA outliers (top", (1 - outlier_percentile) * 100, "%):", nrow(spa_outliers), "\n")
cat("FST outliers (top", (1 - outlier_percentile) * 100, "%):", nrow(fst_outliers), "\n")
cat("Overlap between FST and SPA outliers:", 
    length(intersect(fst_outliers$marker, spa_outliers$marker)), "\n")
```

## Compare eGWAS Markers with SPA Scores

```{r egwas-spa}
# Get SPA scores for significant GWAS markers
eaa_gwas_spa_scores <- egwas_sigmar %>%
  distinct(marker) %>%
  left_join(., select(spa_out, marker, spa_score))

# How many eGWAS markers are also SPA outliers?
n_egwas_spa_overlap <- sum(eaa_gwas_spa_scores$spa_score >= spa_threshold)

cat("\neGWAS markers that are also SPA outliers:", n_egwas_spa_overlap, "\n")

if (n_egwas_spa_overlap > 0) {
  cat("\nThese markers are:\n")
  print(
    eaa_gwas_spa_scores %>%
      filter(spa_score >= spa_threshold) %>%
      arrange(desc(spa_score))
  )
}

# Calculate mean SPA score for eGWAS markers
gwas_mean_spa <- mean(eaa_gwas_spa_scores$spa_score)
gwas_sd_spa <- sd(eaa_gwas_spa_scores$spa_score)

cat("\nMean SPA score for eGWAS markers:", round(gwas_mean_spa, 4), 
    "± ", round(gwas_sd_spa, 4), "\n")
cat("Mean SPA score for all markers:", round(mean(spa_out$spa_score), 4), 
    "± ", round(sd(spa_out$spa_score), 4), "\n")
```

## Permutation Test: eGWAS vs Random Markers (SPA)

```{r permute-spa}
# Number of eGWAS markers
n_eaa_markers <- nrow(eaa_gwas_spa_scores)

# Exclude eGWAS markers from background
spa_out_nogwas_vec <- spa_out %>%
  filter(!marker %in% eaa_gwas_spa_scores$marker) %>%
  pull(spa_score)

# Permutation test
set.seed(209)
random_snps_spa <- replicate(
  n = n_permutations, 
  mean(sample(x = spa_out_nogwas_vec, size = n_eaa_markers))
)

# Calculate empirical p-value
p_value_spa <- mean(random_snps_spa >= gwas_mean_spa)

cat("\n=== Permutation Test Results (SPA) ===\n")
cat("Empirical p-value:", round(p_value_spa, 4), "\n")

# Wilcoxon test
wilcox_result_spa <- wilcox.test(
  x = eaa_gwas_spa_scores$spa_score, 
  y = spa_out_nogwas_vec
)

cat("Wilcoxon test p-value:", format.pval(wilcox_result_spa$p.value, digits = 3), "\n")
```

## Visualize SPA Permutation Results

```{r plot-spa-permutation, fig.height=4, fig.width=6}
# Create plotting data
plot_data_spa <- tibble(spa_score = random_snps_spa)

# Create plot
g_random_eaa_spa <- ggplot(plot_data_spa, aes(x = spa_score)) +
  geom_histogram(bins = 50, fill = "gray70", color = "white") +
  geom_vline(xintercept = gwas_mean_spa, color = "blue", linewidth = 1) +
  annotate(
    "text", 
    x = gwas_mean_spa, 
    y = Inf, 
    label = "EAA significant\nSNPs", 
    color = "blue",
    hjust = -0.1, 
    vjust = 1.5, 
    size = 3.5
  ) +
  annotate(
    "text",
    x = min(random_snps_spa),
    y = Inf,
    label = paste0("Random sets of\n", n_eaa_markers, " SNPs"),
    color = "gray40",
    hjust = 0,
    vjust = 1.5,
    size = 3.5
  ) +
  labs(
    title = "EAA Markers vs Random Background: SPA Scores",
    subtitle = paste0("Permutation test p-value = ", round(p_value_spa, 4)),
    x = "Mean SPA Score",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

print(g_random_eaa_spa)

# Save figure
ggsave(
  filename = paste0(species_name, "_eaaSPA_versus_randomSPA.png"), 
  plot = g_random_eaa_spa, 
  path = fig_dir,
  height = 4, 
  width = 6, 
  dpi = 300
)
```

## Compare eGWAS Markers with F~ST~ Values

```{r egwas-fst}
# Get FST values for significant GWAS markers
eaa_gwas_fst <- egwas_sigmar %>%
  distinct(marker) %>%
  left_join(., select(global_marker_fst, marker, Fst))

# How many eGWAS markers are also FST outliers?
n_egwas_fst_overlap <- sum(eaa_gwas_fst$Fst >= fst_threshold, na.rm = TRUE)

cat("\neGWAS markers that are also FST outliers:", n_egwas_fst_overlap, "\n")

if (n_egwas_fst_overlap > 0) {
  cat("\nThese markers are:\n")
  print(
    eaa_gwas_fst %>%
      filter(Fst >= fst_threshold) %>%
      arrange(desc(Fst))
  )
}

# Calculate mean FST for eGWAS markers
gwas_mean_fst <- mean(eaa_gwas_fst$Fst, na.rm = TRUE)
gwas_sd_fst <- sd(eaa_gwas_fst$Fst, na.rm = TRUE)

cat("\nMean FST for eGWAS markers:", round(gwas_mean_fst, 4), 
    "± ", round(gwas_sd_fst, 4), "\n")
cat("Mean FST for all markers:", round(mean(global_marker_fst$Fst, na.rm = TRUE), 4), 
    "± ", round(sd(global_marker_fst$Fst, na.rm = TRUE), 4), "\n")
```

## Permutation Test: eGWAS vs Random Markers (F~ST~)

```{r permute-fst}
# Number of eGWAS markers
n_eaa_markers_fst <- nrow(eaa_gwas_fst)

# Exclude eGWAS markers from background
fst_nogwas_vec <- global_marker_fst %>%
  filter(!marker %in% eaa_gwas_fst$marker) %>%
  pull(Fst)

# Permutation test
set.seed(209)
random_snps_fst <- replicate(
  n = n_permutations, 
  mean(sample(x = fst_nogwas_vec, size = n_eaa_markers_fst))
)

# Calculate empirical p-value
p_value_fst <- mean(random_snps_fst >= gwas_mean_fst)

cat("\n=== Permutation Test Results (FST) ===\n")
cat("Empirical p-value:", format.pval(p_value_fst, digits = 3), "\n")

# Wilcoxon test
wilcox_result_fst <- wilcox.test(
  x = eaa_gwas_fst$Fst, 
  y = fst_nogwas_vec
)

cat("Wilcoxon test p-value:", format.pval(wilcox_result_fst$p.value, digits = 3), "\n")
```

## Visualize F~ST~ Permutation Results

```{r plot-fst-permutation, fig.height=4, fig.width=6}
# Create plotting data
plot_data_fst <- tibble(fst = random_snps_fst)

# Create plot
g_random_eaa_fst <- ggplot(plot_data_fst, aes(x = fst)) +
  geom_histogram(bins = 50, fill = "gray70", color = "white") +
  geom_vline(xintercept = gwas_mean_fst, color = "red", linewidth = 1) +
  annotate(
    "text", 
    x = gwas_mean_fst, 
    y = Inf, 
    label = "EAA significant\nSNPs", 
    color = "red",
    hjust = 1.1, 
    vjust = 1.5, 
    size = 3.5
  ) +
  annotate(
    "text",
    x = min(random_snps_fst),
    y = Inf,
    label = paste0("Random sets of\n", n_eaa_markers, " SNPs"),
    color = "gray40",
    hjust = 0,
    vjust = 1.5,
    size = 3.5
  ) +
  labs(
    title = expression("EAA Markers vs Random Background:" ~ italic(F)[ST] ~ "Values"),
    subtitle = paste0("Permutation test p-value ", 
                     ifelse(p_value_fst < 0.001, "< 0.001", 
                           paste("=", round(p_value_fst, 4)))),
    x = expression('Mean' ~ italic(F)[ST]),
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

print(g_random_eaa_fst)

# Save figure
ggsave(
  filename = paste0(species_name, "_eaaFST_versus_randomFST.png"), 
  plot = g_random_eaa_fst, 
  path = fig_dir,
  height = 4, 
  width = 6, 
  dpi = 300
)
```

## Correlation Between GWAS Scores and Population Statistics

```{r gwas-correlations}
# Correlation between eGWAS p-values and SPA scores
egwas_score_spa <- gwas_out_best_model %>%
  filter(variable %in% unique(egwas_sigmar$variable)) %>%
  left_join(., select(spa_out, marker, spa_score)) %>%
  group_by(variable) %>%
  summarize(
    spa_egwas_cor = cor(sqrt(score), spa_score, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  arrange(desc(abs(spa_egwas_cor)))

cat("\nCorrelation between eGWAS test statistics and SPA scores:\n")
print(head(egwas_score_spa, 10))

# Correlation between eGWAS p-values and FST
egwas_score_fst <- gwas_out_best_model %>%
  filter(variable %in% unique(egwas_sigmar$variable)) %>%
  left_join(., select(global_marker_fst, marker, Fst)) %>%
  group_by(variable) %>%
  summarize(
    fst_egwas_cor = cor(sqrt(score), Fst, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  arrange(desc(abs(fst_egwas_cor)))

cat("\nCorrelation between eGWAS test statistics and FST:\n")
print(head(egwas_score_fst, 10))
```

---

# Gene Annotation Analysis

This section requires a GFF annotation file for your species.

```{r gene-annotation, eval = !is.null(gff_file_name)}
# Check if GFF file exists
if (!is.null(gff_file_name) && file.exists(gff_file)) {
  
  cat("\n=== Gene Annotation Analysis ===\n")
  cat("Loading GFF file:", gff_file, "\n")
  
  # Read GFF file
  species_gff <- ape::read.gff(file = gff_file)
  
  cat("GFF entries loaded:", nrow(species_gff), "\n")
  
  # Combine all significant markers (eGWAS, FST, SPA)
  all_sigmars <- bind_rows(
    mutate(egwas_sigmar, class = "EAA", variable = variable),
    mutate(fst_outliers, variable = "Fst", class = "Fst"),
    mutate(spa_outliers, variable = "SPA", class = "SPA")
  ) %>% 
    select(class, variable, marker, chrom, pos)
  
  cat("Total significant markers across all analyses:", nrow(all_sigmars), "\n")
  
} else {
  cat("\nGFF annotation file not available. Skipping gene annotation analysis.\n")
}
```

## Find Genes Near Significant Markers

```{r find-nearby-genes, eval = !is.null(gff_file_name)}
if (!is.null(gff_file_name) && file.exists(gff_file)) {
  
  # Analyze each significant marker for nearby genes
  all_sigmar_nearby_annotation <- all_sigmars %>%
    group_by(class, variable, marker) %>%
    group_modify(~ {
      row <- .x
      
      # Get marker location
      snp_info_i <- snp_info %>% filter(marker == row$marker[1])
      
      # Find annotations within this SNP
      snp_i_within_annotation <- species_gff %>%
        filter(
          seqid == as.character(snp_info_i$chrom),
          type != "chromosome"
        ) %>%
        filter(
          (start <= snp_info_i$pos & end >= snp_info_i$pos) | 
          (start >= snp_info_i$pos & end <= snp_info_i$pos)
        )
      
      # Find annotations near this SNP (within LD decay distance)
      sigmar_nearby_ann <- species_gff %>%
        filter(
          seqid == as.character(snp_info_i$chrom), 
          type != "chromosome"
        ) %>%
        filter(
          abs(snp_info_i$pos - start) <= ld_decay_distance | 
          abs(snp_info_i$pos - end) <= ld_decay_distance
        ) %>%
        mutate(
          start_distance = abs(start - snp_info_i$pos),
          end_distance = abs(end - snp_info_i$pos),
          min_distance = pmin(start_distance, end_distance)
        ) %>%
        arrange(min_distance)
      
      tibble(
        nearby_annotation = list(sigmar_nearby_ann),
        within_annotations = list(snp_i_within_annotation)
      )
    }) %>%
    ungroup() %>%
    mutate(region = ifelse(map_int(within_annotations, nrow) > 0, "genic", "non-genic"))
  
  cat("\nSignificant markers by genomic region:\n")
  print(table(all_sigmar_nearby_annotation$region))
  
  # Count markers with nearby genes
  markers_with_genes <- all_sigmar_nearby_annotation %>%
    filter(map_int(nearby_annotation, nrow) > 0)
  
  cat("\nSignificant markers with nearby genes (within", ld_decay_distance, "bp):", 
      nrow(markers_with_genes), "\n")
  
}
```

## Analyze All Markers for Comparison

```{r analyze-all-markers, eval = !is.null(gff_file_name)}
if (!is.null(gff_file_name) && file.exists(gff_file)) {
  
  # Analyze all markers for comparison
  all_markers_nearby_annotation <- snp_info %>%
    group_by(marker) %>%
    group_modify(~ {
      snp_info_i <- .x
      
      # Find annotations within this SNP
      snp_i_within_annotation <- species_gff %>%
        filter(
          seqid == as.character(snp_info_i$chrom),
          type != "chromosome"
        ) %>%
        filter(
          (start <= snp_info_i$pos & end >= snp_info_i$pos) | 
          (start >= snp_info_i$pos & end <= snp_info_i$pos)
        )
      
      # Find annotations near this SNP
      sigmar_nearby_ann <- species_gff %>%
        filter(
          seqid == as.character(snp_info_i$chrom), 
          type != "chromosome"
        ) %>%
        filter(
          abs(snp_info_i$pos - start) <= ld_decay_distance | 
          abs(snp_info_i$pos - end) <= ld_decay_distance
        ) %>%
        mutate(
          start_distance = abs(start - snp_info_i$pos),
          end_distance = abs(end - snp_info_i$pos),
          min_distance = pmin(start_distance, end_distance)
        ) %>%
        arrange(min_distance)
      
      # Get minimum distance to nearest gene
      gene_distances <- sigmar_nearby_ann %>%
        filter(type == "gene") %>%
        pull(min_distance)
      
      tibble(
        min_distance_nearby_gene = ifelse(
          length(gene_distances) > 0, 
          min(gene_distances), 
          NA_real_
        ),
        region = ifelse(nrow(snp_i_within_annotation) == 0, "non-genic", "genic")
      )
    }) %>%
    ungroup()
  
  cat("\nAll markers by genomic region:\n")
  print(table(all_markers_nearby_annotation$region))
  
}
```

## Compare Genic Enrichment

```{r compare-genic-enrichment, eval = !is.null(gff_file_name)}
if (!is.null(gff_file_name) && file.exists(gff_file)) {
  
  # Proportion of genic SNPs in significant vs all SNPs
  prop_genic_sigmar <- all_sigmar_nearby_annotation %>%
    distinct(marker, region) %>%
    pull(region) %>%
    table() %>%
    prop.table()
  
  prop_genic_all <- all_markers_nearby_annotation %>%
    distinct(marker, region) %>%
    pull(region) %>%
    table() %>%
    prop.table()
  
  cat("\n=== Genic Region Enrichment ===\n")
  cat("\nSignificant markers:\n")
  print(round(prop_genic_sigmar, 3))
  
  cat("\nAll markers:\n")
  print(round(prop_genic_all, 3))
  
  # Chi-square test for enrichment
  genic_table <- rbind(
    table(all_sigmar_nearby_annotation %>% 
            distinct(marker, region) %>% 
            pull(region)),
    table(all_markers_nearby_annotation %>% 
            distinct(marker, region) %>% 
            pull(region))
  )
  rownames(genic_table) <- c("Significant", "All")
  
  cat("\nContingency table:\n")
  print(genic_table)
  
  if (all(dim(genic_table) == c(2, 2))) {
    chisq_result <- chisq.test(genic_table)
    cat("\nChi-square test p-value:", format.pval(chisq_result$p.value, digits = 3), "\n")
  }
  
}
```

---

# Save Results

```{r save-results}
# Create list of results to save
results_to_save <- c(
  "egwas_sigmar",
  "eaa_gwas_spa_scores",
  "eaa_gwas_fst",
  "spa_outliers",
  "fst_outliers",
  "gwas_out_best_model",
  "p_value_spa",
  "p_value_fst"
)

# Add annotation results if they exist
if (exists("all_sigmar_nearby_annotation")) {
  results_to_save <- c(results_to_save, 
                       "all_sigmar_nearby_annotation",
                       "all_markers_nearby_annotation")
}

# Save results
output_file <- file.path(result_dir, 
                        paste0(species_name, "_egwas_analysis_results.RData"))

save(list = results_to_save, file = output_file)

cat("\n=== Results Saved ===\n")
cat("Output file:", output_file, "\n")
cat("Objects saved:", length(results_to_save), "\n")
```

---

# Summary

```{r summary}
cat("\n")
cat("=" %R% 70, "\n")
cat("ENVIRONMENTAL ASSOCIATION ANALYSIS SUMMARY\n")
cat("=" %R% 70, "\n\n")

cat("Species:", species_name, "\n")
cat("Analysis date:", format(Sys.Date(), "%B %d, %Y"), "\n\n")

cat("Dataset:\n")
cat("  Total SNPs:", nrow(snp_info), "\n")
cat("  Individuals with environmental data:", nrow(germplasm_bioclim_data), "\n\n")

cat("Significant Associations:\n")
cat("  Total marker-variable associations:", n_total_assoc, "\n")
cat("  Unique markers:", n_unique_markers, "\n")
cat("  Environmental variables:", n_unique_vars, "\n\n")

cat("Population Genetic Outliers:\n")
cat("  SPA outliers:", nrow(spa_outliers), "\n")
cat("  FST outliers:", nrow(fst_outliers), "\n")
cat("  eGWAS markers that are SPA outliers:", n_egwas_spa_overlap, "\n")
cat("  eGWAS markers that are FST outliers:", n_egwas_fst_overlap, "\n\n")

cat("Statistical Tests:\n")
cat("  SPA permutation test p-value:", round(p_value_spa, 4), "\n")
cat("  FST permutation test p-value:", 
    ifelse(p_value_fst < 0.001, "< 0.001", round(p_value_fst, 4)), "\n\n")

if (exists("all_sigmar_nearby_annotation")) {
  cat("Gene Annotation:\n")
  cat("  LD decay distance used:", ld_decay_distance, "bp\n")
  cat("  Significant markers in genic regions:", 
      sum(all_sigmar_nearby_annotation$region == "genic"), "\n")
  cat("  Significant markers with nearby genes:", 
      nrow(markers_with_genes), "\n\n")
}

cat("Output Files:\n")
cat("  Results:", basename(output_file), "\n")
cat("  Figure 1:", paste0(species_name, "_eaaSPA_versus_randomSPA.png"), "\n")
cat("  Figure 2:", paste0(species_name, "_eaaFST_versus_randomFST.png"), "\n")

cat("\n")
cat("=" %R% 70, "\n")
```

---

# Session Information

```{r session-info}
sessionInfo()
```
