---
title: "Geographic Sample Mapping with Regional Zoom"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 14,
  fig.height = 8,
  fig.align = 'center'
)
```

# Introduction

## What This Analysis Does

This workflow creates publication-quality maps showing geographic distribution of samples with:

- **World/continental overview map**: Shows all sample locations
- **Regional zoom map**: Focused view on data extent
- **Optional inset maps**: For specific regions of interest
- **Customizable styling**: Colors, scales, annotations

## Applications

- **Sample collection maps**: Show sampling locations
- **Germplasm diversity**: Display collection sites
- **Population sampling**: Visualize study design
- **Conservation**: Map priority areas
- **Publications**: High-quality figure generation

# Setup

## Load Required Libraries

```{r load-libraries}
# Load required packages
library(tidyverse)         # Data manipulation and ggplot2
library(sf)                # Spatial data handling
library(rnaturalearth)     # Map data
library(rnaturalearthdata) # Additional map data
library(ggspatial)         # Scale bars and north arrows
library(patchwork)         # Combine plots
library(knitr)             # Table formatting
library(kableExtra)        # Enhanced tables

cat("All required packages loaded successfully\n")
```

## Install High-Resolution Map Data (Optional)

```{r install-hires, eval=FALSE}
# For higher resolution maps (optional but recommended)
install.packages("rnaturalearthhires", 
                 repos = "http://packages.ropensci.org", 
                 type = "source")
```

## Set Parameters

```{r set-parameters}
# ==================== INPUT FILE ====================
DATA_CSV <- "your_sample_locations.csv"             # Path to coordinate CSV

# ==================== COLUMN NAMES ====================
SAMPLE_ID_COLUMN <- "SampleID"                      # Sample ID column
LATITUDE_COLUMN <- "Latitude"                       # Latitude column
LONGITUDE_COLUMN <- "Longitude"                     # Longitude column
GROUP_COLUMN <- "Population"                        # Grouping variable (optional)

# ==================== MAP PARAMETERS ====================
# World/Overview map extent (set to NULL for automatic)
WORLD_XLIM <- c(-180, 180)                          # Longitude range
WORLD_YLIM <- c(-90, 90)                            # Latitude range

# Regional map buffer (degrees to add around data extent)
REGIONAL_BUFFER <- 2                                # Buffer in degrees

# ==================== VISUAL PARAMETERS ====================
POINT_SIZE <- 3                                     # Point size
POINT_ALPHA <- 0.8                                  # Point transparency
COLOR_PALETTE <- "Set1"                             # Color palette
MAP_FILL_COLOR <- "wheat"                           # Land color
OCEAN_COLOR <- "lightblue"                          # Ocean color

# ==================== OUTPUT OPTIONS ====================
OUTPUT_PREFIX <- "sample_map"                       # Prefix for output files
OUTPUT_DIR <- "maps"                                # Output directory
SAVE_DPI <- 300                                     # Resolution for saved maps

# Create output directory
if (!dir.exists(OUTPUT_DIR)) {
  dir.create(OUTPUT_DIR, recursive = TRUE)
}

# Display parameters
param_table <- data.frame(
  Parameter = c("Data CSV", "Sample ID Column", "Latitude Column", 
                "Longitude Column", "Group Column", "Regional Buffer"),
  Value = c(DATA_CSV, SAMPLE_ID_COLUMN, LATITUDE_COLUMN, 
            LONGITUDE_COLUMN, GROUP_COLUMN, 
            paste(REGIONAL_BUFFER, "degrees"))
)

kable(param_table, caption = "Analysis Parameters") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Data Import and Preparation

## Read Sample Coordinates

```{r read-data}
cat("Reading sample data:", DATA_CSV, "\n")

# Check if file exists
if (!file.exists(DATA_CSV)) {
  stop("Data file not found: ", DATA_CSV)
}

# Read CSV
sample_data <- read.csv(DATA_CSV, stringsAsFactors = FALSE)

cat("Data loaded successfully\n")
cat("  Total samples:", nrow(sample_data), "\n")

# Display first few rows
kable(head(sample_data), caption = "Sample Data Preview") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%")

# Check required columns exist
required_cols <- c(SAMPLE_ID_COLUMN, LATITUDE_COLUMN, LONGITUDE_COLUMN)
missing_cols <- setdiff(required_cols, colnames(sample_data))

if (length(missing_cols) > 0) {
  stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
}
```

## Filter to Valid Coordinates

```{r filter-valid}
cat("Filtering to valid coordinates...\n")

# Count missing before filtering
missing_before <- sum(is.na(sample_data[[LATITUDE_COLUMN]]) | 
                     is.na(sample_data[[LONGITUDE_COLUMN]]))

# Filter to valid coordinates
sample_valid <- sample_data %>%
  filter(!is.na(.data[[LATITUDE_COLUMN]]), 
         !is.na(.data[[LONGITUDE_COLUMN]]))

cat("Samples with missing coordinates:", missing_before, "\n")
cat("Valid samples:", nrow(sample_valid), "\n")

# Calculate coordinate ranges
lat_range <- range(sample_valid[[LATITUDE_COLUMN]])
lon_range <- range(sample_valid[[LONGITUDE_COLUMN]])

# Create summary
coord_summary <- data.frame(
  Statistic = c("Total Samples", "Valid Coordinates", "Missing Coordinates",
                "Min Latitude", "Max Latitude", "Latitude Range",
                "Min Longitude", "Max Longitude", "Longitude Range"),
  Value = c(
    nrow(sample_data),
    nrow(sample_valid),
    missing_before,
    round(lat_range[1], 4),
    round(lat_range[2], 4),
    round(diff(lat_range), 4),
    round(lon_range[1], 4),
    round(lon_range[2], 4),
    round(diff(lon_range), 4)
  )
)

kable(coord_summary, caption = "Coordinate Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Summary by Group (if applicable)

```{r group-summary}
if (GROUP_COLUMN %in% colnames(sample_valid)) {
  cat("\nSummary by", GROUP_COLUMN, ":\n")
  
  group_summary <- sample_valid %>%
    group_by(.data[[GROUP_COLUMN]]) %>%
    summarise(
      Count = n(),
      Mean_Lat = round(mean(.data[[LATITUDE_COLUMN]], na.rm = TRUE), 4),
      Mean_Lon = round(mean(.data[[LONGITUDE_COLUMN]], na.rm = TRUE), 4),
      .groups = 'drop'
    )
  
  kable(group_summary, 
        caption = paste("Summary by", GROUP_COLUMN)) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
} else {
  cat("\nNo grouping variable specified\n")
}
```

# Load Map Data

## Get Base Map Data

```{r get-maps}
cat("Loading map data...\n")

# Get world countries (medium resolution)
world <- ne_countries(scale = "medium", returnclass = "sf")

cat("Medium resolution world map loaded\n")

# Try to get high-resolution data
world_hires <- NULL
tryCatch({
  library(rnaturalearthhires)
  world_hires <- ne_countries(scale = "large", returnclass = "sf")
  cat("High-resolution map data loaded\n")
}, error = function(e) {
  world_hires <- world
  cat("High-resolution data not available, using medium resolution\n")
  cat("Install with: install.packages('rnaturalearthhires', repos='http://packages.ropensci.org', type='source')\n")
})

# Get coastline for detail
coastline <- ne_coastline(scale = "medium", returnclass = "sf")

cat("Map data ready\n")
```

## Calculate Map Extents

```{r calculate-extents}
# Calculate data extent
data_lon_min <- min(sample_valid[[LONGITUDE_COLUMN]])
data_lon_max <- max(sample_valid[[LONGITUDE_COLUMN]])
data_lat_min <- min(sample_valid[[LATITUDE_COLUMN]])
data_lat_max <- max(sample_valid[[LATITUDE_COLUMN]])

# Add buffer for regional map
regional_lon_min <- max(data_lon_min - REGIONAL_BUFFER, -180)
regional_lon_max <- min(data_lon_max + REGIONAL_BUFFER, 180)
regional_lat_min <- max(data_lat_min - REGIONAL_BUFFER, -90)
regional_lat_max <- min(data_lat_max + REGIONAL_BUFFER, 90)

cat("\nData extent:\n")
cat("  Longitude: [", round(data_lon_min, 2), ",", round(data_lon_max, 2), "]\n")
cat("  Latitude:  [", round(data_lat_min, 2), ",", round(data_lat_max, 2), "]\n")

cat("\nRegional map extent (with buffer):\n")
cat("  Longitude: [", round(regional_lon_min, 2), ",", round(regional_lon_max, 2), "]\n")
cat("  Latitude:  [", round(regional_lat_min, 2), ",", round(regional_lat_max, 2), "]\n")

# Store extents
extent_df <- data.frame(
  Map = c("Data", "Data", "Regional", "Regional"),
  Coordinate = c("Longitude", "Latitude", "Longitude", "Latitude"),
  Min = c(data_lon_min, data_lat_min, regional_lon_min, regional_lat_min),
  Max = c(data_lon_max, data_lat_max, regional_lon_max, regional_lat_max),
  Range = c(
    data_lon_max - data_lon_min,
    data_lat_max - data_lat_min,
    regional_lon_max - regional_lon_min,
    regional_lat_max - regional_lat_min
  )
)

kable(extent_df, digits = 4, caption = "Map Extents") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Create Maps

## Define Common Theme

```{r define-theme}
# Common theme for all maps
map_theme <- theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10),
    panel.grid.major = element_line(color = "gray80", linewidth = 0.2),
    panel.background = element_rect(fill = OCEAN_COLOR, color = NA),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  )
```

## World/Overview Map

```{r world-map, fig.width=12, fig.height=8}
cat("Creating world overview map...\n")

# Determine if we need world view or auto-zoom
if (!is.null(WORLD_XLIM) && !is.null(WORLD_YLIM)) {
  xlim_use <- WORLD_XLIM
  ylim_use <- WORLD_YLIM
  map_title <- "Sample Locations - World Overview"
} else {
  # Auto-zoom to continent/region
  xlim_use <- c(regional_lon_min - 20, regional_lon_max + 20)
  ylim_use <- c(regional_lat_min - 10, regional_lat_max + 10)
  map_title <- "Sample Locations - Overview"
}

# Create base plot
world_map <- ggplot() +
  geom_sf(data = world, fill = MAP_FILL_COLOR, color = "black", linewidth = 0.3) +
  coord_sf(xlim = xlim_use, ylim = ylim_use, expand = FALSE)

# Add points with or without grouping
if (GROUP_COLUMN %in% colnames(sample_valid)) {
  world_map <- world_map +
    geom_point(data = sample_valid, 
               aes(x = .data[[LONGITUDE_COLUMN]], 
                   y = .data[[LATITUDE_COLUMN]], 
                   color = .data[[GROUP_COLUMN]]), 
               size = POINT_SIZE, alpha = POINT_ALPHA) +
    scale_color_brewer(palette = COLOR_PALETTE, name = GROUP_COLUMN)
} else {
  world_map <- world_map +
    geom_point(data = sample_valid, 
               aes(x = .data[[LONGITUDE_COLUMN]], 
                   y = .data[[LATITUDE_COLUMN]]), 
               color = "red", size = POINT_SIZE, alpha = POINT_ALPHA)
}

# Add annotations and theme
world_map <- world_map +
  labs(title = map_title,
       x = "Longitude (°)", y = "Latitude (°)") +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true",
                        height = unit(1, "cm"), width = unit(1, "cm"),
                        style = north_arrow_fancy_orienteering) +
  # Add box showing regional zoom area
  geom_rect(aes(xmin = regional_lon_min, xmax = regional_lon_max, 
                ymin = regional_lat_min, ymax = regional_lat_max),
            fill = NA, color = "red", linewidth = 1, linetype = "dashed") +
  map_theme

print(world_map)
```

## Regional Zoom Map

```{r regional-map, fig.width=12, fig.height=8}
cat("Creating regional zoom map...\n")

# Create regional map with higher detail
regional_map <- ggplot() +
  geom_sf(data = world_hires, fill = MAP_FILL_COLOR, color = "black", linewidth = 0.4) +
  geom_sf(data = coastline, color = "black", linewidth = 0.2) +
  coord_sf(xlim = c(regional_lon_min, regional_lon_max), 
           ylim = c(regional_lat_min, regional_lat_max), 
           expand = FALSE)

# Add points with or without grouping
if (GROUP_COLUMN %in% colnames(sample_valid)) {
  regional_map <- regional_map +
    geom_point(data = sample_valid, 
               aes(x = .data[[LONGITUDE_COLUMN]], 
                   y = .data[[LATITUDE_COLUMN]], 
                   color = .data[[GROUP_COLUMN]]), 
               size = POINT_SIZE * 1.5, alpha = POINT_ALPHA) +
    scale_color_brewer(palette = COLOR_PALETTE, name = GROUP_COLUMN)
} else {
  regional_map <- regional_map +
    geom_point(data = sample_valid, 
               aes(x = .data[[LONGITUDE_COLUMN]], 
                   y = .data[[LATITUDE_COLUMN]]), 
               color = "red", size = POINT_SIZE * 1.5, alpha = POINT_ALPHA)
}

# Add annotations and theme
regional_map <- regional_map +
  labs(title = "Sample Locations - Regional Detail",
       x = "Longitude (°)", y = "Latitude (°)") +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true",
                        height = unit(1, "cm"), width = unit(1, "cm"),
                        style = north_arrow_fancy_orienteering) +
  map_theme

print(regional_map)
```

## Combined Map (Side by Side)

```{r combined-map, fig.width=16, fig.height=8}
cat("Creating combined map...\n")

# Combine maps side by side
combined_map <- world_map + regional_map +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Geographic Distribution of Samples",
    subtitle = paste("Overview and regional detail | N =", nrow(sample_valid), "samples"),
    theme = theme(
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5)
    )
  ) &
  theme(legend.position = "bottom")

print(combined_map)
```

## Map with Sample Labels (Regional Only)

```{r labeled-map, fig.width=12, fig.height=8}
cat("Creating map with sample labels...\n")

# Only add labels if not too many samples
if (nrow(sample_valid) <= 50) {
  
  labeled_map <- ggplot() +
    geom_sf(data = world_hires, fill = MAP_FILL_COLOR, color = "black", linewidth = 0.4) +
    geom_sf(data = coastline, color = "black", linewidth = 0.2) +
    coord_sf(xlim = c(regional_lon_min, regional_lon_max), 
             ylim = c(regional_lat_min, regional_lat_max), 
             expand = FALSE)
  
  # Add points and labels
  if (GROUP_COLUMN %in% colnames(sample_valid)) {
    labeled_map <- labeled_map +
      geom_point(data = sample_valid, 
                 aes(x = .data[[LONGITUDE_COLUMN]], 
                     y = .data[[LATITUDE_COLUMN]], 
                     color = .data[[GROUP_COLUMN]]), 
                 size = POINT_SIZE * 1.5, alpha = POINT_ALPHA) +
      scale_color_brewer(palette = COLOR_PALETTE, name = GROUP_COLUMN)
  } else {
    labeled_map <- labeled_map +
      geom_point(data = sample_valid, 
                 aes(x = .data[[LONGITUDE_COLUMN]], 
                     y = .data[[LATITUDE_COLUMN]]), 
                 color = "red", size = POINT_SIZE * 1.5, alpha = POINT_ALPHA)
  }
  
  # Add labels
  labeled_map <- labeled_map +
    geom_text(data = sample_valid,
              aes(x = .data[[LONGITUDE_COLUMN]], 
                  y = .data[[LATITUDE_COLUMN]], 
                  label = .data[[SAMPLE_ID_COLUMN]]),
              size = 2.5, hjust = -0.2, vjust = 0.5, check_overlap = TRUE) +
    labs(title = "Sample Locations with Labels",
         x = "Longitude (°)", y = "Latitude (°)") +
    annotation_scale(location = "bl", width_hint = 0.3) +
    annotation_north_arrow(location = "tr", which_north = "true",
                          height = unit(1, "cm"), width = unit(1, "cm"),
                          style = north_arrow_fancy_orienteering) +
    map_theme
  
  print(labeled_map)
  
} else {
  cat("Too many samples (", nrow(sample_valid), ") for labeling\n")
  cat("Skipping labeled map (recommended for ≤50 samples)\n")
}
```

# Interactive Map (Optional)

```{r interactive-map, eval=FALSE}
# Create interactive map using leaflet (requires leaflet package)
if (requireNamespace("leaflet", quietly = TRUE)) {
  library(leaflet)
  
  interactive_map <- leaflet(sample_valid) %>%
    addTiles() %>%
    addCircleMarkers(
      lng = ~get(LONGITUDE_COLUMN),
      lat = ~get(LATITUDE_COLUMN),
      popup = ~get(SAMPLE_ID_COLUMN),
      radius = 6,
      color = "red",
      fillOpacity = 0.7
    )
  
  # Display interactive map
  interactive_map
  
  # Save interactive map
  library(htmlwidgets)
  saveWidget(interactive_map, 
             file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_interactive.html")))
  cat("Interactive map saved\n")
} else {
  cat("Install leaflet package for interactive maps: install.packages('leaflet')\n")
}
```

# Save Maps

```{r save-maps}
cat("Saving maps to disk...\n")

# Save world overview map
ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_world_overview.png")),
       plot = world_map,
       width = 12, height = 8, dpi = SAVE_DPI, bg = "white")
cat("Saved: ", OUTPUT_PREFIX, "_world_overview.png\n")

# Save regional map
ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_regional.png")),
       plot = regional_map,
       width = 12, height = 8, dpi = SAVE_DPI, bg = "white")
cat("Saved: ", OUTPUT_PREFIX, "_regional.png\n")

# Save combined map
ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_combined.png")),
       plot = combined_map,
       width = 16, height = 8, dpi = SAVE_DPI, bg = "white")
cat("Saved: ", OUTPUT_PREFIX, "_combined.png\n")

# Save labeled map if created
if (exists("labeled_map")) {
  ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_labeled.png")),
         plot = labeled_map,
         width = 12, height = 8, dpi = SAVE_DPI, bg = "white")
  cat("Saved: ", OUTPUT_PREFIX, "_labeled.png\n")
}

# Save coordinate data
write.csv(sample_valid, 
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_coordinates.csv")),
          row.names = FALSE)
cat("Saved: ", OUTPUT_PREFIX, "_coordinates.csv\n")

# Save summary
write.csv(coord_summary,
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_summary.csv")),
          row.names = FALSE)
cat("Saved: ", OUTPUT_PREFIX, "_summary.csv\n")

cat("\nAll maps saved successfully!\n")
cat("Output directory:", OUTPUT_DIR, "\n")
```

# Summary Statistics

```{r summary-stats}
# Create comprehensive summary
cat("\n=== SAMPLE MAPPING SUMMARY ===\n")
cat("Total samples in file:", nrow(sample_data), "\n")
cat("Samples with valid coordinates:", nrow(sample_valid), "\n")
cat("Missing coordinates:", nrow(sample_data) - nrow(sample_valid), "\n")

cat("\nGeographic extent:\n")
cat("  Latitude range:", round(diff(lat_range), 2), "degrees\n")
cat("  Longitude range:", round(diff(lon_range), 2), "degrees\n")

cat("\nCenter point:\n")
cat("  Latitude:", round(mean(lat_range), 4), "\n")
cat("  Longitude:", round(mean(lon_range), 4), "\n")

if (GROUP_COLUMN %in% colnames(sample_valid)) {
  cat("\nSamples by", GROUP_COLUMN, ":\n")
  print(table(sample_valid[[GROUP_COLUMN]]))
}

cat("\n=== FILES CREATED ===\n")
cat("1. World overview map\n")
cat("2. Regional detail map\n")
cat("3. Combined side-by-side map\n")
if (exists("labeled_map")) {
  cat("4. Labeled map\n")
}
cat("5. Coordinate data CSV\n")
cat("6. Summary statistics CSV\n")
```

# Session Information

```{r session-info}
sessionInfo()
```

# References

1. **rnaturalearth**: South, A. (2017). rnaturalearth: World Map Data from Natural Earth. R package version 0.1.0.

2. **sf**: Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. *The R Journal*, 10(1), 439-446.

3. **ggspatial**: Dunnington, D. (2021). ggspatial: Spatial Data Framework for ggplot2. R package.

---

**Analysis completed:** `r Sys.time()`
