---
title: "Extract Bioclim Data from Geographic Coordinates"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This script extracts bioclim variables for a set of geographic coordinates and visualizes them on a world map.

**Requirements:**
- A CSV file with columns named "Latitude" and "Longitude" (case-sensitive)
- Optional: An ID column to identify each location

## Load Libraries

```{r}
library(terra)
library(geodata)
library(raster)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggplot2)
```

## Configuration

**Edit these parameters to customize for your data:**

```{r}
# Input CSV file (must contain "Latitude" and "Longitude" columns)
input_file <- "your_data.csv"

# Output file name for extracted bioclim data
output_file <- "bioclim_data_extracted.csv"

# Optional: Name of ID column in your CSV (set to NULL if none)
id_column <- NULL  # e.g., "SampleID", "Accession", etc.

# Bioclim resolution: 10, 5, 2.5, or 0.5 (in minutes of a degree)
# Lower resolution = faster download but less detailed
bioclim_resolution <- 2.5

# Directory to store downloaded bioclim rasters
bioclim_dir <- "bioclim_data/"
```

## Download Bioclim Data

Downloads WorldClim bioclim variables if not already present.

```{r}
# Create directory if it doesn't exist
if (!dir.exists(bioclim_dir)) {
  dir.create(bioclim_dir, recursive = TRUE)
}

# Download bioclim data (will skip if already downloaded)
bioclim_rasters <- worldclim_global(
  var = "bio",
  res = bioclim_resolution,
  path = bioclim_dir
)

# Preview first bioclim variable
plot(bioclim_rasters[[1]], main = "BIO1: Annual Mean Temperature")
```

## Read and Clean Coordinate Data

```{r}
# Read input data
obs_data <- read.csv(input_file, header = TRUE)

# Check required columns
if (!all(c("Latitude", "Longitude") %in% colnames(obs_data))) {
  stop("Input file must contain 'Latitude' and 'Longitude' columns (case-sensitive)")
}

cat("Original data dimensions:", nrow(obs_data), "rows x", ncol(obs_data), "columns\n")

# Remove rows with missing coordinates
obs_data <- obs_data[complete.cases(obs_data[, c("Longitude", "Latitude")]), ]
cat("After removing NA coordinates:", nrow(obs_data), "rows\n")

# Remove duplicate coordinates
dups <- duplicated(obs_data[, c("Longitude", "Latitude")])
cat("Duplicate coordinates found:", sum(dups), "\n")
obs_data <- obs_data[!dups, ]
cat("After removing duplicates:", nrow(obs_data), "rows\n\n")

# Display coordinate summary
cat("Coordinate ranges:\n")
cat("  Latitude: ", min(obs_data$Latitude), "to", max(obs_data$Latitude), "\n")
cat("  Longitude:", min(obs_data$Longitude), "to", max(obs_data$Longitude), "\n\n")

head(obs_data[, c("Longitude", "Latitude")])
```

## Visualize Points on World Map

```{r, fig.width=12, fig.height=6}
# Get world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Create plot
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray40", linewidth = 0.3) +
  geom_point(
    data = obs_data,
    aes(x = Longitude, y = Latitude),
    color = "red",
    size = 2,
    alpha = 0.6
  ) +
  coord_sf(expand = FALSE) +
  theme_minimal() +
  labs(
    title = "Geographic Distribution of Sample Locations",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme(
    panel.background = element_rect(fill = "aliceblue"),
    panel.grid = element_line(color = "gray80", linewidth = 0.2)
  )
```

## Zoomed View (Optional)

Automatically zooms to the extent of your data points.

```{r, fig.width=10, fig.height=8}
# Calculate extent with buffer
lon_range <- range(obs_data$Longitude)
lat_range <- range(obs_data$Latitude)
lon_buffer <- diff(lon_range) * 0.1
lat_buffer <- diff(lat_range) * 0.1

ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray40", linewidth = 0.3) +
  geom_point(
    data = obs_data,
    aes(x = Longitude, y = Latitude),
    color = "darkred",
    size = 3,
    alpha = 0.7
  ) +
  coord_sf(
    xlim = c(lon_range[1] - lon_buffer, lon_range[2] + lon_buffer),
    ylim = c(lat_range[1] - lat_buffer, lat_range[2] + lat_buffer),
    expand = FALSE
  ) +
  theme_minimal() +
  labs(
    title = "Sample Locations (Zoomed)",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme(
    panel.background = element_rect(fill = "aliceblue"),
    panel.grid = element_line(color = "gray80", linewidth = 0.2)
  )
```

## Extract Bioclim Values

```{r}
# Convert coordinates to spatial points
coords_only <- obs_data[, c("Longitude", "Latitude")]
points_sv <- vect(coords_only, geom = c("Longitude", "Latitude"), crs = crs(bioclim_rasters))

# Extract bioclim values at each point
bioclim_values <- extract(bioclim_rasters, points_sv)

# Prepare final dataset
if (!is.null(id_column) && id_column %in% colnames(obs_data)) {
  # Include ID column if specified and exists
  final_data <- cbind(
    obs_data[, c(id_column, "Longitude", "Latitude")],
    bioclim_values[, -1]  # Remove terra's internal ID column
  )
} else {
  # Just coordinates
  final_data <- cbind(
    coords_only,
    bioclim_values[, -1]
  )
}

# Display summary
cat("Extracted bioclim data dimensions:", nrow(final_data), "rows x", ncol(final_data), "columns\n")
cat("\nBioclim variables extracted:\n")
cat(paste(colnames(final_data)[-(1:ncol(coords_only))], collapse = ", "), "\n\n")

head(final_data)
```

## Save Results

```{r}
# Write to CSV
write.csv(final_data, output_file, row.names = FALSE)

cat("Results saved to:", output_file, "\n")
cat("Total locations processed:", nrow(final_data), "\n")
```

## Bioclim Variable Descriptions

The 19 bioclim variables are:

- **BIO1**: Annual Mean Temperature
- **BIO2**: Mean Diurnal Range
- **BIO3**: Isothermality
- **BIO4**: Temperature Seasonality
- **BIO5**: Max Temperature of Warmest Month
- **BIO6**: Min Temperature of Coldest Month
- **BIO7**: Temperature Annual Range
- **BIO8**: Mean Temperature of Wettest Quarter
- **BIO9**: Mean Temperature of Driest Quarter
- **BIO10**: Mean Temperature of Warmest Quarter
- **BIO11**: Mean Temperature of Coldest Quarter
- **BIO12**: Annual Precipitation
- **BIO13**: Precipitation of Wettest Month
- **BIO14**: Precipitation of Driest Month
- **BIO15**: Precipitation Seasonality
- **BIO16**: Precipitation of Wettest Quarter
- **BIO17**: Precipitation of Driest Quarter
- **BIO18**: Precipitation of Warmest Quarter
- **BIO19**: Precipitation of Coldest Quarter

---

**Session Info**

```{r}
sessionInfo()
```
