---
title: "Population Structure Analysis via PCA"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This document performs Principal Component Analysis (PCA) on genomic data from a VCF file to visualize population structure.

## Configuration

```{r config}
# ---------- USER CONFIGURATION ----------
# Modify these variables for your dataset

# Input VCF file name
vcf_file <- "your_data.vcf.gz"

# Output GDS file name (will be created)
gds_file <- "your_data.gds"

# Metadata file (optional - set to NULL if not available)
# Should contain columns: SampleID and a grouping variable
metadata_file <- "metadata.csv"

# Name of the grouping column in metadata (e.g., "Population", "Species", "Cluster")
group_column <- "Population"

# Plot title
plot_title <- "Population Structure"

# Output file for PCA results
output_file <- "PCA_results.csv"
```

## Load Required Libraries

```{r libraries, message=FALSE, warning=FALSE}
# ---------- 1. Install and Load Required Packages ----------
library(SNPRelate)
library(ggplot2)
```

## Convert VCF to GDS Format

```{r vcf_to_gds}
# ---------- 2. Convert VCF to GDS ----------
# This step only needs to be run once per VCF file
if (!file.exists(gds_file)) {
  snpgdsVCF2GDS(vcf_file, gds_file, method = "biallelic.only")
  cat("GDS file created successfully\n")
} else {
  cat("GDS file already exists, skipping conversion\n")
}
```

## Open GDS File

```{r open_gds}
# ---------- 3. Open GDS File ----------
genofile <- snpgdsOpen(gds_file)
```

## LD Pruning

```{r ld_pruning}
# ---------- 4. LD Pruning ----------
# Remove SNPs in linkage disequilibrium to reduce redundancy
pruned_snp <- snpgdsLDpruning(genofile, autosome.only = FALSE)
snpset <- unlist(pruned_snp, use.names = FALSE)

cat("Number of SNPs after LD pruning:", length(snpset), "\n")
```

## Perform PCA

```{r run_pca}
# ---------- 5. Perform PCA ----------
pca <- snpgdsPCA(genofile, snp.id = snpset, autosome.only = FALSE)

cat("PCA completed successfully\n")
cat("Number of principal components:", ncol(pca$eigenvect), "\n")
```

## Extract PCA Results

```{r extract_pca}
# ---------- 6. Extract PCA Results ----------
# Calculate variance explained by each PC
pc_percent <- pca$varprop * 100

# Create data frame with PCA results
pca_df <- data.frame(
  SampleID = pca$sample.id,
  PC1 = pca$eigenvect[, 1],
  PC2 = pca$eigenvect[, 2],
  PC3 = pca$eigenvect[, 3],
  PC4 = pca$eigenvect[, 4],
  PC5 = pca$eigenvect[, 5]
)

# Display variance explained
cat("\nVariance explained by first 5 PCs:\n")
for (i in 1:5) {
  cat(sprintf("PC%d: %.2f%%\n", i, pc_percent[i]))
}
```

## Add Metadata (Optional)

```{r add_metadata}
# ---------- 7. Add Population/Group Labels ----------
if (!is.null(metadata_file) && file.exists(metadata_file)) {
  # Read metadata
  metadata <- read.csv(metadata_file)
  
  # Check if required columns exist
  if (!("SampleID" %in% colnames(metadata))) {
    stop("Metadata file must contain a 'SampleID' column")
  }
  
  if (!(group_column %in% colnames(metadata))) {
    stop(paste0("Metadata file must contain a '", group_column, "' column"))
  }
  
  # Merge with PCA data
  pca_df <- merge(pca_df, metadata[, c("SampleID", group_column)], 
                  by = "SampleID", all.x = TRUE)
  
  # Check for sample ID mismatches
  n_matched <- sum(!is.na(pca_df[[group_column]]))
  cat(sprintf("\n%d of %d samples matched with metadata\n", 
              n_matched, nrow(pca_df)))
  
} else {
  # If no metadata, create a default group
  cat("\nNo metadata file provided - all samples assigned to single group\n")
  pca_df$Group <- "All Samples"
  group_column <- "Group"
}

head(pca_df)
```

## Visualize PCA

### PC1 vs PC2

```{r plot_pc12, fig.width=8, fig.height=6}
# ---------- 8. Plot PCA (PC1 vs PC2) ----------
ggplot(pca_df, aes(x = PC1, y = PC2, color = .data[[group_column]])) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(type = "norm", level = 0.95, linetype = 2, linewidth = 1) +
  labs(
    title = plot_title,
    x = paste0("PC1 (", round(pc_percent[1], 2), "%)"),
    y = paste0("PC2 (", round(pc_percent[2], 2), "%)"),
    color = group_column
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )
```

### PC1 vs PC3

```{r plot_pc13, fig.width=8, fig.height=6}
# ---------- 9. Plot PCA (PC1 vs PC3) ----------
ggplot(pca_df, aes(x = PC1, y = PC3, color = .data[[group_column]])) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(type = "norm", level = 0.95, linetype = 2, linewidth = 1) +
  labs(
    title = plot_title,
    x = paste0("PC1 (", round(pc_percent[1], 2), "%)"),
    y = paste0("PC3 (", round(pc_percent[3], 2), "%)"),
    color = group_column
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )
```

### PC2 vs PC3

```{r plot_pc23, fig.width=8, fig.height=6}
# ---------- 10. Plot PCA (PC2 vs PC3) ----------
ggplot(pca_df, aes(x = PC2, y = PC3, color = .data[[group_column]])) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(type = "norm", level = 0.95, linetype = 2, linewidth = 1) +
  labs(
    title = plot_title,
    x = paste0("PC2 (", round(pc_percent[2], 2), "%)"),
    y = paste0("PC3 (", round(pc_percent[3], 2), "%)"),
    color = group_column
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )
```

## Save Results

```{r save_results}
# ---------- 11. Save PCA Results ----------
write.csv(pca_df, output_file, row.names = FALSE)
cat("\nPCA results saved to:", output_file, "\n")
```

## Close GDS File

```{r close_gds}
# ---------- 12. Close the GDS File ----------
snpgdsClose(genofile)
cat("GDS file closed\n")
```

## Session Information

```{r session_info}
sessionInfo()
```
