---
title: "Spatial Population Structure Analysis with TESS3"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = 'center'
)
```

# Introduction

## What is TESS3?

**TESS3** (Terrain-based Estimation of Spatial Structure) is a spatial population genetics method that estimates ancestry coefficients while accounting for geographic information.

**Key features:**

- **Spatially explicit**: Uses geographic coordinates directly
- **No prior population assignment**: Infers structure from data
- **Continuous variation**: Models smooth transitions across space
- **Visualization**: Generates spatial interpolation maps
- **K selection**: Cross-validation to determine optimal number of clusters

**Advantages over STRUCTURE/ADMIXTURE:**

- Incorporates spatial information directly
- Better for continuously distributed populations
- Produces beautiful spatial maps
- Faster computation
- No need to define discrete populations

## Applications

- **Landscape genetics**: Understand gene flow patterns
- **Conservation**: Identify management units
- **Biogeography**: Map genetic clines
- **Invasion biology**: Track introduction routes
- **Agriculture**: Characterize germplasm structure
- **Climate adaptation**: Identify locally adapted populations

# Setup

## Load Required Libraries

```{r load-libraries}
# Load required packages
library(vcfR)          # VCF file handling
library(tess3r)        # TESS3 analysis
library(maps)          # Base R maps
library(rworldmap)     # World map data
library(ggplot2)       # Plotting
library(dplyr)         # Data manipulation
library(tidyr)         # Data reshaping
library(patchwork)     # Combine plots
library(knitr)         # Table formatting
library(kableExtra)    # Enhanced tables

cat("All required packages loaded successfully\n")
```

## Set File Paths and Parameters

```{r set-parameters}
# ==================== INPUT FILES ====================
VCF_FILE <- "your_genotypes.vcf"                    # Path to VCF file
COORD_CSV <- "your_coordinates.csv"                 # Path to coordinate CSV

# ==================== COORDINATE COLUMNS ====================
SAMPLE_ID_COLUMN <- "SampleID"                      # Sample ID column name
LATITUDE_COLUMN <- "Latitude"                       # Latitude column name
LONGITUDE_COLUMN <- "Longitude"                     # Longitude column name

# ==================== TESS3 PARAMETERS ====================
K_VALUES <- 1:7                                     # Range of K values to test
PLOIDY <- 2                                         # Ploidy level (2 for diploid)
N_CORES <- 4                                        # Number of CPU cores to use

# ==================== VISUALIZATION PARAMETERS ====================
# Geographic extent for maps (adjust based on your data)
MAP_XLIM <- c(-180, 180)                            # Longitude range
MAP_YLIM <- c(-90, 90)                              # Latitude range

# ==================== OUTPUT OPTIONS ====================
OUTPUT_PREFIX <- "tess3"                            # Prefix for output files
OUTPUT_DIR <- "tess3_results"                       # Output directory

# Create output directory
if (!dir.exists(OUTPUT_DIR)) {
  dir.create(OUTPUT_DIR, recursive = TRUE)
}

# Display parameters
param_table <- data.frame(
  Parameter = c("VCF File", "Coordinate CSV", "K Values", "Ploidy", "CPU Cores"),
  Value = c(VCF_FILE, COORD_CSV, paste(range(K_VALUES), collapse = " to "), 
            PLOIDY, N_CORES)
)

kable(param_table, caption = "Analysis Parameters") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Data Import and Preparation

## Read VCF File

```{r read-vcf}
cat("Reading VCF file:", VCF_FILE, "\n")

# Check if file exists
if (!file.exists(VCF_FILE)) {
  stop("VCF file not found: ", VCF_FILE)
}

# Read VCF
vcf <- read.vcfR(VCF_FILE)

cat("VCF loaded successfully\n")
cat("  SNPs:", nrow(vcf@fix), "\n")
cat("  Samples:", ncol(vcf@gt) - 1, "\n")

# Extract genotypes as numeric matrix (0, 1, 2 for diploid)
gt_numeric <- extract.gt(vcf, element = "GT", as.numeric = TRUE)

# Transpose: TESS3 requires samples as rows, SNPs as columns
genotype <- t(gt_numeric)

cat("  Genotype matrix dimensions (samples x SNPs):", dim(genotype), "\n")

# Get sample names
sample_names <- rownames(genotype)

# Create VCF summary
vcf_summary <- data.frame(
  Metric = c("Total SNPs", "Total Samples", "Missing Genotypes"),
  Value = c(
    ncol(genotype),
    nrow(genotype),
    paste0(round(100 * sum(is.na(genotype)) / length(genotype), 2), "%")
  )
)

kable(vcf_summary, caption = "VCF Data Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Read Coordinate Data

```{r read-coordinates}
cat("Reading coordinate data:", COORD_CSV, "\n")

# Check if file exists
if (!file.exists(COORD_CSV)) {
  stop("Coordinate CSV file not found: ", COORD_CSV)
}

# Read CSV
coordinates <- read.csv(COORD_CSV, header = TRUE, stringsAsFactors = FALSE)

cat("Coordinate data loaded\n")
cat("  Total samples:", nrow(coordinates), "\n")

# Display first few rows
kable(head(coordinates), caption = "Coordinate Data Preview") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%")

# Check required columns exist
required_cols <- c(SAMPLE_ID_COLUMN, LATITUDE_COLUMN, LONGITUDE_COLUMN)
missing_cols <- setdiff(required_cols, colnames(coordinates))

if (length(missing_cols) > 0) {
  stop("Missing required columns in coordinate CSV: ", 
       paste(missing_cols, collapse = ", "))
}
```

## Match Samples Between VCF and Coordinates

```{r match-samples}
cat("Matching samples between VCF and coordinate data...\n")

# Get sample IDs from both datasets
vcf_samples <- sample_names
coord_samples <- coordinates[[SAMPLE_ID_COLUMN]]

cat("VCF samples:", length(vcf_samples), "\n")
cat("Coordinate samples:", length(coord_samples), "\n")

# Find common samples
common_samples <- intersect(vcf_samples, coord_samples)

cat("Common samples:", length(common_samples), "\n")

if (length(common_samples) == 0) {
  stop("ERROR: No matching samples found!\n",
       "  VCF samples (first 5): ", paste(head(vcf_samples, 5), collapse = ", "), "\n",
       "  Coordinate samples (first 5): ", paste(head(coord_samples, 5), collapse = ", "))
}

# Create matching summary
match_summary <- data.frame(
  Category = c("VCF samples", "Coordinate samples", "Matched samples", 
               "VCF-only", "Coordinates-only"),
  Count = c(
    length(vcf_samples),
    length(coord_samples),
    length(common_samples),
    length(vcf_samples) - length(common_samples),
    length(coord_samples) - length(common_samples)
  )
)

kable(match_summary, caption = "Sample Matching Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Subset both datasets to common samples
genotype_matched <- genotype[rownames(genotype) %in% common_samples, ]
coordinates_matched <- coordinates[coordinates[[SAMPLE_ID_COLUMN]] %in% common_samples, ]

# Ensure same order
coordinates_matched <- coordinates_matched[match(rownames(genotype_matched), 
                                                 coordinates_matched[[SAMPLE_ID_COLUMN]]), ]

# Verify alignment
if (!all(rownames(genotype_matched) == coordinates_matched[[SAMPLE_ID_COLUMN]])) {
  stop("ERROR: Sample alignment verification failed!")
}

cat("Sample alignment verified\n")
```

## Remove Samples with Missing Coordinates

```{r remove-missing-coords}
cat("Checking for missing coordinates...\n")

# Identify samples with missing lat/long
coord_cols <- c(LATITUDE_COLUMN, LONGITUDE_COLUMN)
missing_coords <- which(!complete.cases(coordinates_matched[, coord_cols]))

if (length(missing_coords) > 0) {
  cat("Removing", length(missing_coords), "samples with missing coordinates\n")
  cat("Sample indices:", paste(missing_coords, collapse = ", "), "\n")
  
  genotype_final <- genotype_matched[-missing_coords, ]
  coordinates_final <- coordinates_matched[-missing_coords, ]
} else {
  cat("No missing coordinates found\n")
  genotype_final <- genotype_matched
  coordinates_final <- coordinates_matched
}

cat("Final dataset:", nrow(genotype_final), "samples\n")
```

## Impute Missing Genotypes

```{r impute-genotypes}
cat("Imputing missing genotypes...\n")

# Count missing before imputation
missing_before <- sum(is.na(genotype_final))
missing_pct_before <- 100 * missing_before / length(genotype_final)

# Impute missing genotypes with column (SNP) means
genotype_final <- apply(genotype_final, 2, function(x) {
  ifelse(is.na(x), mean(x, na.rm = TRUE), x)
})

# Count missing after imputation
missing_after <- sum(is.na(genotype_final))

cat("Missing genotypes before imputation:", missing_before, 
    paste0("(", round(missing_pct_before, 2), "%)"), "\n")
cat("Missing genotypes after imputation:", missing_after, "\n")
```

## Prepare Coordinate Matrix

```{r prepare-coords}
# Extract coordinate matrix for TESS3
coord_matrix <- as.matrix(coordinates_final[, c(LONGITUDE_COLUMN, LATITUDE_COLUMN)])

# Ensure numeric
coord_matrix <- apply(coord_matrix, 2, as.numeric)

cat("Final dataset dimensions:\n")
cat("  Samples:", nrow(genotype_final), "\n")
cat("  SNPs:", ncol(genotype_final), "\n")
cat("  Coordinates:", nrow(coord_matrix), "x", ncol(coord_matrix), "\n")

# Display coordinate summary
coord_summary <- data.frame(
  Statistic = c("Min Longitude", "Max Longitude", "Min Latitude", "Max Latitude",
                "Longitude Range", "Latitude Range"),
  Value = c(
    round(min(coord_matrix[, 1]), 4),
    round(max(coord_matrix[, 1]), 4),
    round(min(coord_matrix[, 2]), 4),
    round(max(coord_matrix[, 2]), 4),
    round(max(coord_matrix[, 1]) - min(coord_matrix[, 1]), 4),
    round(max(coord_matrix[, 2]) - min(coord_matrix[, 2]), 4)
  )
)

kable(coord_summary, caption = "Coordinate Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Visualize Sampling Locations

```{r plot-sampling-locations, fig.width=10, fig.height=6}
# Create map of sampling locations
plot(coord_matrix, pch = 19, cex = 0.8, col = "red",
     xlab = "Longitude (°)", ylab = "Latitude (°)",
     main = "Sampling Locations")
maps::map(add = TRUE, interior = TRUE, col = "gray70")
grid()
```

# TESS3 Analysis

## Run TESS3 for Multiple K Values

```{r run-tess3}
cat("Running TESS3 analysis for K =", paste(K_VALUES, collapse = ", "), "...\n")
cat("This may take several minutes...\n")

# Start timer
start_time <- Sys.time()

# Run TESS3
tess3_obj <- tess3(
  X = genotype_final, 
  coord = coord_matrix, 
  K = K_VALUES,
  method = "projected.ls", 
  ploidy = PLOIDY, 
  openMP.core.num = N_CORES
)

# End timer
end_time <- Sys.time()
elapsed <- as.numeric(difftime(end_time, start_time, units = "mins"))

cat("TESS3 analysis complete!\n")
cat("Elapsed time:", round(elapsed, 2), "minutes\n")
```

## Extract Cross-Validation Scores

```{r extract-cv}
cat("Extracting cross-validation scores...\n")

# Extract CV scores
if (!is.null(tess3_obj$crossvalid)) {
  cv_scores <- tess3_obj$crossvalid
} else if (!is.null(tess3_obj$rmse)) {
  cv_scores <- tess3_obj$rmse
} else {
  # Manually extract from each K
  cv_scores <- sapply(K_VALUES, function(k) {
    tess3_obj[[paste0("K", k)]]$crossvalid
  })
}

# Create CV results table
cv_table <- data.frame(
  K = K_VALUES,
  CV_Score = cv_scores
)

kable(cv_table, digits = 4, caption = "Cross-Validation Scores") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Determine optimal K (minimum CV score)
optimal_K <- K_VALUES[which.min(cv_scores)]
cat("\nOptimal K (minimum CV):", optimal_K, "\n")
```

## Plot Cross-Validation Scores

```{r plot-cv, fig.width=8, fig.height=6}
# Plot CV scores
plot(K_VALUES, cv_scores, 
     type = "b", pch = 19, col = "blue", lwd = 2,
     xlab = "Number of Ancestral Populations (K)",
     ylab = "Cross-Validation Score",
     main = "TESS3 Cross-Validation Scores")

# Mark optimal K
points(optimal_K, cv_scores[which.min(cv_scores)], 
       pch = 19, col = "red", cex = 2)
text(optimal_K, cv_scores[which.min(cv_scores)], 
     paste("Optimal K =", optimal_K), 
     pos = 3, col = "red", font = 2)
grid()
```

# Visualization: Ancestry Plots

## STRUCTURE-like Bar Plots

```{r structure-barplots, fig.width=14, fig.height=18}
# Define color palette
my.colors <- c("tomato", "orange", "purple", "olivedrab", "gold", 
               "lightblue", "grey", "pink", "brown")
my.palette <- CreatePalette(my.colors, 9)

# Set up multi-panel plot
par(mfrow = c(3, 2), mar = c(4, 4, 3, 1))

# Plot for each K value (skip K=1)
for (K in K_VALUES[K_VALUES > 1]) {
  # Get Q matrix
  q.matrix <- qmatrix(tess3_obj, K = K)
  
  # Create barplot
  barplot(q.matrix, 
          border = NA, 
          space = 0, 
          xlab = "Individuals", 
          ylab = "Ancestry Proportions", 
          main = paste("Ancestry Coefficients (K =", K, ")"),
          col = my.palette[1:K],
          las = 1)
}

# Reset plotting parameters
par(mfrow = c(1, 1))
```

## Individual Barplots (High Resolution)

```{r individual-barplots, eval=FALSE}
# Save individual high-resolution plots for each K
for (K in K_VALUES[K_VALUES > 1]) {
  
  filename <- file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_barplot_K", K, ".png"))
  
  png(filename, width = 12, height = 6, units = "in", res = 300)
  
  q.matrix <- qmatrix(tess3_obj, K = K)
  
  barplot(q.matrix, 
          border = NA, 
          space = 0, 
          xlab = "Individuals", 
          ylab = "Ancestry Proportions", 
          main = paste("Ancestry Coefficients (K =", K, ")"),
          col = my.palette[1:K],
          las = 1,
          cex.main = 1.5,
          cex.lab = 1.3,
          cex.axis = 1.2)
  
  dev.off()
  
  cat("Saved:", filename, "\n")
}
```

# Visualization: Spatial Maps

## Base R Spatial Interpolation Maps

```{r spatial-maps-base, fig.width=12, fig.height=16}
# Set up multi-panel plot
par(mfrow = c(3, 2), mar = c(4, 4, 3, 2))

# Plot for each K value (skip K=1)
for (K in K_VALUES[K_VALUES > 1]) {
  q.matrix <- qmatrix(tess3_obj, K = K)
  
  plot(q.matrix, coord_matrix, 
       method = "map.max", 
       interpol = FieldsKrigModel(10),
       main = paste("Spatial Ancestry (K =", K, ")"),
       xlab = "Longitude", 
       ylab = "Latitude",
       resolution = c(300, 300), 
       cex = 0.4,
       col.palette = my.palette)
}

# Reset plotting parameters
par(mfrow = c(1, 1))
```

## ggplot2 Spatial Maps

```{r spatial-maps-ggplot, fig.width=14, fig.height=18}
# Get world map polygon
map.polygon <- getMap(resolution = "low")

# Create list to store plots
plot_list <- list()

# Determine map extent based on data
lon_range <- range(coord_matrix[, 1])
lat_range <- range(coord_matrix[, 2])

# Add buffer
lon_buffer <- diff(lon_range) * 0.2
lat_buffer <- diff(lat_range) * 0.2

xlim_map <- c(lon_range[1] - lon_buffer, lon_range[2] + lon_buffer)
ylim_map <- c(lat_range[1] - lat_buffer, lat_range[2] + lat_buffer)

# Create plots for each K
for (K in K_VALUES[K_VALUES > 1]) {
  # Extract Q-matrix
  q.matrix <- qmatrix(tess3_obj, K = K)
  
  # Create map using ggtess3Q
  pl <- ggtess3Q(q.matrix, coord_matrix, map.polygon = map.polygon)
  
  # Customize plot
  p <- pl +
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group)) +
    xlim(xlim_map) +
    ylim(ylim_map) +
    coord_equal() +
    geom_point(data = data.frame(
                 Longitude = coord_matrix[, 1], 
                 Latitude = coord_matrix[, 2]),
               aes(x = Longitude, y = Latitude),
               size = 1, color = "black") +
    xlab("Longitude") +
    ylab("Latitude") +
    theme_bw() +
    ggtitle(paste("Spatial Ancestry (K =", K, ")")) +
    theme(
      plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
      axis.title = element_text(size = 11),
      axis.text = element_text(size = 10),
      legend.title = element_text(size = 10),
      legend.text = element_text(size = 9)
    )
  
  # Store plot
  plot_list[[paste0("K", K)]] <- p
}

# Combine all plots
combined_plot <- wrap_plots(plot_list, ncol = 2)
print(combined_plot)
```

# Extract and Save Q-Matrices

```{r save-qmatrices}
cat("Saving Q-matrices...\n")

# Save Q-matrix for each K
for (K in K_VALUES[K_VALUES > 1]) {
  q.matrix <- qmatrix(tess3_obj, K = K)
  
  # Create data frame with sample names
  q_df <- data.frame(
    Sample = rownames(genotype_final),
    q.matrix
  )
  colnames(q_df) <- c("Sample", paste0("Cluster_", 1:K))
  
  # Add coordinates
  q_df$Longitude <- coord_matrix[, 1]
  q_df$Latitude <- coord_matrix[, 2]
  
  # Save to CSV
  filename <- file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_qmatrix_K", K, ".csv"))
  write.csv(q_df, filename, row.names = FALSE)
  
  cat("Saved Q-matrix for K =", K, "\n")
}
```

# Save Results

```{r save-results}
cat("Saving results to disk...\n")

# Save combined barplot
png(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_barplots_combined.png")), 
    width = 14, height = 18, units = "in", res = 300)

par(mfrow = c(3, 2), mar = c(4, 4, 3, 1))

for (K in K_VALUES[K_VALUES > 1]) {
  q.matrix <- qmatrix(tess3_obj, K = K)
  
  barplot(q.matrix, 
          border = NA, 
          space = 0, 
          xlab = "Individuals", 
          ylab = "Ancestry Proportions", 
          main = paste("Ancestry Coefficients (K =", K, ")"),
          col = my.palette[1:K],
          las = 1)
}

par(mfrow = c(1, 1))
dev.off()

cat("Saved combined barplot\n")

# Save combined spatial maps (base R)
png(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_spatial_maps_baseR.png")), 
    width = 12, height = 16, units = "in", res = 300)

par(mfrow = c(3, 2), mar = c(4, 4, 3, 2))

for (K in K_VALUES[K_VALUES > 1]) {
  q.matrix <- qmatrix(tess3_obj, K = K)
  
  plot(q.matrix, coord_matrix, 
       method = "map.max", 
       interpol = FieldsKrigModel(10),
       main = paste("Spatial Ancestry (K =", K, ")"),
       xlab = "Longitude", ylab = "Latitude",
       resolution = c(300, 300), cex = 0.4,
       col.palette = my.palette)
}

par(mfrow = c(1, 1))
dev.off()

cat("Saved spatial maps (base R)\n")

# Save combined spatial maps (ggplot)
ggsave(file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_spatial_maps_ggplot.png")), 
       combined_plot, 
       width = 14, height = 18, dpi = 300)

cat("Saved spatial maps (ggplot)\n")

# Save cross-validation results
write.csv(cv_table, 
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_cross_validation.csv")), 
          row.names = FALSE)

cat("Saved cross-validation results\n")

# Save summary
summary_data <- data.frame(
  Metric = c("Total Samples", "Total SNPs", "Optimal K", "Min CV Score"),
  Value = c(
    nrow(genotype_final),
    ncol(genotype_final),
    optimal_K,
    round(min(cv_scores), 4)
  )
)

write.csv(summary_data, 
          file.path(OUTPUT_DIR, paste0(OUTPUT_PREFIX, "_summary.csv")), 
          row.names = FALSE)

cat("Saved analysis summary\n")

cat("\nAll results saved successfully!\n")
cat("Output directory:", OUTPUT_DIR, "\n")
```

# Interpretation Guide

## Understanding TESS3 Results

### Cross-Validation Scores

**Lower CV score = better model fit**

- **Elbow method**: Look for where CV score plateaus
- **Minimum CV**: Typically indicates optimal K
- **Biological meaning**: Consider biological interpretation alongside statistics

### Ancestry Coefficients (Q-matrix)

**Each row = one individual**
**Each column = one ancestry cluster**
**Values sum to 1** across clusters for each individual

**Interpretation:**

- **Pure ancestry** (e.g., 0.9, 0.05, 0.05): Individual mainly from one cluster
- **Admixed** (e.g., 0.4, 0.3, 0.3): Individual has mixed ancestry
- **Recent admixture**: Sharp transitions in geographic space
- **Ancient admixture**: Broad clines

### Spatial Maps

**Color patterns indicate:**

- **Distinct regions**: Sharp color boundaries = population differentiation
- **Gradual transitions**: Color clines = isolation by distance
- **Patchwork**: Complex admixture patterns
- **Geographic barriers**: Correspond to mountains, rivers, etc.

### Choosing K

**Consider multiple lines of evidence:**

1. **Cross-validation score**: Minimum value
2. **Biological knowledge**: Known populations/subspecies
3. **Geographic structure**: Match to landscape features
4. **Stability**: Consistent results across runs
5. **Interpretation**: Meaningful biological clusters

**Common patterns:**

- **K = 2**: Often captures major geographic split
- **K = 3-5**: Finer-scale regional structure
- **K > 5**: May reveal fine-scale or family structure

## Biological Interpretation

### Population Structure

**Strong structure (distinct clusters):**
- Limited gene flow between regions
- Geographic barriers to dispersal
- Different selective environments
- Long-term isolation

**Weak structure (gradual clines):**
- Continuous gene flow
- Isolation by distance
- No major barriers
- Recent colonization

### Gene Flow Patterns

**Look for:**

- **Source populations**: High ancestry in one cluster
- **Admixture zones**: High mixing between clusters
- **Migration routes**: Gradual changes along corridors
- **Barriers**: Sharp transitions at boundaries

### Management Implications

**For conservation:**

- **Genetic units**: Distinct clusters may warrant separate management
- **Connectivity**: Admixture zones indicate gene flow corridors
- **Priorities**: Unique clusters (high differentiation) are priorities
- **Translocation**: Avoid mixing divergent genetic units

**For breeding:**

- **Diversity**: Sample across all major clusters
- **Heterosis**: Cross individuals from different clusters
- **Local adaptation**: Consider cluster-environment associations

## Next Steps

1. **Validate with other methods**
   - Compare with STRUCTURE/ADMIXTURE
   - Run PCA for comparison
   - Calculate pairwise FST

2. **Environmental associations**
   - Test cluster-environment correlations
   - Run RDA or LFMM
   - Identify adaptive loci

3. **Demographic history**
   - Test for isolation by distance
   - Estimate migration rates
   - Model demographic scenarios

4. **Functional analysis**
   - Annotate differentiated loci
   - Test for selection signatures
   - Link to phenotypes

# Session Information

```{r session-info}
sessionInfo()
```

# References

1. Caye, K., Deist, T. M., Martins, H., Michel, O., & François, O. (2016). TESS3: Fast inference of spatial population structure and genome scans for selection. *Molecular Ecology Resources*, 16(2), 540-548.

2. Frichot, E., Mathieu, F., Trouillon, T., Bouchard, G., & François, O. (2014). Fast and efficient estimation of individual ancestry coefficients. *Genetics*, 196(4), 973-983.

3. Guillot, G., Mortier, F., & Estoup, A. (2005). GENELAND: a computer package for landscape genetics. *Molecular Ecology Notes*, 5(3), 712-715.

---

**Analysis completed:** `r Sys.time()`
