---
title: "Linkage Disequilibrium (LD) Decay Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)
```

## Overview

This R Markdown document performs a comprehensive linkage disequilibrium (LD) decay analysis from VCF files. The analysis calculates pairwise LD (r²) between SNPs and models the decay of LD with physical distance.

**What this analysis does:**

- Calculates pairwise r² values between SNPs within chromosomes
- Bins SNP pairs by distance and calculates mean r²
- Fits exponential decay models to estimate LD decay
- Generates genome-wide and per-chromosome visualizations
- Calculates key metrics: half-decay distance and r²=0.2 threshold

## Configuration

**Edit these parameters to customize your analysis:**

```{r config}
# ============================================
# INPUT/OUTPUT CONFIGURATION
# ============================================

# VCF file path (required)
vcf_file <- "your_file.vcf.gz"  # Change to your VCF filename

# Output prefix for all files
output_prefix <- "ld_decay"

# ============================================
# ANALYSIS PARAMETERS
# ============================================

# Maximum distance for LD calculation (in bp)
max_distance <- 1000000  # 1 Mb

# Maximum number of SNP pairs to sample per chromosome
sample_size <- 100000

# Maximum number of SNPs to randomly sample per chromosome
# (reduces computation for very dense marker sets)
max_snps <- 20000

# Minimum number of individuals required for LD calculation
min_individuals <- 10

# Random seed for reproducibility (set to NULL for random sampling)
random_seed <- 12345

# ============================================
# DISTANCE BINS FOR SMOOTHING
# ============================================

# Distance bins for grouping SNP pairs (in bp)
distance_bins <- c(0, 1000, 2500, 5000, 10000, 25000, 50000, 
                   100000, 250000, 500000, 1000000)
```

## Load Required Libraries

```{r libraries}
cat("Loading required libraries...\n")

# Core packages
required_packages <- c("vcfR", "ggplot2", "dplyr")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Required package '", pkg, "' is not installed.\n",
         "Please install it with: install.packages('", pkg, "')")
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# Optional color palette
use_metbrewer <- FALSE
if (requireNamespace("MetBrewer", quietly = TRUE)) {
  suppressPackageStartupMessages(library(MetBrewer))
  use_metbrewer <- TRUE
  cat("Using MetBrewer color palettes\n")
} else {
  cat("Note: MetBrewer package not found. Using default colors.\n")
}

cat("All libraries loaded successfully.\n\n")
```

## Validate Configuration

```{r validate}
# Check if VCF file exists
if (!file.exists(vcf_file)) {
  stop("Error: VCF file '", vcf_file, "' not found!\n",
       "Please update the 'vcf_file' variable in the configuration section.")
}

# Print analysis configuration
cat("==========================================\n")
cat("LD Decay Analysis Configuration\n")
cat("==========================================\n")
cat("VCF file:           ", vcf_file, "\n")
cat("Max distance:       ", format(max_distance, big.mark = ","), " bp\n")
cat("Sample size:        ", format(sample_size, big.mark = ","), " SNP pairs per chromosome\n")
cat("Max SNPs sampled:   ", format(max_snps, big.mark = ","), " per chromosome\n")
cat("Min individuals:    ", min_individuals, "\n")
cat("Output prefix:      ", output_prefix, "\n")
cat("Random seed:        ", ifelse(is.null(random_seed), "NULL (random)", random_seed), "\n")
cat("==========================================\n\n")
```

## Read VCF File

```{r read_vcf}
cat("Reading VCF file...\n")

vcf <- tryCatch({
  read.vcfR(vcf_file, verbose = FALSE)
}, error = function(e) {
  stop("Error reading VCF file: ", e$message, "\n",
       "Please ensure the file is a valid VCF format.")
})

n_variants <- nrow(vcf@fix)
n_samples <- ncol(vcf@gt) - 1

cat("VCF file loaded successfully!\n")
cat("  Variants: ", format(n_variants, big.mark = ","), "\n")
cat("  Samples:  ", n_samples, "\n\n")

# Validation checks
if (n_variants < 10) {
  stop("Error: VCF file contains too few variants (", n_variants, 
       "). Need at least 10 variants.")
}

if (n_samples < min_individuals) {
  stop("Error: VCF file contains too few samples (", n_samples, "). ",
       "Need at least ", min_individuals, " samples.")
}
```

## Extract Genotype Data

```{r extract_genotypes}
cat("Extracting genotype data...\n")

# Extract genotype matrix
gt_matrix <- tryCatch({
  extract.gt(vcf, element = "GT", as.numeric = TRUE)
}, error = function(e) {
  stop("Error extracting genotypes: ", e$message)
})

# Get chromosome and position information
chrom_info <- data.frame(
  chromosome = as.character(vcf@fix[, "CHROM"]),
  position = as.numeric(vcf@fix[, "POS"]),
  snp_id = 1:nrow(vcf@fix),
  stringsAsFactors = FALSE
)

# Summary of chromosomes
chrom_summary <- chrom_info %>%
  group_by(chromosome) %>%
  summarise(
    n_snps = n(),
    min_pos = min(position),
    max_pos = max(position),
    span_mb = round((max(position) - min(position)) / 1e6, 2)
  ) %>%
  arrange(desc(n_snps))

cat("\nChromosome summary:\n")
print(chrom_summary)
```

## Define Core Functions

### Calculate r² Between SNPs

```{r function_r2}
# Function to calculate r-squared between two SNP vectors
calculate_r2 <- function(snp1, snp2, min_ind = min_individuals) {
  # Remove individuals with missing data
  valid_idx <- !is.na(snp1) & !is.na(snp2)
  
  if (sum(valid_idx) < min_ind) return(NA)
  
  snp1_clean <- snp1[valid_idx]
  snp2_clean <- snp2[valid_idx]
  
  # Check for variation
  if (sd(snp1_clean) == 0 || sd(snp2_clean) == 0) return(NA)
  
  # Calculate correlation and square it for r²
  r2 <- cor(snp1_clean, snp2_clean)^2
  
  # Ensure r² is between 0 and 1
  if (is.na(r2) || is.nan(r2) || r2 < 0 || r2 > 1) {
    return(NA)
  }
  
  return(r2)
}
```

### Estimate LD Decay per Chromosome

```{r function_ld_decay}
# Function to estimate LD decay for a single chromosome
estimate_ld_decay_chromosome <- function(chrom_name, max_dist = max_distance, 
                                         samp_size = sample_size, max_snp = max_snps) {
  cat("Processing chromosome:", chrom_name, "\n")
  
  # Get SNPs for this chromosome
  chrom_snps <- chrom_info[chrom_info$chromosome == chrom_name, ]
  
  if (nrow(chrom_snps) < 10) {
    cat("  Skipping - insufficient SNPs (", nrow(chrom_snps), ")\n")
    return(NULL)
  }
  
  # Sample SNPs if too many
  if (nrow(chrom_snps) > max_snp) {
    if (!is.null(random_seed)) set.seed(random_seed + match(chrom_name, unique(chrom_info$chromosome)))
    sampled_idx <- sample(1:nrow(chrom_snps), max_snp)
    chrom_snps <- chrom_snps[sampled_idx, ]
    cat("  Sampled", max_snp, "SNPs from", nrow(chrom_snps), "\n")
  }
  
  # Get genotype data for this chromosome
  chrom_gt <- gt_matrix[chrom_snps$snp_id, , drop = FALSE]
  chrom_positions <- chrom_snps$position
  
  # Generate all possible SNP pairs
  n_snps <- nrow(chrom_snps)
  pairs_df <- expand.grid(i = 1:n_snps, j = 1:n_snps)
  pairs_df <- pairs_df[pairs_df$i < pairs_df$j, ]  # Only unique pairs
  
  # Calculate distances
  pairs_df$distance <- abs(chrom_positions[pairs_df$i] - chrom_positions[pairs_df$j])
  
  # Filter by max distance
  pairs_df <- pairs_df[pairs_df$distance <= max_dist & pairs_df$distance > 0, ]
  
  if (nrow(pairs_df) == 0) {
    cat("  No SNP pairs within max distance\n")
    return(NULL)
  }
  
  # Sample pairs if too many
  if (nrow(pairs_df) > samp_size) {
    if (!is.null(random_seed)) set.seed(random_seed + match(chrom_name, unique(chrom_info$chromosome)) + 1000)
    pairs_df <- pairs_df[sample(nrow(pairs_df), samp_size), ]
  }
  
  cat("  Calculating r² for", nrow(pairs_df), "SNP pairs...\n")
  
  # Calculate r² for each pair
  pairs_df$r2 <- mapply(function(i, j) {
    calculate_r2(chrom_gt[i, ], chrom_gt[j, ])
  }, pairs_df$i, pairs_df$j)
  
  # Remove NA values
  pairs_df <- pairs_df[!is.na(pairs_df$r2), ]
  
  if (nrow(pairs_df) == 0) {
    cat("  No valid r² values calculated\n")
    return(NULL)
  }
  
  cat("  Complete:", nrow(pairs_df), "valid r² values\n")
  
  # Return results
  data.frame(
    chromosome = chrom_name,
    distance = pairs_df$distance,
    r2 = pairs_df$r2,
    stringsAsFactors = FALSE
  )
}
```

### Fit Exponential Decay Model

```{r function_fit_decay}
# Function to fit exponential decay model
fit_exponential_decay <- function(distances, r2_values) {
  # Remove NA and infinite values
  valid_idx <- !is.na(r2_values) & !is.na(distances) & 
                is.finite(r2_values) & is.finite(distances) & 
                distances > 0
  
  if (sum(valid_idx) < 10) {
    return(list(a = NA, b = NA, r_squared = NA))
  }
  
  dist <- distances[valid_idx]
  r2 <- r2_values[valid_idx]
  
  # Try to fit exponential decay: r² = a * exp(-b * distance)
  tryCatch({
    # Initial parameter guesses
    start_a <- max(r2, na.rm = TRUE)
    start_b <- -log(0.5) / median(dist, na.rm = TRUE)
    
    model <- nls(r2 ~ a * exp(-b * dist), 
                 start = list(a = start_a, b = start_b),
                 control = list(maxiter = 100, warnOnly = TRUE))
    
    params <- coef(model)
    
    # Calculate R-squared
    predictions <- predict(model)
    ss_res <- sum((r2 - predictions)^2)
    ss_tot <- sum((r2 - mean(r2))^2)
    r_squared <- 1 - (ss_res / ss_tot)
    
    list(
      a = params["a"],
      b = params["b"],
      r_squared = r_squared
    )
  }, error = function(e) {
    list(a = NA, b = NA, r_squared = NA)
  })
}
```

## Calculate LD Decay

```{r calculate_ld, results='hide'}
cat("\nCalculating LD decay for each chromosome...\n")
cat("==========================================\n")

# Get unique chromosomes
chromosomes <- unique(chrom_info$chromosome)

# Process each chromosome
ld_results_list <- lapply(chromosomes, function(chrom) {
  estimate_ld_decay_chromosome(chrom)
})

# Combine results
all_ld_results <- do.call(rbind, ld_results_list[!sapply(ld_results_list, is.null)])

cat("\n==========================================\n")
cat("LD calculation complete!\n")
cat("Total SNP pairs analyzed:", format(nrow(all_ld_results), big.mark = ","), "\n")
cat("Chromosomes processed:", length(unique(all_ld_results$chromosome)), "\n")
cat("==========================================\n\n")
```

## Summary Statistics

```{r summary_stats}
# Create distance bins and summarize
all_ld_results$distance_bin <- cut(
  all_ld_results$distance,
  breaks = distance_bins,
  include.lowest = TRUE
)

ld_summary <- all_ld_results %>%
  group_by(chromosome, distance_bin) %>%
  summarise(
    n_pairs = n(),
    mean_r2 = mean(r2, na.rm = TRUE),
    median_r2 = median(r2, na.rm = TRUE),
    sd_r2 = sd(r2, na.rm = TRUE),
    mean_distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(distance_bin))

# Display summary table
cat("LD Summary Statistics (first 20 rows):\n")
print(head(ld_summary, 20))
```

## Fit Decay Models

```{r fit_models}
cat("\nFitting exponential decay models...\n")

# Fit model for each chromosome
decay_parameters <- data.frame()
smooth_decay_data <- data.frame()

for (chrom in unique(all_ld_results$chromosome)) {
  chrom_data <- all_ld_results[all_ld_results$chromosome == chrom, ]
  
  # Fit model
  model_fit <- fit_exponential_decay(chrom_data$distance, chrom_data$r2)
  
  if (!is.na(model_fit$a) && !is.na(model_fit$b)) {
    # Calculate half-decay distance
    half_decay <- log(2) / model_fit$b
    
    # Calculate r²=0.2 threshold distance
    r2_0.2_dist <- if (model_fit$a > 0.2) {
      -log(0.2 / model_fit$a) / model_fit$b
    } else {
      NA
    }
    
    # Store parameters
    decay_parameters <- rbind(
      decay_parameters,
      data.frame(
        chromosome = chrom,
        a = model_fit$a,
        b = model_fit$b,
        r_squared = model_fit$r_squared,
        half_decay_distance = half_decay,
        r2_0.2_distance = r2_0.2_dist
      )
    )
    
    # Generate smooth curve
    dist_range <- seq(1, max_distance, length.out = 100)
    predicted_r2 <- model_fit$a * exp(-model_fit$b * dist_range)
    
    smooth_decay_data <- rbind(
      smooth_decay_data,
      data.frame(
        chromosome = chrom,
        distance = dist_range,
        predicted_r2 = predicted_r2
      )
    )
  }
}

# Display decay parameters
cat("\nDecay Parameters:\n")
print(decay_parameters)
```

## Genome-wide Analysis

```{r genome_wide}
cat("\nCalculating genome-wide LD decay...\n")

# Bin all data across genome
all_ld_results$genome_bin <- cut(
  all_ld_results$distance,
  breaks = distance_bins,
  include.lowest = TRUE
)

genome_wide_summary <- all_ld_results %>%
  group_by(genome_bin) %>%
  summarise(
    n_pairs = n(),
    mean_r2 = mean(r2, na.rm = TRUE),
    median_r2 = median(r2, na.rm = TRUE),
    sd_r2 = sd(r2, na.rm = TRUE),
    mean_distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(genome_bin))

# Fit genome-wide model
gw_model <- fit_exponential_decay(all_ld_results$distance, all_ld_results$r2)

if (!is.na(gw_model$a) && !is.na(gw_model$b)) {
  gw_half_decay <- log(2) / gw_model$b
  gw_r2_0.2 <- if (gw_model$a > 0.2) {
    -log(0.2 / gw_model$a) / gw_model$b
  } else {
    NA
  }
  
  genome_wide_params <- data.frame(
    a = gw_model$a,
    b = gw_model$b,
    r_squared = gw_model$r_squared,
    half_decay_distance = gw_half_decay,
    r2_0.2_distance = gw_r2_0.2
  )
  
  # Generate smooth curve
  gw_dist_range <- seq(1, max_distance, length.out = 100)
  gw_smooth_data <- data.frame(
    distance = gw_dist_range,
    predicted_r2 = gw_model$a * exp(-gw_model$b * gw_dist_range)
  )
  
  cat("\nGenome-wide Decay Parameters:\n")
  cat("  Half-decay distance:", format(round(gw_half_decay), big.mark = ","), "bp\n")
  if (!is.na(gw_r2_0.2)) {
    cat("  r²=0.2 distance:    ", format(round(gw_r2_0.2), big.mark = ","), "bp\n")
  }
  cat("  Model R²:           ", round(gw_model$r_squared, 3), "\n")
} else {
  genome_wide_params <- NULL
  gw_smooth_data <- NULL
  cat("Warning: Could not fit genome-wide model\n")
}
```

## Visualizations

### Setup Colors

```{r setup_colors}
# Set up color palette
n_chroms <- length(unique(all_ld_results$chromosome))

if (use_metbrewer && n_chroms <= 12) {
  color_palette <- met.brewer("Hiroshige", n = n_chroms)
} else {
  # Use rainbow colors as fallback
  color_palette <- rainbow(n_chroms, s = 0.6, v = 0.8)
}

names(color_palette) <- unique(all_ld_results$chromosome)
```

### Genome-wide LD Decay (Binned)

```{r plot_genome_wide_binned, fig.width=10, fig.height=8}
if (!is.null(genome_wide_params)) {
  subtitle_text <- paste0(
    "Half-decay: ", format(round(genome_wide_params$half_decay_distance), big.mark = ","), " bp"
  )
  
  if (!is.na(genome_wide_params$r2_0.2_distance)) {
    subtitle_text <- paste0(
      subtitle_text, 
      " | r²=0.2: ", 
      format(round(genome_wide_params$r2_0.2_distance), big.mark = ","), 
      " bp"
    )
  }
  
  genome_wide_plot <- ggplot() +
    geom_point(
      data = genome_wide_summary,
      aes(x = mean_distance/1000, y = mean_r2),
      color = "darkblue",
      size = 3,
      alpha = 0.6
    ) +
    geom_line(
      data = gw_smooth_data,
      aes(x = distance/1000, y = predicted_r2),
      color = "red",
      size = 1.5
    ) +
    geom_hline(yintercept = 0.2, color = "orange", linetype = "dotted", size = 0.8) +
    geom_vline(
      xintercept = genome_wide_params$half_decay_distance/1000,
      color = "red",
      linetype = "dashed",
      size = 1
    )
  
  if (!is.na(genome_wide_params$r2_0.2_distance)) {
    genome_wide_plot <- genome_wide_plot +
      geom_vline(
        xintercept = genome_wide_params$r2_0.2_distance/1000,
        color = "orange",
        linetype = "dashed",
        size = 1
      )
  }
  
  genome_wide_plot <- genome_wide_plot +
    scale_x_log10(breaks = c(1, 10, 100, 1000), labels = c("1", "10", "100", "1000")) +
    labs(
      title = "Genome-wide Average LD Decay (Binned)",
      subtitle = subtitle_text,
      x = "Distance (kb, log scale)",
      y = "Mean r²"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray50"),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      panel.grid.minor = element_blank()
    )
  
  print(genome_wide_plot)
}
```

### Genome-wide LD Decay (Raw Data)

```{r plot_genome_wide_raw, fig.width=10, fig.height=8}
if (!is.null(genome_wide_params)) {
  # Sample raw data if too many points
  max_raw_points <- 5000
  if (nrow(all_ld_results) > max_raw_points) {
    sampled_raw <- all_ld_results[sample(nrow(all_ld_results), max_raw_points), ]
    points_label <- paste0(format(max_raw_points, big.mark = ","), " sampled points of ", 
                          format(nrow(all_ld_results), big.mark = ","))
  } else {
    sampled_raw <- all_ld_results
    points_label <- paste0(format(nrow(all_ld_results), big.mark = ","), " SNP pairs")
  }
  
  subtitle_text_raw <- paste0(
    "Half-decay: ", format(round(genome_wide_params$half_decay_distance), big.mark = ","), " bp"
  )
  
  if (!is.na(genome_wide_params$r2_0.2_distance)) {
    subtitle_text_raw <- paste0(
      subtitle_text_raw,
      " | r²=0.2: ",
      format(round(genome_wide_params$r2_0.2_distance), big.mark = ","),
      " bp | "
    )
  } else {
    subtitle_text_raw <- paste0(subtitle_text_raw, " | ")
  }
  subtitle_text_raw <- paste0(subtitle_text_raw, points_label)
  
  genome_wide_raw_plot <- ggplot() +
    geom_point(
      data = sampled_raw,
      aes(x = distance/1000, y = r2),
      color = "darkblue",
      size = 1,
      alpha = 0.2
    ) +
    geom_line(
      data = gw_smooth_data,
      aes(x = distance/1000, y = predicted_r2),
      color = "red",
      size = 1.5
    ) +
    geom_hline(yintercept = 0.2, color = "orange", linetype = "dotted", size = 0.8) +
    geom_vline(
      xintercept = genome_wide_params$half_decay_distance/1000,
      color = "red",
      linetype = "dashed",
      size = 1
    )
  
  if (!is.na(genome_wide_params$r2_0.2_distance)) {
    genome_wide_raw_plot <- genome_wide_raw_plot +
      geom_vline(
        xintercept = genome_wide_params$r2_0.2_distance/1000,
        color = "orange",
        linetype = "dashed",
        size = 1
      )
  }
  
  genome_wide_raw_plot <- genome_wide_raw_plot +
    scale_x_log10(breaks = c(1, 10, 100, 1000), labels = c("1", "10", "100", "1000")) +
    labs(
      title = "Genome-wide LD Decay (Raw Data)",
      subtitle = subtitle_text_raw,
      x = "Distance (kb, log scale)",
      y = "r²"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray50"),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      panel.grid.minor = element_blank()
    )
  
  print(genome_wide_raw_plot)
}
```

### Per-Chromosome LD Decay (Combined)

```{r plot_combined, fig.width=14, fig.height=8}
if (nrow(smooth_decay_data) > 0) {
  ld_decay_plot <- ggplot() +
    geom_point(
      data = ld_summary,
      aes(x = mean_distance/1000, y = mean_r2, color = chromosome),
      size = 2,
      alpha = 0.6
    ) +
    geom_line(
      data = smooth_decay_data,
      aes(x = distance/1000, y = predicted_r2, color = chromosome),
      size = 1.2,
      alpha = 0.9
    ) +
    scale_x_log10(breaks = c(1, 10, 100, 1000), labels = c("1", "10", "100", "1000")) +
    labs(
      title = "Linkage Disequilibrium Decay Across Chromosomes",
      subtitle = "Exponential decay curves",
      x = "Distance (kb, log scale)",
      y = "Mean r²",
      color = "Chromosome"
    ) +
    theme_minimal() +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray50"),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12)
    ) +
    scale_color_manual(values = color_palette)
  
  print(ld_decay_plot)
}
```

### Per-Chromosome LD Decay (Faceted)

```{r plot_faceted, fig.width=12, fig.height=8}
if (nrow(smooth_decay_data) > 0) {
  faceted_ld_plot <- ggplot() +
    geom_point(
      data = ld_summary,
      aes(x = mean_distance/1000, y = mean_r2),
      color = "darkblue",
      size = 2,
      alpha = 0.6
    ) +
    geom_line(
      data = smooth_decay_data,
      aes(x = distance/1000, y = predicted_r2),
      color = "steelblue",
      size = 1.2
    ) +
    facet_wrap(~ chromosome, scales = "free") +
    scale_x_log10(breaks = c(1, 10, 100, 1000), labels = c("1", "10", "100", "1000")) +
    labs(
      title = "LD Decay by Chromosome",
      x = "Distance (kb, log scale)",
      y = "Mean r²"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(size = 10, face = "bold"),
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.text = element_text(size = 8)
    )
  
  print(faceted_ld_plot)
}
```

### Decay Distance Comparison

```{r plot_comparison, fig.width=10, fig.height=6}
if (nrow(decay_parameters) > 1) {
  # Reshape data for grouped bar plot
  decay_comparison <- data.frame(
    chromosome = rep(decay_parameters$chromosome, 2),
    distance_type = rep(c("Half-decay", "r²=0.2"), each = nrow(decay_parameters)),
    distance_kb = c(
      decay_parameters$half_decay_distance/1000,
      decay_parameters$r2_0.2_distance/1000
    )
  )
  
  # Remove NA values
  decay_comparison <- decay_comparison[!is.na(decay_comparison$distance_kb), ]
  
  decay_comparison_plot <- ggplot(
    decay_comparison,
    aes(x = chromosome, y = distance_kb, fill = distance_type)
  ) +
    geom_col(position = "dodge", alpha = 0.8) +
    scale_fill_manual(values = c("Half-decay" = "steelblue", "r²=0.2" = "orange")) +
    labs(
      title = "LD Decay Distances by Chromosome",
      subtitle = "Comparison of half-decay and r²=0.2 thresholds",
      x = "Chromosome",
      y = "Distance (kb)",
      fill = "Threshold"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray50"),
      legend.position = "top"
    )
  
  print(decay_comparison_plot)
}
```

## Save Results

```{r save_results}
# Save raw LD results
raw_file <- paste0(output_prefix, "_raw_results.csv")
write.csv(all_ld_results, raw_file, row.names = FALSE)
cat("Saved:", raw_file, "\n")

# Save summary statistics
summary_file <- paste0(output_prefix, "_summary.csv")
write.csv(ld_summary, summary_file, row.names = FALSE)
cat("Saved:", summary_file, "\n")

# Save decay parameters
if (nrow(decay_parameters) > 0) {
  params_file <- paste0(output_prefix, "_parameters.csv")
  write.csv(decay_parameters, params_file, row.names = FALSE)
  cat("Saved:", params_file, "\n")
}

# Save genome-wide parameters
if (!is.null(genome_wide_params)) {
  gw_params_file <- paste0(output_prefix, "_genome_wide_parameters.csv")
  write.csv(genome_wide_params, gw_params_file, row.names = FALSE)
  cat("Saved:", gw_params_file, "\n")
}
```

## Final Summary

```{r final_summary}
cat("\n==========================================\n")
cat("LD Decay Analysis Complete!\n")
cat("==========================================\n")
cat("Total SNP pairs analyzed:  ", format(nrow(all_ld_results), big.mark = ","), "\n")
cat("Chromosomes processed:     ", length(unique(all_ld_results$chromosome)), "\n")

if (!is.null(genome_wide_params)) {
  cat("\nGenome-wide metrics:\n")
  cat("  Half-decay distance:   ", format(round(genome_wide_params$half_decay_distance), big.mark = ","), " bp\n")
  if (!is.na(genome_wide_params$r2_0.2_distance)) {
    cat("  r²=0.2 distance:       ", format(round(genome_wide_params$r2_0.2_distance), big.mark = ","), " bp\n")
  }
  cat("  Model R²:              ", round(genome_wide_params$r_squared, 3), "\n")
}

cat("\nOutput files created with prefix '", output_prefix, "':\n", sep = "")
cat("  - Raw LD calculations\n")
cat("  - Summary statistics (binned)\n")
cat("  - Decay parameters (per chromosome)\n")
cat("  - Genome-wide parameters\n")
cat("==========================================\n")
```

## Session Information

```{r session_info}
sessionInfo()
```
